---
output: github_document
---
# Shennong: Toolkit for Interpreting Omics Data

<!-- badges: start -->
[![R-CMD-check](https://github.com/zerostwo/shennong/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zerostwo/shennong/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/zerostwo/shennong/branch/main/graph/badge.svg)](https://app.codecov.io/gh/zerostwo/shennong?branch=main)
[![lifecycle](https://img.shields.io/badge/lifecycle-Experimental-important.svg)](https://lifecycle.r-lib.org/articles/stages.html)
<!-- badges: end -->

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## 🤪 Overview

`Shennong` is a toolkit for interpreting omics data and maintained by [Songqi Duan](https://blog.catplot.com).

## 📦 Installation

Install the latest version from [GitHub](https://github.com/zerostwo/shennong) as follow:

```{r, eval = FALSE}
if (!require(devtools)) install.packages("devtools")
devtools::install_github("zerostwo/shennong")
```

## 🧩 Code of Conduct
  
Please note that the shennong project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
