---
title: "Differential Abundance and State Priority"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential Abundance and State Priority}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Shennong treats biological samples—not cells—as the inferential unit for
differential abundance. It also separates three distinct state-priority
questions: perturbation separability, bulk-phenotype-guided cell selection, and
rare-state discovery.

## Sample-level differential abundance

```{r, eval=FALSE}
library(Shennong)

object <- sn_test_abundance(
  object,
  method = "propeller",
  sample_by = "patient",
  condition_by = "response",
  cell_type_by = "cell_type",
  contrast = c("responder", "nonresponder"),
  design = c("batch", "sex"),
  store_name = "response_abundance"
)
```

Use the permutation backend as a transparent validation and Milo for
neighborhood-level effects:

```{r, eval=FALSE}
permutation <- sn_test_abundance(
  object, method = "permutation", sample_by = "patient",
  condition_by = "response", cell_type_by = "cell_type",
  contrast = c("responder", "nonresponder"), permutations = 999,
  return_object = FALSE
)

milo <- sn_test_abundance(
  object, method = "milo", sample_by = "patient",
  condition_by = "response", cell_type_by = "cell_type",
  contrast = c("responder", "nonresponder"),
  backend_control = list(milo = list(reduction = "pca")),
  return_object = FALSE
)
```

For Bayesian compositional inference, run scCODA or pertpy in the Python
environment owned by the project and pass its effect table through the explicit
adapter. The table needs a `feature` or `cell_type` column and an `effect` or
`estimate` column; inclusion probabilities and credible-effect flags are
retained when supplied. The adapter does not convert posterior inclusion
probabilities into frequentist p values; those fields remain missing unless the
backend explicitly provides them.

```{r, eval=FALSE}
sccoda <- sn_test_abundance(
  object, method = "sccoda", sample_by = "patient",
  condition_by = "response", cell_type_by = "cell_type",
  contrast = c("responder", "nonresponder"),
  backend_control = list(result = list(table = sccoda_effects)),
  return_object = FALSE
)
sccoda$tables$primary
```

Discover and retrieve the durable result:

```{r, eval=FALSE}
sn_list_results(object, type = "differential_abundance")
abundance <- sn_get_result(
  object, "differential_abundance", "response_abundance"
)
abundance$tables$primary
abundance$tables$sample_proportions
abundance$tables$sample_contributions
sn_plot_abundance(abundance)
```

## Prioritize responsive states

The default method holds out biological samples when `sample_by` is supplied,
predicts the condition independently within each state, and records a
sample-label permutation null.

```{r, eval=FALSE}
object <- sn_prioritize_states(
  object,
  method = "augur",
  phenotype = "response",
  sample_by = "patient",
  state_by = "cell_type",
  contrast = c("responder", "nonresponder"),
  store_name = "response_priority"
)

priority <- sn_get_result(object, "state_priority", "response_priority")
priority$tables$primary
priority$tables$sample_contributions
priority$tables$permutation_null
sn_plot_state_priority(priority)
```

RareQ first discovers topology-supported rare states and then tests their
sample-level abundance association. Scissor answers a different question and
therefore requires a gene-by-bulk-sample expression matrix plus its phenotype;
a cell metadata label is not accepted as a substitute.

```{r, eval=FALSE}
rareq <- sn_prioritize_states(
  object, method = "rareq", phenotype = "response",
  sample_by = "patient", contrast = c("responder", "nonresponder"),
  return_object = FALSE
)

scissor <- sn_prioritize_states(
  object, method = "scissor", phenotype = "bulk_response",
  state_by = "cell_type", sample_by = "patient",
  bulk_expression = bulk_expression,
  bulk_phenotype = bulk_response,
  family = "binomial",
  return_object = FALSE
)
```
