---
title: "Differential Abundance and State Priority"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential Abundance and State Priority}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
run_vignette <- identical(tolower(Sys.getenv("SHENNONG_RUN_VIGNETTES")), "true")
real_profile <- tolower(Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core"))
root_candidates <- unique(c(
  Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = ""),
  file.path(getwd(), "data-local", "pkgdown-real"),
  file.path(getwd(), "..", "data-local", "pkgdown-real")
))
root_candidates <- root_candidates[nzchar(root_candidates)]
existing_roots <- root_candidates[dir.exists(root_candidates)]
real_root <- if (length(existing_roots)) existing_roots[[1]] else root_candidates[[1]]
pbmc_path <- file.path(real_root, "single-cell", "kotliarov_pbmc.qs2")
melanoma_path <- file.path(real_root, "cancer", "melanoma_gse72056.qs2")
tcga_path <- file.path(real_root, "bulk", "tcga_skcm.qs2")
core_ready <- run_vignette && file.exists(pbmc_path)
scissor_ready <- run_vignette && identical(real_profile, "all") &&
  file.exists(melanoma_path) && file.exists(tcga_path) &&
  requireNamespace("Scissor", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 5
)
```

This article uses public observations rather than simulated cell labels. The
core analysis uses 2,000 cells sampled across all 20 biological samples in the
Kotliarov PBMC CITE-seq vaccine-response cohort. Scissor is a separate extended
analysis linking the public GSE72056 melanoma cells to the curated
TCGA-SKCM phenotype.

```{r runtime-status}
status <- data.frame(
  workflow = c("sample-level abundance", "state priority", "Scissor"),
  mode = c("core", "core", "extended"),
  data = c("Kotliarov PBMC", "Kotliarov PBMC", "GSE72056 + TCGA-SKCM"),
  status = c(
    if (core_ready) "executed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture missing",
    if (core_ready) "executed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture missing",
    if (scissor_ready) "executed below" else if (!identical(real_profile, "all")) "disabled: requires SHENNONG_REAL_PROFILE=all" else if (!requireNamespace("Scissor", quietly = TRUE)) "dependency missing: Scissor" else "fixture missing"
  ),
  check.names = FALSE
)
knitr::kable(status)
```

## Prepare real sample and state labels

The author-provided `real_sample`, response, and batch fields remain the
replicate design. Cell states are obtained by clustering the real RNA counts;
they are not assigned from a fabricated metadata vector.

```{r prepare-pbmc, eval=core_ready}
library(Shennong)
library(Seurat)

pbmc <- qs2::qs_read(pbmc_path)
pbmc <- sn_run_cluster(
  pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.5,
  species = "human",
  verbose = FALSE
)
pbmc$cell_state <- paste0("cluster_", pbmc$seurat_clusters)

pbmc_metadata <- pbmc[[]]
sample_design <- do.call(rbind, lapply(
  split(seq_len(nrow(pbmc_metadata)), pbmc_metadata$real_sample),
  function(index) data.frame(
    real_sample = pbmc_metadata$real_sample[index[[1]]],
    real_response = unique(pbmc_metadata$real_response[index]),
    observed_batches = paste(sort(unique(pbmc_metadata$real_batch[index])), collapse = "+"),
    cells = length(index),
    row.names = NULL
  )
))
sample_design <- sample_design[order(sample_design$real_sample), ]
sample_design
```

## Sample-level differential abundance

Propeller models the 20 biological samples. The contrast is high versus low
baseline vaccine response. One donor has cells in both acquisition batches, so
batch is reported above but is not forced into an invalid one-value-per-sample
covariate. A sample-label permutation analysis provides a second result whose
null is easy to audit.

```{r abundance, eval=core_ready}
abundance <- sn_test_abundance(
  pbmc,
  method = "propeller",
  sample_by = "real_sample",
  condition_by = "real_response",
  cell_type_by = "cell_state",
  contrast = c("d0 high", "d0 low"),
  return_object = FALSE
)

permutation <- sn_test_abundance(
  pbmc,
  method = "permutation",
  sample_by = "real_sample",
  condition_by = "real_response",
  cell_type_by = "cell_state",
  contrast = c("d0 high", "d0 low"),
  permutations = 199,
  seed = 717,
  return_object = FALSE
)

abundance$tables$primary[order(abundance$tables$primary$adjusted_p_value), ]
head(abundance$tables$sample_proportions)
head(permutation$tables$permutation_null)
sn_plot_abundance(abundance)
```

## Prioritize response-separable states

The built-in sample-aware Augur path holds out complete biological samples and
permutes sample labels. Thus, the displayed AUC and empirical p-value do not
count cells as independent replicates.

```{r priority, eval=core_ready}
priority <- sn_prioritize_states(
  pbmc,
  method = "augur",
  phenotype = "real_response",
  sample_by = "real_sample",
  state_by = "cell_state",
  contrast = c("d0 high", "d0 low"),
  max_features = 250,
  max_cells_per_state = 300,
  permutations = 49,
  seed = 717,
  return_object = FALSE
)

priority$tables$primary
head(priority$tables$sample_contributions)
sn_plot_state_priority(priority)
```

## Extended: phenotype-guided Scissor

This chunk runs only with `SHENNONG_REAL_PROFILE=all` and an installed Scissor
backend. The single-cell input is the author-normalized GSE72056 matrix; the
bulk input is TCGA-SKCM upper-quartile log2-normalized RSEM TPM. The phenotype
is the curated primary-versus-metastatic label, named and aligned to the bulk
columns. A backend error is printed as a failed status and is never replaced by
a hand-made selection table.

```{r scissor, eval=scissor_ready}
melanoma <- qs2::qs_read(melanoma_path)
tcga <- qs2::qs_read(tcga_path)
bulk_phenotype <- stats::setNames(
  tcga$sample_data$sample_type,
  rownames(tcga$sample_data)
)

scissor_attempt <- tryCatch(
  sn_run_scissor(
    melanoma,
    state_by = "cell_type",
    sample_by = "tumor",
    bulk_expression = as.matrix(tcga$log2_uq_rsem_tpm),
    bulk_phenotype = bulk_phenotype,
    family = "binomial",
    assay = "RNA",
    layer = "data",
    seed = 717,
    backend_control = list(nfeatures = 500L, npcs = 10L, cutoff = 0.2),
    return_object = FALSE
  ),
  error = identity
)

if (inherits(scissor_attempt, "error")) {
  knitr::kable(data.frame(
    backend = "Scissor", status = "failed",
    detail = conditionMessage(scissor_attempt), check.names = FALSE
  ))
} else {
  scissor <- scissor_attempt
  print(scissor$tables$model)
  print(scissor$tables$states)
  print(head(scissor$tables$correlations))
  print(sn_plot_scissor(scissor, type = "states"))
  print(sn_plot_scissor(scissor, type = "correlations"))
}
```

Scissor's cutoff governs alpha-search stopping; it does not guarantee a final
selected-cell fraction. Always inspect
`selection_cutoff_satisfied` in the model table before interpreting selected
cells. Bootstrap reliability remains intentionally outside this documentation
build because it is a separate confirmatory computation.
