---
title: "Analysis methods and stored-result contracts"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis methods and stored-result contracts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

Shennong separates method discovery from method execution. The registry shows
the preferred backend, its requirements, whether the adapter has been
implemented, and whether the current machine can run it. A method can therefore
remain visible in the roadmap without being misrepresented as available.

## Discover methods before running a workflow

```{r method-registry}
library(Shennong)

sn_list_methods("trajectory")[, c(
  "name", "default", "implemented", "available", "runtime", "package"
)]

sn_method_status("slingshot", task = "trajectory")
```

Use `implemented` to distinguish a shipped Shennong adapter from a roadmap
entry. Use `available` to determine whether its optional R package, executable,
or pixi runtime is present in the current environment. Installation is always
explicit; discovery never installs software.

## The common result shape

New workflows store results with one versioned contract. Tables, embeddings,
graphs, models, diagnostics, warnings, parameters, input summaries, and
provenance have stable locations.

```{r result-contract}
result <- list(
  schema_version = "1.0",
  analysis_type = "trajectory",
  name = "cd8_slingshot",
  method = "slingshot",
  backend = "slingshot",
  input = list(reduction = "pca", cells = 4L, features = 2L),
  parameters = list(start = "naive"),
  tables = list(
    primary = data.frame(
      cell = paste0("cell", 1:4),
      pseudotime = c(0, 0.3, 0.7, 1)
    )
  ),
  embeddings = list(),
  graphs = list(),
  models = list(),
  diagnostics = list(converged = TRUE),
  warnings = character(),
  provenance = list(
    package_versions = list(Shennong = as.character(packageVersion("Shennong"))),
    random_seed = 1L,
    timestamp = "2026-07-15 UTC"
  )
)

sn_validate_result(result, error = FALSE)
```

## Store, discover, and retrieve from Seurat

The generic functions support new result types while existing helpers such as
`sn_get_de_result()` remain available. Registered legacy result writers are
upgraded to the same contract on write and read.

```{r stored-results, eval = FALSE}
object <- sn_store_result(
  object,
  type = "trajectory",
  name = "cd8_slingshot",
  result = result
)

sn_list_results(object)
sn_list_results(object, type = "trajectory")

trajectory <- sn_get_result(
  object,
  type = "trajectory",
  name = "cd8_slingshot"
)

trajectory$tables$primary
trajectory$provenance

object <- sn_delete_result(
  object,
  type = "trajectory",
  name = "cd8_slingshot"
)
```

Prefer these APIs over direct `object@misc` access. Stable names and explicit
provenance make stored results discoverable by scripts, reports, and future
agent integrations without allowing an LLM to overwrite computational output.
