---
title: "Analysis methods and stored-result contracts"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis methods and stored-result contracts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_root <- Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = "")
pbmc_fixture <- if (nzchar(real_data_root)) {
  file.path(real_data_root, "single-cell", "kotliarov_pbmc.qs2")
} else {
  ""
}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") &&
  nzchar(pbmc_fixture) && file.exists(pbmc_fixture)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  eval = run_vignette
)
```

Shennong separates method discovery from method execution. The registry shows
the preferred backend, its requirements, whether the adapter has been
implemented, and whether the current machine can run it. A method can therefore
remain visible in the roadmap without being misrepresented as available.

## Discover methods before running a workflow

```{r method-registry}
library(Shennong)

method_table <- sn_list_methods("trajectory")[, c(
  "name", "default", "implemented", "available", "runtime", "package"
)]
knitr::kable(method_table)

sn_get_method_status("slingshot", task = "trajectory")
```

Use `implemented` to distinguish a shipped Shennong adapter from a roadmap
entry. Use `available` to determine whether its optional R package, executable,
or pixi runtime is present in the current environment. Installation is always
explicit; discovery never installs software.

## The common result shape on real data

New workflows store results with one versioned contract. This example runs QC
assessment on the public Kotliarov PBMC CITE-seq cohort, stores the result on
the Seurat object, and then validates the actual stored envelope. Tables,
embeddings, graphs, models, diagnostics, warnings, parameters, input summaries,
and provenance have stable locations.

```{r result-contract}
object <- qs2::qs_read(pbmc_fixture)
object <- sn_assess_qc(
  object,
  sample_by = "real_sample",
  result_id = "kotliarov_qc",
  return_object = TRUE,
  verbose = FALSE
)

result <- sn_get_result(object, "qc_assessment", "kotliarov_qc")
sn_validate_result(result, error = FALSE)
result$tables$primary |> head()
result$tables$overall
```

## Store, discover, and retrieve from Seurat

The generic functions support new result types while table-focused helpers such
as `sn_get_de_result()` remain available.

Every public result writer accepts `result_id` as its stable identifier. The
same ID is used to store, list, retrieve, plot, interpret, and delete the
result. Runtime and backend artifacts use `artifact_id` instead.

```{r stored-results}
object <- sn_store_result(
  object,
  type = "qc_assessment",
  result = result,
  result_id = "qc_copy"
)

sn_list_results(object)
sn_list_results(object, type = "qc_assessment")
sn_list_results(object, include_artifacts = TRUE)

qc_copy <- sn_get_result(
  object,
  type = "qc_assessment",
  result_id = "qc_copy_explicit"
)

qc_copy$tables$primary |> head()
qc_copy$provenance

object <- sn_delete_result(
  object,
  type = "qc_assessment",
  result_id = "qc_copy_explicit"
)

sn_list_results(object)
```

Prefer these APIs over direct `object@misc` access. Stable result IDs and explicit
provenance make stored results discoverable by scripts, reports, and future
agent integrations without allowing an LLM to overwrite computational output.

## Export a service-neutral Result Bundle

`sn_build_result_bundle()` turns one retrieved canonical result into the
package-owned JSON handoff contract
`shennong.dev/analysis-result-bundle/v1`. The bundle may reference immutable
inputs and candidate output artifacts, but it never connects to ShennongDB,
calls an orchestrator, or carries credentials. An authorized external service
must independently verify the referenced digests before promoting output
bytes.

```{r result-bundle}
# Discover and retrieve the exact stored result first.
sn_list_results(object, type = "qc_assessment")
qc_result <- sn_get_result(
  object,
  type = "qc_assessment",
  result_id = "kotliarov_qc"
)

bundle <- sn_build_result_bundle(
  qc_result,
  inputs = list(list(
    role = "expression",
    resource_id = "kotliarov-pbmc",
    revision = "replace-with-immutable-revision",
    digest = list(
      algorithm = "sha256",
      value = paste(rep("0", 64L), collapse = "")
    )
  )),
  execution = list(runtime = "interactive-r")
)

sn_validate_result_bundle(bundle, error = FALSE)
bundle_path <- tempfile(fileext = ".json")
handoff <- sn_export_result_bundle(bundle, bundle_path)
handoff[c("size_bytes", "digest")]
unlink(bundle_path)
```

Replace the placeholder revision and digest with the exact values supplied by
the data owner. Artifact paths, when present, must be bundle-relative. The
export return value includes the JSON file's SHA-256 digest for the next
trusted boundary to verify.

## Audit and upgrade existing objects

`sn_audit_results()` checks the canonical result store and every other
populated top-level `object@misc` entry without changing the object. Canonical
results report `valid`, normalizable malformed envelopes report `repairable`,
runtime/cache payloads report `artifact`, and
unknown payloads report `unregistered` rather than being silently omitted or
misclassified as analysis results.
Every analytical result uses `schema_version = "2.0.0"` and a canonical
`tables$primary`; additional named tables carry analysis-specific outputs.

```{r audit-results}
audit <- sn_audit_results(object)
audit[, c(
  "type", "result_id", "contract_scope", "status", "schema_version",
  "primary_rows", "errors"
)]

# Normalize only analytical results. Artifacts and unregistered
# payloads are left untouched.
object <- sn_upgrade_results(object)
stopifnot(all(
  sn_audit_results(object, include_artifacts = FALSE)$status == "valid"
))

sn_audit_results(object, include_artifacts = FALSE)
```

Table-returning helpers such as `sn_get_de_result()` expose the same canonical
primary table. Use `sn_get_result()` when the
full envelope, diagnostics, or provenance is required.

## Use optional ShennongOpt acceleration

Shennong never patches another package during loading. Selected hot paths are
accelerated by the companion R package **ShennongOpt**, which registers guarded
fast replacements (Seurat `RunPCA`/`ScaleData`, scran, decontX, scDblFinder,
Coralysis, UCell, LISI, and Rogue) and rebinds them only while a compatible
Shennong workflow call runs. Unsupported inputs fall back to the captured
upstream implementation, so results stay bit-exact by construction.

```{r acceleration-check}
sn_check_acceleration()
```

Automatic workflow scopes can be disabled per session:

```{r acceleration-default, eval = FALSE}
# Prevent automatic workflow scopes in this R session.
options(shennong.acceleration = FALSE)

# Explicit helpers intentionally ignore the automatic opt-out.
out <- sn_with_acceleration(
  sn_run_cluster(object, resolution = 0.8),
  name = "seurat"
)

# Keep a patch active for direct upstream calls until it is disabled again.
sn_enable_acceleration("seurat")
sn_disable_acceleration("seurat")
```

`options(shennong.acceleration = FALSE)` or `SHENNONG_ACCELERATION_DISABLED=true`
block automatic scopes; neither deactivates a patch that was activated manually.
Set `shennong.opt.threads` (or `SHENNONG_OPT_THREADS`) to cap the thread budget
used by accelerated parallel sections. Accelerated paths never change scientific
output; conformance contracts compare candidate calls against direct upstream
oracles with acceleration disabled.

Hot paths without a ShennongOpt counterpart (for example CellChat, NicheNetR,
SoupX, clusterProfiler caches, tradeSeq, WGCNA, and Seurat merge/JoinLayers)
currently execute their plain upstream implementations. They will gain
accelerations as new ShennongOpt patches become available; provenance records
which patches each run used and which were requested but unavailable.

## Give agents a read-only discovery surface

Install the packaged Shennong Agent Skills and register the bundled stdio MCP
server with an MCP-capable client:

```{r mcp, eval=FALSE}
sn_install_codex_skill(
  path = "~/.agents/skills",
  type = "package_skills"
)

sn_get_mcp_server_config()
```

The MCP server exposes `list_methods`, `method_status`, `function_help`,
`workflow_guide`, and `package_info`. These tools only read installed package
metadata and documentation. Analysis execution remains explicit R code, and
the server offers no arbitrary-code or file-mutation tool.
