---
title: "Analysis methods and stored-result contracts"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis methods and stored-result contracts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_root <- Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = "")
pbmc_fixture <- if (nzchar(real_data_root)) {
  file.path(real_data_root, "single-cell", "kotliarov_pbmc.qs2")
} else {
  ""
}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") &&
  nzchar(pbmc_fixture) && file.exists(pbmc_fixture)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  eval = run_vignette
)
```

Shennong separates method discovery from method execution. The registry shows
the preferred backend, its requirements, whether the adapter has been
implemented, and whether the current machine can run it. A method can therefore
remain visible in the roadmap without being misrepresented as available.

## Discover methods before running a workflow

```{r method-registry}
library(Shennong)

sn_list_methods("trajectory")[, c(
  "name", "default", "implemented", "available", "runtime", "package"
)]

sn_method_status("slingshot", task = "trajectory")
```

Use `implemented` to distinguish a shipped Shennong adapter from a roadmap
entry. Use `available` to determine whether its optional R package, executable,
or pixi runtime is present in the current environment. Installation is always
explicit; discovery never installs software.

## The common result shape on real data

New workflows store results with one versioned contract. This example runs QC
assessment on the public Kotliarov PBMC CITE-seq cohort, stores the result on
the Seurat object, and then validates the actual stored envelope. Tables,
embeddings, graphs, models, diagnostics, warnings, parameters, input summaries,
and provenance have stable locations.

```{r result-contract}
object <- qs2::qs_read(pbmc_fixture)
object <- sn_assess_qc(
  object,
  sample_by = "real_sample",
  store_name = "kotliarov_qc",
  return_object = TRUE,
  verbose = FALSE
)

result <- sn_get_result(object, "qc_assessment", "kotliarov_qc")
sn_validate_result(result, error = FALSE)
result$tables$primary |> head()
result$overall
```

## Store, discover, and retrieve from Seurat

The generic functions support new result types while existing helpers such as
`sn_get_de_result()` remain available. Registered legacy result writers are
upgraded to the same contract on write and read.

```{r stored-results}
object <- sn_store_result(
  object,
  type = "qc_assessment",
  name = "qc_copy",
  result = result
)

sn_list_results(object)
sn_list_results(object, type = "qc_assessment")

qc_copy <- sn_get_result(
  object,
  type = "qc_assessment",
  name = "qc_copy"
)

qc_copy$tables$primary |> head()
qc_copy$provenance

object <- sn_delete_result(
  object,
  type = "qc_assessment",
  name = "qc_copy"
)

sn_list_results(object)
```

Prefer these APIs over direct `object@misc` access. Stable names and explicit
provenance make stored results discoverable by scripts, reports, and future
agent integrations without allowing an LLM to overwrite computational output.

## Export a service-neutral Result Bundle

`sn_build_result_bundle()` turns one retrieved canonical result into the
package-owned JSON handoff contract
`shennong.dev/analysis-result-bundle/v1`. The bundle may reference immutable
inputs and candidate output artifacts, but it never connects to ShennongDB,
calls an orchestrator, or carries credentials. An authorized external service
must independently verify the referenced digests before promoting output
bytes.

```{r result-bundle}
# Discover and retrieve the exact stored result first.
sn_list_results(object, type = "qc_assessment")
qc_result <- sn_get_result(
  object,
  type = "qc_assessment",
  name = "kotliarov_qc"
)

bundle <- sn_build_result_bundle(
  qc_result,
  inputs = list(list(
    role = "expression",
    resource_id = "kotliarov-pbmc",
    revision = "replace-with-immutable-revision",
    digest = list(
      algorithm = "sha256",
      value = paste(rep("0", 64L), collapse = "")
    )
  )),
  execution = list(runtime = "interactive-r")
)

sn_validate_result_bundle(bundle, error = FALSE)
bundle_path <- tempfile(fileext = ".json")
handoff <- sn_export_result_bundle(bundle, bundle_path)
handoff[c("size_bytes", "digest")]
unlink(bundle_path)
```

Replace the placeholder revision and digest with the exact values supplied by
the data owner. Artifact paths, when present, must be bundle-relative. The
export return value includes the JSON file's SHA-256 digest for the next
trusted boundary to verify.

## Audit and upgrade existing objects

`sn_audit_results()` checks every registered result collection, the generic
result store, and every other populated top-level `object@misc` entry without
changing the object. Canonical results report `valid`, older compatible
envelopes report `legacy`, runtime/cache payloads report `artifact`, and
unknown payloads report `unregistered` rather than being silently omitted or
misclassified as analysis results.
Every analytical result uses `schema_version = "1.0.0"` and a canonical
`tables$primary`; specialized table names remain synchronized aliases.

```{r audit-results}
audit <- sn_audit_results(object)
audit[, c(
  "type", "name", "contract_scope", "status", "schema_version",
  "primary_rows", "errors"
)]

# Upgrade only compatible analytical results. Artifacts and unregistered
# payloads are left untouched.
object <- sn_upgrade_results(object)
stopifnot(all(
  sn_audit_results(object, include_artifacts = FALSE)$status == "valid"
))

sn_audit_results(object, include_artifacts = FALSE)
```

Table-returning helpers such as `sn_get_de_result()` remain compatibility
views over the same canonical primary table. Use `sn_get_result()` when the
full envelope, diagnostics, or provenance is required.

## Use validated AutoZyme acceleration

Shennong never patches another package during loading. When an integrated
workflow first needs one of the lazy defaults—CellChat, NicheNetR,
clusterProfiler, fgsea, Seurat, tradeSeq, or WGCNA—it checks that relevant patch.
Each patch requires the pinned AutoZyme build plus an installed, exactly
validated upstream-package version. Eligible patches are activated only inside
the compatible Shennong workflow call, and the pre-call state is restored after
success or error. Missing or drifted dependencies are skipped safely.
Approximate patches are never part of the automatic set. AutoZyme's load-time
change to `future.globals.maxSize` is restored before the analytical expression
runs.

```{r autozyme-check}
autozyme_defaults <- c(
  "cellchat", "nichenetr", "clusterprofiler", "fgsea",
  "seurat", "tradeseq", "wgcna"
)
sn_check_autozyme(autozyme_defaults)
```

```{r autozyme-default, eval = FALSE}
communication <- sn_run_cell_communication(
  object,
  method = "cellchat",
  group_by = "cell_type"
)

# Prevent automatic workflow scopes in this R session.
options(shennong.autozyme = FALSE)

# Explicit helpers intentionally ignore the automatic opt-out.
communication <- sn_with_autozyme({
  sn_run_cell_communication(
    object,
    method = "cellchat",
    group_by = "cell_type"
  )
}, patches = "cellchat")
```

`options(shennong.autozyme = FALSE)`, `AUTOZYME_DISABLED=true`, and
`AUTOZYME_DISABLE=true` block automatic scopes; none deactivates a patch that
the user activated manually, and explicit helpers ignore these controls. For
memory safety, BPCells-backed Seurat layers bypass the Seurat fast patch because
the pinned implementation would coerce them to an in-memory `dgCMatrix`. An
already active Seurat patch is suspended only for that BPCells-backed call and
restored afterward. NicheNetR automatic use remains limited to `single = TRUE`
and a dense numeric ligand-target matrix. Relaxing `strict` or setting
`allow_approximate = TRUE` remains an explicit manual reproducibility decision.
Shennong does not install AutoZyme, upstream packages, or a Python environment.
Result-producing scopes add the patches active for the compatible call to
provenance before restoring the automatic scope. The BPCells bypass applies to
Seurat patch calls;
it does not make every downstream backend BPCells-native. CellChat and tradeSeq,
for example, may still need a controlled sparse materialization or an aggregated
input that fits memory.

## Give agents a read-only discovery surface

Install the packaged Shennong Agent Skills and register the bundled stdio MCP
server with an MCP-capable client:

```{r mcp, eval=FALSE}
sn_install_codex_skill(
  path = "~/.agents/skills",
  type = "package_skills"
)

sn_mcp_server_config()
```

The MCP server exposes `list_methods`, `method_status`, `function_help`,
`workflow_guide`, and `package_info`. These tools only read installed package
metadata and documentation. Analysis execution remains explicit R code, and
the server offers no arbitrary-code or file-mutation tool.
