---
title: "Markers, signatures, pathways, and annotation"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Markers, signatures, pathways, and annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_file <- file.path(
  Sys.getenv("SHENNONG_REAL_DATA_DIR"),
  "single-cell",
  "kotliarov_pbmc.qs2"
)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_requested <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")
run_vignette <- run_requested &&
  requireNamespace("qs2", quietly = TRUE) &&
  file.exists(real_data_file)
run_extended <- run_vignette && identical(real_profile, "all")
celltypist_model <- Sys.getenv("SHENNONG_CELLTYPIST_MODEL")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

After clustering, users usually ask three linked questions:

1. Which genes define each cluster?
2. Which known signatures or pathways explain those genes?
3. How can the evidence be reused for plots and annotation?

Shennong stores DE and enrichment results on the Seurat object so the same
tables can feed dot plots, interpretation prompts, and later reports.

All core results below are computed from the public Kotliarov PBMC CITE-seq
fixture. Its 20 `real_sample` values are the biological replicates,
`real_response` is constant within each sample, and RNA and ADT measurements
remain aligned. No donor, condition, or cell-label column is fabricated.

## Cluster the real PBMC cells and find markers

```{r setup-markers}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- qs2::qs_read(real_data_file)

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

pbmc <- sn_find_de(
  object = pbmc,
  analysis = "markers",
  group_by = "seurat_clusters",
  layer = "data",
  min_pct = 0.25,
  logfc_threshold = 0.25,
  store_name = "cluster_markers",
  return_object = TRUE,
  verbose = FALSE
)

table(pbmc$seurat_clusters, pbmc$real_response)
```

The result is stored under `object@misc$de_results`. Retrieve it by name instead
of relying on a temporary variable from an earlier script.

```{r retrieve-markers}
marker_tbl <- sn_get_de_result(
  pbmc,
  de_name = "cluster_markers",
  top_n = 5
)

head(marker_tbl)
names(pbmc@misc$de_results)
```

## Prioritize interpretable marker classes

Marker tables often contain hundreds of significant genes. Use
`sn_annotate_de_features()` to flag marker genes that are especially useful for
mechanistic interpretation or validation, such as transcription factors,
surface or plasma-membrane genes, cytokines, and chemokines. When called on a
Seurat object, the annotated table is stored as another DE result, so it can be
discovered and retrieved with the same result helpers.

```{r annotate-marker-classes}
pbmc <- sn_annotate_de_features(
  pbmc,
  de_name = "cluster_markers",
  species = "human"
)

sn_list_results(pbmc)

sn_get_de_result(
  pbmc,
  de_name = "cluster_markers_feature_classes",
  top_n = 5
) |>
  dplyr::select(
    cluster,
    gene,
    avg_log2FC,
    feature_classes,
    starts_with("is_")
  ) |>
  head()
```

## Plot stored markers without hand-copying gene lists

Because the DE result is stored, `sn_plot_dot()` can select top markers per
cluster directly.

```{r marker-dotplot}
sn_plot_dot(
  x = pbmc,
  features = "top_markers",
  de_name = "cluster_markers",
  n = 4,
  group_by = "seurat_clusters",
  palette = "RdBu",
  title = "Top markers in the real PBMC subset"
)
```

For canonical checks, pass marker genes explicitly.

```{r canonical-dotplot}
sn_plot_dot(
  x = pbmc,
  features = c("IL7R", "CCR7", "MS4A1", "CD79A", "LYZ", "S100A8", "NKG7"),
  group_by = "seurat_clusters",
  palette = "Purples",
  direction = 1,
  title = "Canonical PBMC markers"
)
```

## Run traceable consensus annotation

`sn_run_annotation()` is the stable annotation entry point. Its default
`method = "consensus"` scores the bundled canonical marker database, retains
the best and second-best candidate, calibrates a confidence margin, assigns
hierarchical labels, and maps known labels to a versioned Cell Ontology
snapshot. It supports both cluster- and cell-level output; without a reference,
each cell inherits the traceable consensus of its cluster.

```{r consensus-annotation}
pbmc <- sn_run_annotation(
  pbmc,
  group_by = "seurat_clusters",
  method = "consensus",
  species = "human",
  tissue = "peripheral blood",
  ontology = TRUE,
  store_name = "pbmc_consensus"
)

annotation_index <- sn_list_results(pbmc, type = "annotation")
annotation_index

annotation <- sn_get_result(
  pbmc,
  type = "annotation",
  name = "pbmc_consensus"
)

head(annotation$tables$clusters)
head(annotation$tables$cells)
annotation$diagnostics

confidence_from_evidence <- sn_annotation_confidence(
  annotation$tables$evidence
)
consensus_from_evidence <- sn_annotation_consensus(
  annotation$tables$evidence,
  ontology = TRUE
)

head(confidence_from_evidence)
head(consensus_from_evidence)

sn_plot_reference_projection(
  pbmc,
  store_name = "pbmc_consensus",
  reduction = "umap",
  color_by = "prediction"
)
```

Every final label remains linked to marker support, conflicts, reference
coverage, confidence, and provenance. LLM helpers can explain this evidence but
cannot replace the stored computational label. Review uncertain assignments
before using them as biological truth.

```{r annotation-review-plots}
sn_review_annotation(pbmc, "pbmc_consensus")

sn_plot_annotation_confidence(
  pbmc,
  store_name = "pbmc_consensus",
  level = "cluster"
)

sn_plot_annotation_markers(
  pbmc,
  store_name = "pbmc_consensus",
  top_n = 5
)
```

If trusted cell-level labels are present, a confusion heatmap makes
disagreements visible. This fixture has no author-curated cell-type truth
column, so the real-data example below compares consensus labels with labels
mapped independently from a disjoint reference subset. That plot measures
reference-mapping consistency; it is not an accuracy estimate or biological
ground truth.

Ontology mapping can also be inspected directly. The `all` profile reports the
versioned mappings for the real consensus labels without changing the stored
annotation.

```{r map-cell-ontology, eval = run_extended}
sn_map_cell_ontology(
  unique(annotation$tables$clusters$prediction),
  strict = FALSE
)
```

## Choose a reference backend deliberately

Use `sn_list_methods("annotation")` and `sn_method_status()` to inspect current
availability. Reference methods should not be used when tissue, disease,
species, assay chemistry, or cell-state coverage is badly mismatched; a high
similarity to an incomplete reference is not evidence that the missing cell
type is absent.

| Backend | Best use | Input and runtime | Stored output |
|---|---|---|---|
| SingleR | Default reference prediction with per-label scores | Query and annotated Seurat/SummarizedExperiment reference; CPU; optional `SingleR` | Full score evidence, pruned label, delta, coverage |
| CellTypist | External pretrained immune atlases | Counts plus model; CPU CLI on `PATH` | Original label columns and consensus evidence |
| Seurat | Anchor transfer from a compatible Seurat reference | Query/reference assays and labels; CPU; optional `Seurat` | Transferred labels and all returned score columns |
| Symphony | Repeated mapping to a compressed atlas | Built reference or Seurat reference; CPU; optional `symphony` | kNN confidence and mapped embedding |
| scmap | Conservative projection to indexed expression centroids | Shared features and annotated reference; CPU; optional `scmap` | Label, similarity, and unassigned status |
| scANVI | Semi-supervised mapping with batch structure | Raw counts, labels, batch metadata; CPU or CUDA through pixi | Labels and managed backend artifacts |

Backend-specific functions stay internal rather than creating parallel result
types. Select the backend with `sn_run_annotation(method = ...)` and retrieve
every result with the same `sn_get_result(object, "annotation", name)`
contract.

The underlying methods should be cited in publications: SingleR (Aran et al.,
2019), CellTypist (Dominguez Conde et al., 2022), Seurat mapping (Hao et al.,
2021), Symphony (Kang et al., 2021), scmap (Kiselev et al., 2018), and scANVI
(Xu et al., 2021). Cell Ontology identifiers come from the bundled 2026-06-08
CL snapshot; verify current author and ontology guidelines when preparing a
submission.

The optional reference examples split the 20 real biological samples into
non-overlapping reference and query sets. They run only in the `all` profile
when the requested backend is installed.

```{r singleR-annotation, eval = run_extended && requireNamespace("SingleR", quietly = TRUE)}
reference_samples <- sort(unique(as.character(pbmc$real_sample)))[seq(1, 20, by = 2)]
reference_cells <- colnames(pbmc)[pbmc$real_sample %in% reference_samples]
reference <- subset(pbmc, cells = reference_cells)
query <- subset(pbmc, cells = setdiff(colnames(pbmc), reference_cells))

query_singleR <- sn_run_annotation(
  query,
  group_by = "seurat_clusters",
  method = "consensus",
  reference = reference,
  reference_label_by = "pbmc_consensus_label",
  consensus_reference_method = "singleR",
  species = "human",
  store_name = "kotliarov_reference_consensus"
)

reference_result <- sn_get_result(
  query_singleR,
  type = "annotation",
  name = "kotliarov_reference_consensus"
)

head(reference_result$tables$backend_predictions)
head(reference_result$tables$backend_evidence)
```

## Transfer labels from a reference

Marker tables are useful when you want to name clusters manually. When a
trusted reference already exists, `sn_transfer_labels()` follows Seurat's
anchor workflow and writes the projected label plus a confidence score back to
the query metadata. The wrapper keeps the source label and transfer settings in
`query@misc$label_transfer`, so the annotation is not just a loose metadata
column. The example below uses the consensus labels computed above as the
reference annotation and keeps the reference/query biological samples
disjoint. These are computational labels, not author-provided truth.

```{r label-transfer, eval = run_vignette}
reference_samples <- sort(unique(as.character(pbmc$real_sample)))[seq(1, 20, by = 2)]
reference_cells <- colnames(pbmc)[pbmc$real_sample %in% reference_samples]
reference <- subset(pbmc, cells = reference_cells)
query <- subset(pbmc, cells = setdiff(colnames(pbmc), reference_cells))

query <- sn_transfer_labels(
  object = query,
  reference = reference,
  label_by = "pbmc_consensus_label",
  prediction_prefix = "kotliarov_reference",
  dims = 1:15,
  verbose = FALSE
)

table(query$kotliarov_reference_label, query$real_response)
head(query[[]][, c(
  "real_sample", "real_response",
  "kotliarov_reference_label", "kotliarov_reference_score"
)])
```

The mapped query labels carry label-transfer provenance in
`query@misc$label_transfer`. To compare them with the stored consensus result,
place only those mapped query labels back into a cell-aligned metadata vector;
reference cells remain missing and are excluded from the contingency table.

```{r annotation-reference-consistency}
reference_mapped_labels <- stats::setNames(
  rep(NA_character_, ncol(pbmc)),
  colnames(pbmc)
)
reference_mapped_labels[colnames(query)] <- as.character(
  query$kotliarov_reference_label
)
pbmc$kotliarov_reference_label <- unname(
  reference_mapped_labels[colnames(pbmc)]
)

sn_plot_annotation_confusion(
  pbmc,
  truth = "kotliarov_reference_label",
  store_name = "pbmc_consensus"
)
```

The same query-first interface can use Coralysis reference mapping when the
reference was trained with `sn_run_cluster(integration_method = "coralysis")`.
By default, Shennong keeps the native Coralysis-trained SingleCellExperiment
and PCA model under `reference@misc$coralysis`, so the returned object can be
used directly as a label-transfer reference.

```{r coralysis-label-transfer, eval = run_extended && requireNamespace("Coralysis", quietly = TRUE)}
reference <- sn_run_cluster(
  reference,
  batch = "real_batch",
  integration_method = "coralysis",
  normalization_method = "seurat",
  verbose = FALSE
)

query <- sn_transfer_labels(
  object = query,
  reference = reference,
  label_by = "pbmc_consensus_label",
  method = "coralysis",
  prediction_prefix = "coral_reference",
  transfer_control = list(k.nn = 10),
  verbose = FALSE
)

table(query$coral_reference_label, query$real_response)
```

For durable handoffs, prepare a compact transfer-ready reference instead of
saving the full analysis object. Coralysis references keep the trained models,
PCA model, feature names, and selected labels while dropping the large
reference assay matrices and joint-probability tables.

The core profile also builds a compact Seurat reference in a temporary path,
then maps the same disjoint query samples back to it. No local fixture or
reference object is written into the repository.

```{r compact-label-transfer-reference, eval = run_vignette}
compact_path <- tempfile(fileext = ".qs2")
compact_reference <- sn_prepare_label_transfer_reference(
  reference,
  label_by = "pbmc_consensus_label",
  method = "seurat",
  path = compact_path,
  overwrite = TRUE
)

compact_query <- sn_transfer_labels(
  object = query,
  reference = compact_reference,
  label_by = "pbmc_consensus_label",
  prediction_prefix = "compact_reference",
  dims = 1:15,
  verbose = FALSE
)

data.frame(
  reference_cells = ncol(compact_reference),
  query_cells = ncol(compact_query),
  labels_transferred = sum(!is.na(compact_query$compact_reference_label))
)
```

## Discover and manage bundled signatures

Shennong ships a signature catalog so blocking genes, marker sets, and report
features do not need to be redefined in every analysis.

```{r signatures}
signature_index <- sn_list_signatures(species = "human")
head(signature_index)

immune_signatures <- sn_get_signatures(
  species = "human",
  category = c("mito", "ribo")
)

head(immune_signatures)
```

Custom signatures can be added, renamed, and deleted through the same API. The
example copies the shipped catalog to a temporary `.rda`, exercises the full
lifecycle with real PBMC genes, and leaves the package catalog unchanged.

```{r custom-signature, eval = run_vignette}
catalog_path <- tempfile(fileext = ".rda")
file.copy(
  system.file("data", "shennong_signature_catalog.rda", package = "Shennong"),
  catalog_path
)

added_catalog <- sn_add_signature(
  species = "human",
  path = "custom/t_cell_activation",
  genes = c("IL7R", "CCR7", "LTB"),
  catalog_path = catalog_path,
  source = "project"
)

updated_catalog <- sn_update_signature(
  species = "human",
  path = "custom/t_cell_activation",
  rename_to = "naive_t_cell",
  catalog_path = catalog_path
)

deleted_catalog <- sn_delete_signature(
  species = "human",
  path = "custom/naive_t_cell",
  catalog_path = catalog_path
)

data.frame(
  operation = c("add", "update", "delete"),
  catalog = c(added_catalog, updated_catalog, deleted_catalog)
)
file.info(catalog_path)[, c("size", "mtime")]
```

## Score programs in cells and samples

`sn_score_programs()` accepts named gene-set lists, program/gene tables, named
gene vectors, or bundled signature queries. UCell is the default for per-cell
work because its rank-based score works directly with sparse expression and is
less sensitive to dataset composition than a raw mean. AUCell is useful when
area-under-recovery-curve activity is the intended statistic, but its ranking
step is more expensive. GSVA and ssGSEA are intended primarily for samples or
aggregated groups; supply `group_by` for large single-cell objects. The `mean`
backend is dependency-free and transparent, but it is more sensitive to assay
scale and signature size.

```{r score-programs}
immune_programs <- list(
  B_cell = c("MS4A1", "CD79A", "CD37", "CD74"),
  T_cell = c("CD3D", "CD3E", "TRAC", "LTB"),
  Myeloid = c("LYZ", "FCN1", "S100A8", "CTSS")
)

pbmc <- sn_score_programs(
  pbmc,
  signatures = immune_programs,
  method = "ucell",
  assay = "RNA",
  layer = "data",
  name = "immune_programs",
  min_genes = 2
)

program_result <- sn_get_result(
  pbmc,
  type = "program_scoring",
  name = "immune_programs"
)

program_result$tables$coverage
head(program_result$tables$scores)
```

UCell and AUCell require their optional Bioconductor packages and run on CPU.
GSVA/ssGSEA require `GSVA` and run on CPU; Shennong blocks per-cell GSVA on more
than 5,000 sparse columns unless users aggregate explicitly. Cite UCell
(Andreatta and Carmona, 2021), AUCell (Aibar et al., 2017), GSVA (Hänzelmann et
al., 2013), or ssGSEA (Barbie et al., 2009) according to the selected backend.

```{r plot-programs}
sn_plot_program_activity(
  pbmc,
  name = "immune_programs",
  group_by = "seurat_clusters"
)

sn_plot_program_heatmap(
  pbmc,
  name = "immune_programs",
  group_by = "seurat_clusters"
)
```

For comparative studies, preserve the sample or patient as the inferential
unit. `sn_test_programs()` first averages cell scores within each
sample/condition/stratum, then runs the selected test; omitting `sample_by` is
explicitly marked as exploratory cell-level inference.

```{r test-programs, eval = run_vignette}
pbmc <- sn_test_programs(
  pbmc,
  score_name = "immune_programs",
  condition_by = "real_response",
  sample_by = "real_sample",
  group_by = "seurat_clusters",
  contrast = c("d0 high", "d0 low"),
  method = "wilcox",
  store_name = "immune_program_response"
)

sn_get_result(
  pbmc,
  type = "program_comparison",
  name = "immune_program_response"
)$tables$primary
```

## Run enrichment from stored marker results

`sn_enrich()` can accept a gene vector, a ranked vector, a data frame, or a
Seurat object with stored DE results. For cluster markers, the Seurat-object
path is the most reproducible because the enrichment knows which DE result it
came from.

```{r enrich-markers}
pbmc <- sn_enrich(
  x = pbmc,
  source_de_name = "cluster_markers",
  species = "human",
  database = "GOBP",
  store_name = "cluster_gobp",
  return_object = TRUE
)

pathways <- sn_get_enrichment_result(
  pbmc,
  enrichment_name = "cluster_gobp",
  top_n = 5
)

head(pathways)
```

If you already have a real enrichment table from another stage or tool, store
it with `sn_store_enrichment()` so the interpretation layer can find it. Here
the retrieved Kotliarov GO table is re-registered under a review-specific name;
no placeholder term or p-value is introduced.

```{r store-external-enrichment, eval = run_vignette}
review_terms <- pathways

pbmc <- sn_store_enrichment(
  object = pbmc,
  result = review_terms,
  store_name = "reviewed_cluster_gobp",
  analysis = "ora",
  database = "GOBP",
  species = "human",
  source_de_name = "cluster_markers",
  return_object = TRUE
)

head(sn_get_enrichment_result(pbmc, "reviewed_cluster_gobp"))
```

## Cell communication and regulatory activity

Cell communication should be run through a real backend rather than an ad hoc
ligand-receptor table. `sn_run_cell_communication()` keeps the same stored-result
pattern used by DE and enrichment. Use CellChat for a global interaction
network, NicheNet when the question is sender-to-receiver ligand activity, or
LIANA when the optional LIANA package is available and consensus scoring is
preferred.

Communication and regulatory backends are part of the `all` profile. They run
only when their upstream R package and, for NicheNet, versioned prior matrices
are already available. Sample-aware comparisons use `real_sample` as the
inferential unit and `real_response` as its sample-level phenotype.

```{r cellchat-communication, eval = run_extended && requireNamespace("CellChat", quietly = TRUE)}
pbmc <- sn_run_cell_communication(
  object = pbmc,
  method = "cellchat",
  group_by = "seurat_clusters",
  sample_by = "real_sample",
  condition_by = "real_response",
  contrast = c("d0 high", "d0 low"),
  species = "human",
  store_name = "cluster_cellchat"
)

cellchat_tbl <- sn_get_cell_communication_result(
  pbmc,
  communication_name = "cluster_cellchat"
)

head(cellchat_tbl)
sn_plot_communication(pbmc, name = "cluster_cellchat", type = "bubble", n = 25)
```

NicheNet needs its ligand-target matrix and ligand-receptor network. Shennong
requires those priors explicitly so the output remains traceable to the real
NicheNet model rather than an inferred placeholder network. Set
`SHENNONG_NICHENET_LIGAND_TARGET` and `SHENNONG_NICHENET_LR_NETWORK` to
versioned local RDS files to enable this extended chunk.

```{r nichenet-communication, eval = run_extended && requireNamespace("nichenetr", quietly = TRUE) && file.exists(Sys.getenv("SHENNONG_NICHENET_LIGAND_TARGET")) && file.exists(Sys.getenv("SHENNONG_NICHENET_LR_NETWORK"))}
ligand_target_matrix <- readRDS(Sys.getenv("SHENNONG_NICHENET_LIGAND_TARGET"))
lr_network <- readRDS(Sys.getenv("SHENNONG_NICHENET_LR_NETWORK"))
cluster_levels <- names(sort(table(pbmc$seurat_clusters), decreasing = TRUE))
receiver_markers <- marker_tbl$gene[
  as.character(marker_tbl$cluster) == cluster_levels[[3]]
]

pbmc <- sn_run_cell_communication(
  object = pbmc,
  method = "nichenet",
  group_by = "seurat_clusters",
  sender = cluster_levels[1:2],
  receiver = cluster_levels[[3]],
  geneset = receiver_markers,
  background_genes = rownames(pbmc),
  sample_by = "real_sample",
  condition_by = "real_response",
  contrast = c("d0 high", "d0 low"),
  ligand_target_matrix = ligand_target_matrix,
  lr_network = lr_network,
  store_name = "sender_receiver_nichenet"
)

head(sn_get_cell_communication_result(pbmc, "sender_receiver_nichenet"))
```

For transcription-factor and pathway activity, `sn_run_regulatory_activity()`
uses fast footprint methods through `decoupleR`. DoRothEA reports TF activity;
PROGENy reports pathway activity.

```{r regulatory-activity, eval = run_extended && requireNamespace("decoupleR", quietly = TRUE) && requireNamespace("dorothea", quietly = TRUE)}
pbmc <- sn_run_regulatory_activity(
  object = pbmc,
  method = "dorothea",
  group_by = "seurat_clusters",
  species = "human",
  store_name = "cluster_dorothea"
)

tf_activity <- sn_get_regulatory_activity_result(
  pbmc,
  activity_name = "cluster_dorothea",
  sources = c("NFKB1", "STAT1")
)

head(tf_activity)
```

## Optional reference annotation with CellTypist

CellTypist is a Python command-line tool, so Shennong keeps it explicit. When
the input is a Seurat object, predictions are written back to metadata. When the
input is a file path, Shennong returns the prediction table because there is no
object to update.

The CellTypist chunk additionally requires a local model path in
`SHENNONG_CELLTYPIST_MODEL` and an existing `celltypist` executable. The
article never downloads a model during pkgdown rendering.

```{r celltypist-seurat, eval = run_extended && nzchar(celltypist_model) && file.exists(celltypist_model) && nzchar(Sys.which("celltypist"))}
pbmc <- sn_run_celltypist(
  x = pbmc,
  model = celltypist_model,
  over_clustering = "seurat_clusters",
  majority_voting = TRUE,
  quiet = TRUE
)

grep("celltypist", colnames(pbmc[[]]), value = TRUE, ignore.case = TRUE)
```

The intended pattern is evidence first, label second: markers and pathways are
computed and stored before automated or LLM-assisted annotation consumes them.
