---
title: "Markers, signatures, pathways, and annotation"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Markers, signatures, pathways, and annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

After clustering, users usually ask three linked questions:

1. Which genes define each cluster?
2. Which known signatures or pathways explain those genes?
3. How can the evidence be reused for plots and annotation?

Shennong stores DE and enrichment results on the Seurat object so the same
tables can feed dot plots, interpretation prompts, and later reports.

## Cluster PBMC3k and find markers

```{r setup-markers}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- sn_load_data("pbmc3k")

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

pbmc <- sn_find_de(
  object = pbmc,
  analysis = "markers",
  group_by = "seurat_clusters",
  layer = "data",
  min_pct = 0.25,
  logfc_threshold = 0.25,
  store_name = "cluster_markers",
  return_object = TRUE,
  verbose = FALSE
)
```

The result is stored under `object@misc$de_results`. Retrieve it by name instead
of relying on a temporary variable from an earlier script.

```{r retrieve-markers}
marker_tbl <- sn_get_de_result(
  pbmc,
  de_name = "cluster_markers",
  top_n = 5
)

head(marker_tbl)
names(pbmc@misc$de_results)
```

## Plot stored markers without hand-copying gene lists

Because the DE result is stored, `sn_plot_dot()` can select top markers per
cluster directly.

```{r marker-dotplot}
sn_plot_dot(
  x = pbmc,
  features = "top_markers",
  de_name = "cluster_markers",
  n = 4,
  group_by = "seurat_clusters",
  palette = "RdBu",
  title = "Top PBMC3k markers"
)
```

For canonical checks, pass marker genes explicitly.

```{r canonical-dotplot}
sn_plot_dot(
  x = pbmc,
  features = c("IL7R", "CCR7", "MS4A1", "CD79A", "LYZ", "S100A8", "NKG7"),
  group_by = "seurat_clusters",
  palette = "Purples",
  direction = 1,
  title = "Canonical PBMC markers"
)
```

## Transfer labels from a reference

Marker tables are useful when you want to name clusters manually. When a
trusted reference already exists, `sn_transfer_labels()` follows Seurat's
anchor workflow and writes the projected label plus a confidence score back to
the query metadata. The wrapper keeps the source label and transfer settings in
`query@misc$label_transfer`, so the annotation is not just a loose metadata
column.

```{r label-transfer, eval = FALSE}
reference <- pbmc
query <- pbmc

reference$cell_type <- reference$seurat_clusters

query <- sn_transfer_labels(
  object = query,
  reference = reference,
  label_col = "cell_type",
  prediction_prefix = "pbmc_reference",
  dims = 1:15,
  verbose = FALSE
)

table(query$pbmc_reference_label)
head(query[[]][, c("pbmc_reference_label", "pbmc_reference_score")])
```

## Discover and manage bundled signatures

Shennong ships a signature catalog so blocking genes, marker sets, and report
features do not need to be redefined in every analysis.

```{r signatures}
signature_index <- sn_list_signatures(species = "human")
head(signature_index)

immune_signatures <- sn_get_signatures(
  species = "human",
  category = c("mito", "ribo")
)

head(immune_signatures)
```

Custom signatures can be added, renamed, and deleted through the same API. Use
a project-specific catalog path when you do not want to modify the package-level
catalog.

```{r custom-signature, eval = FALSE}
catalog_path <- "config/signatures/pbmc_signatures.csv"

sn_add_signature(
  species = "human",
  path = "custom/t_cell_activation",
  genes = c("IL7R", "CCR7", "LTB"),
  catalog_path = catalog_path,
  source = "project"
)

sn_update_signature(
  species = "human",
  path = "custom/t_cell_activation",
  rename_to = "custom/naive_t_cell",
  catalog_path = catalog_path
)

sn_delete_signature(
  species = "human",
  path = "custom/naive_t_cell",
  catalog_path = catalog_path
)
```

## Run enrichment from stored marker results

`sn_enrich()` can accept a gene vector, a ranked vector, a data frame, or a
Seurat object with stored DE results. For cluster markers, the Seurat-object
path is the most reproducible because the enrichment knows which DE result it
came from.

```{r enrich-markers}
pbmc <- sn_enrich(
  x = pbmc,
  source_de_name = "cluster_markers",
  species = "human",
  database = "GOBP",
  store_name = "cluster_gobp",
  return_object = TRUE
)

pathways <- sn_get_enrichment_result(
  pbmc,
  enrichment_name = "cluster_gobp",
  top_n = 5
)

head(pathways)
```

If you already have an enrichment table from another tool, store it with
`sn_store_enrichment()` so the interpretation layer can find it.

```{r store-external-enrichment, eval = FALSE}
external_terms <- data.frame(
  cluster = "0",
  ID = "GO:0006955",
  Description = "immune response",
  p.adjust = 0.001,
  geneID = "IL7R/CCR7/LTB"
)

pbmc <- sn_store_enrichment(
  object = pbmc,
  result = external_terms,
  store_name = "external_gobp",
  analysis = "ora",
  database = "GOBP",
  species = "human",
  source_de_name = "cluster_markers",
  return_object = TRUE
)
```

## Optional reference annotation with CellTypist

CellTypist is a Python command-line tool, so Shennong keeps it explicit. When
the input is a Seurat object, predictions are written back to metadata. When the
input is a file path, Shennong returns the prediction table because there is no
object to update.

```{r celltypist-seurat, eval = FALSE}
pbmc <- sn_run_celltypist(
  x = pbmc,
  model = "Immune_All_Low.pkl",
  over_clustering = "seurat_clusters",
  majority_voting = TRUE,
  quiet = TRUE
)

grep("Immune_All_Low", colnames(pbmc[[]]), value = TRUE)
```

```{r celltypist-path, eval = FALSE}
prediction_tbl <- sn_run_celltypist(
  x = "pbmc3k_counts.csv",
  model = "Immune_All_Low.pkl",
  over_clustering = "obs_cluster",
  quiet = TRUE
)

head(prediction_tbl)
```

The intended pattern is evidence first, label second: markers and pathways are
computed and stored before automated or LLM-assisted annotation consumes them.
