---
title: "Annotation, markers, and pathways workflow"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotation, markers, and pathways workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") ||
  identical(Sys.getenv("IN_PKGDOWN"), "true")

required_pkgs <- c("Seurat", "clusterProfiler", "org.Hs.eg.db")
run_vignette <- run_vignette && all(vapply(
  required_pkgs,
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  eval = run_vignette
)
```

This article covers the main biological-interpretation stage after clustering:

- inspect bundled signatures
- score marker structure with differential expression
- visualize top markers
- run ORA or GSEA with `sn_enrich()`
- optionally bring in reference-based annotation with `sn_run_celltypist()`

```{r setup}
library(Shennong)
library(dplyr)
library(knitr)
library(Seurat)

if (!exists("pbmc_small", inherits = FALSE)) {
  try(data("pbmc_small", package = "Shennong", envir = environment()), silent = TRUE)
}
if (!exists("pbmc_small", inherits = FALSE) && file.exists(file.path("data", "pbmc_small.rda"))) {
  load(file.path("data", "pbmc_small.rda"))
}
```

```{r prepare-annotation-demo, include = FALSE}
pbmc_clustered <- suppressWarnings(
  sn_run_cluster(
    pbmc_small,
    normalization_method = "seurat",
    nfeatures = 800,
    npcs = 20,
    dims = 1:15,
    resolution = 0.4,
    verbose = FALSE
  )
)

signature_tbl <- sn_list_signatures(species = "human")
stress_genes <- sn_get_signatures(
  species = "human",
  category = c("Programs/HeatShock", "Compartments/Mito")
)

pbmc_clustered <- suppressWarnings(
  sn_find_de(
    pbmc_clustered,
    analysis = "markers",
    group_by = "seurat_clusters",
    layer = "data",
    store_name = "cluster_markers",
    return_object = TRUE,
    verbose = FALSE
  )
)

marker_tbl <- sn_get_de_result(
  pbmc_clustered,
  de_name = "cluster_markers",
  top_n = 3
)

enrichment_object <- tryCatch(
  sn_enrich(
    x = pbmc_clustered,
    source_de_name = "cluster_markers",
    gene_clusters = gene ~ cluster,
    database = c("H", "GOBP"),
    species = "human",
    pvalue_cutoff = 0.05,
    store_name = "cluster_pathways",
    return_object = TRUE
  ),
  error = function(...) pbmc_clustered
)

stored_pathways <- tryCatch(
  sn_get_enrichment_result(
    enrichment_object,
    enrichment_name = "cluster_pathways",
    top_n = 5
  ),
  error = function(...) NULL
)
```

## 1. Start from bundled signatures

Bundled signatures provide a stable set of marker programs and technical gene
sets for filtering, scoring, or interpretation.

```{r}
knitr::kable(head(signature_tbl[, c("path", "n_genes")], 8))
length(stress_genes)
```

## 2. Find marker genes

Cluster markers are usually the first evidence layer for annotation.

```{r}
knitr::kable(marker_tbl)
```

## 3. Visualize top markers

`sn_plot_dot()` can reuse the stored DE result directly, which avoids manual
gene selection in routine cluster review.

```{r}
sn_plot_dot(
  pbmc_clustered,
  features = "top_markers",
  de_name = "cluster_markers",
  n = 3
)
```

## 4. Run pathway analysis

`sn_enrich()` now supports grouped ORA and ranked GSEA through the same
`gene_clusters` formula interface. It can also query multiple databases in one
call.

```{r}
if (!is.null(stored_pathways)) {
  knitr::kable(stored_pathways, digits = 4)
}
```

Grouped ORA uses a categorical right-hand side:

```r
ora_result <- sn_enrich(
  x = pbmc_clustered,
  source_de_name = "cluster_markers",
  gene_clusters = gene ~ cluster,
  database = c("H", "C2:CP:REACTOME"),
  species = "human"
)
```

Ranked GSEA uses a numeric right-hand side:

```r
gsea_result <- sn_enrich(
  marker_table,
  gene_clusters = gene ~ avg_log2FC,
  database = "GOBP",
  species = "human"
)
```

## 5. Optional reference-based annotation

If CellTypist is configured in the local environment, Shennong can add
reference labels to the Seurat object:

```r
pbmc_annotated <- sn_run_celltypist(
  pbmc_clustered,
  model = "Immune_All_Low.pkl"
)
```

That annotation can then be reused as the `label` input for downstream
integration metrics or cluster summaries.
