---
title: "Validated AutoZyme Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validated AutoZyme Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.8
)
benchmark_path <- system.file(
  "benchmarks",
  "single-cell-autozyme-benchmark.json",
  package = "Shennong"
)
if (!nzchar(benchmark_path)) {
  benchmark_path <- file.path(
    "inst", "benchmarks", "single-cell-autozyme-benchmark.json"
  )
}
benchmark <- jsonlite::read_json(benchmark_path, simplifyVector = TRUE)
results <- as.data.frame(benchmark$summary, stringsAsFactors = FALSE)
operation_labels <- c(
  merge = "Seurat merge",
  joinlayers = "JoinLayers",
  scdblfinder = "scDblFinder"
)
results$workflow <- unname(operation_labels[results$operation])
merge_result <- results[results$operation == "merge", , drop = FALSE]
join_result <- results[results$operation == "joinlayers", , drop = FALSE]
doublet_result <- results[results$operation == "scdblfinder", , drop = FALSE]
merge_speedup <- sprintf("%.2fx", merge_result$speedup)
merge_rss_increase <- sprintf(
  "%.1f%%", -merge_result$peak_rss_reduction_percent
)
join_speedup <- sprintf("%.2fx", join_result$speedup)
join_rss_reduction <- sprintf(
  "%.1f%%", join_result$peak_rss_reduction_percent
)
doublet_speedup <- sprintf("%.2fx", doublet_result$speedup)
doublet_rss_reduction <- sprintf(
  "%.1f%%", doublet_result$peak_rss_reduction_percent
)
```

## What was measured

This is a reproducible benchmark of Shennong's vendored exact-scoped
`seurat_merge`, `seurat_joinlayers`, and `scdblfinder` patches. It uses the
validated 2,000-cell Kotliarov PBMC CITE-seq subset distributed by Bioconductor
`scRNAseq`. The local fixture retains all 32,738 RNA features and 20 biological
samples. Its SHA-256 is recorded in the report; the data itself remains under
the ignored local real-data root and is not copied into the package.

Each baseline and accelerated condition ran three times in a fresh R process.
The order alternated between baseline-first and accelerated-first. The table
uses median operation time and median whole-worker peak resident memory. The
memory number therefore includes loading Shennong, reading the fixture, and
preparing identical inputs; it is a conservative end-to-end process measure,
not an allocation profile of one function body.

```{r benchmark-table}
display <- data.frame(
  Workflow = results$workflow,
  `Baseline (s)` = round(results$baseline_seconds, 3),
  `Accelerated (s)` = round(results$accelerated_seconds, 3),
  Speedup = sprintf("%.2fx", results$speedup),
  `Baseline peak RSS (MB)` = round(results$baseline_peak_rss_mb, 1),
  `Accelerated peak RSS (MB)` = round(results$accelerated_peak_rss_mb, 1),
  `Peak RSS change` = sprintf(
    "%+.1f%%",
    -results$peak_rss_reduction_percent
  ),
  `Exact output` = ifelse(results$output_identical, "yes", "no"),
  check.names = FALSE
)
knitr::kable(display, align = c("l", rep("r", 6), "c"))
```

```{r benchmark-figure, echo=FALSE}
plot_data <- rbind(
  data.frame(
    workflow = results$workflow,
    metric = "Elapsed time (seconds)",
    condition = "Baseline",
    value = results$baseline_seconds
  ),
  data.frame(
    workflow = results$workflow,
    metric = "Elapsed time (seconds)",
    condition = "Accelerated",
    value = results$accelerated_seconds
  ),
  data.frame(
    workflow = results$workflow,
    metric = "Peak worker RSS (MB)",
    condition = "Baseline",
    value = results$baseline_peak_rss_mb
  ),
  data.frame(
    workflow = results$workflow,
    metric = "Peak worker RSS (MB)",
    condition = "Accelerated",
    value = results$accelerated_peak_rss_mb
  )
)
plot_data$workflow <- factor(
  plot_data$workflow,
  levels = unname(operation_labels)
)
plot_data$condition <- factor(
  plot_data$condition,
  levels = c("Baseline", "Accelerated")
)

ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = workflow, y = value, fill = condition)
) +
  ggplot2::geom_col(position = "dodge", width = 0.72) +
  ggplot2::facet_wrap(~metric, scales = "free_y") +
  ggplot2::scale_fill_manual(values = c("#8C8C8C", "#0072B2")) +
  ggplot2::labs(x = NULL, y = NULL, fill = NULL) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position = "top",
    panel.grid.major.x = ggplot2::element_blank()
  )
```

## Interpretation

All three accelerated outputs were byte-stable under the benchmark's canonical
snapshots, and every worker reported its patch inactive after the operation.

- `merge()` improved from `r round(merge_result$baseline_seconds, 3)` to
  `r round(merge_result$accelerated_seconds, 3)` seconds
  (`r merge_speedup`). For this small end-to-end worker, peak RSS increased by
  `r merge_rss_increase`. This is a real regression for this fixture and measurement
  boundary, even though larger patch-level fixtures previously showed lower
  peak memory. The patch is therefore a speed win here, not a universal memory
  win.
- `JoinLayers()` improved by `r join_speedup` and reduced whole-worker peak RSS
  by `r join_rss_reduction`.
- `sn_find_doublets()` with the default scDblFinder path improved by
  `r doublet_speedup` and reduced peak RSS by `r doublet_rss_reduction`.

These numbers describe one machine, package set, and real-data fixture. Use the
included script on the intended deployment host before making capacity claims.

## Compatibility boundaries

The benchmark does not widen the patches' verified scope:

- `seurat_merge` replaces only `SeuratObject::merge.Assay5`; top-level Seurat
  merge benefits only through normal dispatch. Unsupported classes and call
  shapes use captured upstream behavior.
- `seurat_joinlayers` accelerates an explicit counts-to-counts join on a plain
  `Assay5`. Unsupported layers, renamed outputs, custom subclasses, or mixed
  backends fall back upstream.
- `scdblfinder` is limited to a target-function-fingerprint-compatible default
  call on exact in-memory `dgCMatrix` counts with at most 33,000 cells. Grouped
  and BPCells-backed Shennong workflows remain outside this fast path.

Check eligibility before a run:

```{r check-patches}
Shennong::sn_check_autozyme(c(
  "seurat_merge",
  "seurat_joinlayers",
  "scdblfinder"
))[, c(
  "patch", "installed_version", "version_match",
  "bundled_by_shennong", "eligible", "active"
)]
```

## Reproduce the benchmark

The heavy benchmark is deliberately not executed during pkgdown builds. Point
the script at the locally prepared fixture and choose an output path:

```{r reproduce, eval=FALSE}
system2(
  file.path(R.home("bin"), "Rscript"),
  c(
    "scripts/real-data/benchmark-single-cell-autozyme.R",
    "--root", "data-local/pkgdown-real",
    "--output", "inst/benchmarks/single-cell-autozyme-benchmark.json",
    "--repetitions", "3"
  )
)
```

The script validates the fixture manifest before running, launches every
condition in a new process with GNU `time`, checks exact output fingerprints,
checks patch rollback, and writes both JSON evidence and a compact CSV summary.

Benchmark generated at `r benchmark$generated_at_utc` on
`r benchmark$environment$cpu`, using R `r benchmark$environment$r`, Shennong
`r benchmark$environment$shennong`, AutoZyme
`r benchmark$environment$autozyme`, SeuratObject
`r benchmark$environment$seurat_object`, and scDblFinder
`r benchmark$environment$scdblfinder`.
