---
title: "Validated AutoZyme Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validated AutoZyme Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.8
)
benchmark_path <- system.file(
  "benchmarks",
  "single-cell-autozyme-benchmark.json",
  package = "Shennong"
)
if (!nzchar(benchmark_path)) {
  candidates <- c(
    file.path("inst", "benchmarks", "single-cell-autozyme-benchmark.json"),
    file.path("..", "inst", "benchmarks", "single-cell-autozyme-benchmark.json"),
    file.path(
      dirname(knitr::current_input(dir = TRUE)),
      "..", "inst", "benchmarks", "single-cell-autozyme-benchmark.json"
    )
  )
  benchmark_path <- candidates[file.exists(candidates)][[1L]]
}
benchmark <- jsonlite::read_json(benchmark_path, simplifyVector = TRUE)
results <- as.data.frame(benchmark$summary, stringsAsFactors = FALSE)
operation_labels <- c(
  merge = "Seurat merge",
  joinlayers = "JoinLayers",
  scdblfinder = "scDblFinder"
)
results$workflow <- unname(operation_labels[results$operation])
merge_result <- results[results$operation == "merge", , drop = FALSE]
join_result <- results[results$operation == "joinlayers", , drop = FALSE]
doublet_result <- results[results$operation == "scdblfinder", , drop = FALSE]
merge_speedup <- sprintf("%.2fx", merge_result$speedup)
merge_rss_increase <- sprintf(
  "%.1f%%", -merge_result$peak_rss_reduction_percent
)
join_speedup <- sprintf("%.2fx", join_result$speedup)
join_rss_reduction <- sprintf(
  "%.1f%%", join_result$peak_rss_reduction_percent
)
doublet_speedup <- sprintf("%.2fx", doublet_result$speedup)
doublet_rss_reduction <- sprintf(
  "%.1f%%", doublet_result$peak_rss_reduction_percent
)
```

## What was measured

This is a reproducible benchmark of Shennong's vendored exact-scoped
`seurat_merge`, `seurat_joinlayers`, and `scdblfinder` patches. It uses the
validated 2,000-cell Kotliarov PBMC CITE-seq subset distributed by Bioconductor
`scRNAseq`. The local fixture retains all 32,738 RNA features and 20 biological
samples. Its SHA-256 is recorded in the report; the data itself remains under
the ignored local real-data root and is not copied into the package.

Each baseline and accelerated condition ran three times in a fresh R process.
The order alternated between baseline-first and accelerated-first. The table
uses median operation time and median whole-worker peak resident memory. The
memory number therefore includes loading Shennong, reading the fixture, and
preparing identical inputs; it is a conservative end-to-end process measure,
not an allocation profile of one function body.

```{r benchmark-table}
display <- data.frame(
  Workflow = results$workflow,
  `Baseline (s)` = round(results$baseline_seconds, 3),
  `Accelerated (s)` = round(results$accelerated_seconds, 3),
  Speedup = sprintf("%.2fx", results$speedup),
  `Baseline peak RSS (MB)` = round(results$baseline_peak_rss_mb, 1),
  `Accelerated peak RSS (MB)` = round(results$accelerated_peak_rss_mb, 1),
  `Peak RSS change` = sprintf(
    "%+.1f%%",
    -results$peak_rss_reduction_percent
  ),
  `Exact output` = ifelse(results$output_identical, "yes", "no"),
  check.names = FALSE
)
knitr::kable(display, align = c("l", rep("r", 6), "c"))
```

```{r benchmark-figure, echo=FALSE}
plot_data <- rbind(
  data.frame(
    workflow = results$workflow,
    metric = "Elapsed time (seconds)",
    condition = "Baseline",
    value = results$baseline_seconds
  ),
  data.frame(
    workflow = results$workflow,
    metric = "Elapsed time (seconds)",
    condition = "Accelerated",
    value = results$accelerated_seconds
  ),
  data.frame(
    workflow = results$workflow,
    metric = "Peak worker RSS (MB)",
    condition = "Baseline",
    value = results$baseline_peak_rss_mb
  ),
  data.frame(
    workflow = results$workflow,
    metric = "Peak worker RSS (MB)",
    condition = "Accelerated",
    value = results$accelerated_peak_rss_mb
  )
)
plot_data$workflow <- factor(
  plot_data$workflow,
  levels = unname(operation_labels)
)
plot_data$condition <- factor(
  plot_data$condition,
  levels = c("Baseline", "Accelerated")
)

ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = workflow, y = value, fill = condition)
) +
  ggplot2::geom_col(position = "dodge", width = 0.72) +
  ggplot2::facet_wrap(~metric, scales = "free_y") +
  ggplot2::scale_fill_manual(values = c("#8C8C8C", "#0072B2")) +
  ggplot2::labs(x = NULL, y = NULL, fill = NULL) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position = "top",
    panel.grid.major.x = ggplot2::element_blank()
  )
```

## Interpretation

All three accelerated outputs were byte-stable under the benchmark's canonical
snapshots, and every worker reported its patch inactive after the operation.

- `merge()` improved from `r round(merge_result$baseline_seconds, 3)` to
  `r round(merge_result$accelerated_seconds, 3)` seconds
  (`r merge_speedup`). For this small end-to-end worker, peak RSS increased by
  `r merge_rss_increase`. This is a real regression for this fixture and measurement
  boundary, even though larger patch-level fixtures previously showed lower
  peak memory. The patch is therefore a speed win here, not a universal memory
  win.
- `JoinLayers()` improved by `r join_speedup` and reduced whole-worker peak RSS
  by `r join_rss_reduction`.
- `sn_find_doublets()` with the default scDblFinder path improved by
  `r doublet_speedup` and reduced peak RSS by `r doublet_rss_reduction`.

These numbers describe one machine, package set, and real-data fixture. Use the
included script on the intended deployment host before making capacity claims.

## Three-arm 30,000-cell workflow audit

The newer `benchmark-autozyme-workflow-patches.R` runner separates three
questions that a two-arm patch benchmark cannot distinguish:

1. Does the Shennong wrapper agree with the direct upstream public API?
2. Does enabling the patch preserve the Shennong result?
3. Does the complete Shennong call become faster on a realistic input?

The 2026-08-21 maintainer-host audit used 30,000 cells from three independent
public PBMC captures (10x PBMC 10k v3, TENxPBMCData PBMC 33k, and 10x Fresh
PBMC 68k donor A), with 10,000 cells sampled from each capture at seed 717.
The derived fixture contains 20,453 common gene symbols and has SHA-256
`20b884fb6c8f20c932ae0eb3e56484e40089386208653156b447439032945013`.
Every arm ran in a fresh process; three repetitions used rotating arm order.
The benchmark library contained the exact AutoZyme revision
`8fc2e9c3a7f70302f97589aaa9b0395dcf86f9bc`.

| Workflow target | Direct upstream (s) | Shennong off (s) | Shennong on (s) | Off/on | Off RSS (MiB) | On RSS (MiB) | Result |
|:--|--:|--:|--:|--:|--:|--:|:--|
| LISI | 6.186 | 6.251 | 0.686 | 9.11x | 3988.7 | 3988.5 | equal |
| UCell | 57.857 | 57.678 | 1.734 | 33.26x | 4073.3 | 2575.1 | equal |
| Seurat NormalizeData | 2.694 | 3.013 | 0.671 | 4.49x | 2267.6 | 2131.2 | within `1e-8` |
| Assay5 merge | 0.621 | 0.627 | 0.206 | 3.04x | 2577.1 | 2343.3 | exact |
| JoinLayers | 1.700 | 1.819 | 1.078 | 1.69x | 3566.3 | 2808.2 | exact |

The scDblFinder 30,000-cell audit is run separately because each unaccelerated
arm takes roughly 100 seconds. Its pinned-source, three-order median improved
from 100.15 to 57.29 seconds (`1.75x`), while peak worker RSS increased from
about 4,038 to 4,327 MiB. That patch is therefore a time-for-memory tradeoff at
this scale.

The NormalizeData row was rerun after Shennong learned to bypass temporary
matrix snapshots when the input is already a single standard `counts` layer.
That removed most wrapper-only preparation cost: the unaccelerated wrapper is
now about 12% above the direct call on this host, rather than about 57% in the
pre-fix main report.

The repeated GO annotation-cache scope was only `1.04x` in a one-repetition
1,000-cell query preparation test (28.68 versus 27.50 seconds) and should not be
presented as a general ORA/GSEA kernel acceleration. The clusterProfiler patch
targets annotation retrieval only. The current `sn_enrich()` GSEA backend is
enrichit, so the separate `fgsea` patch has no call intersection and is not
benchmarked as a Shennong enrichment acceleration.

The current Coralysis patch is also not an end-to-end `sn_run_cluster()` speed
claim. Although its target is `RunParallelDivisiveICP`, the released fast-path
guard requires the literal metadata field `batch.label = "benchmark_batch"`,
`build.train.set = FALSE`, empty `build.train.params`, and a fixed ICP
configuration. Shennong passes the caller's real batch-column name and generated
training parameters, so this workflow currently falls back to upstream even
when the patch is active. The PBMC integration timing already shows Coralysis as
a substantial clustering stage; widening that patch requires a new parity
contract rather than changing Shennong's scientific defaults to fit a benchmark
guard.

These host measurements are evidence for the tested call shapes, not a claim
that activation proves an internal fast-path hit. The report records target
intersection and scoped activation separately and leaves `fast_path_hit`
unknown because the current AutoZyme patch API exposes no per-call hit counter.
Seurat 5.5.1 and UCell 2.17.0 are also explicit upstream-version-drift cases;
their results pass this fixture but remain outside strict admission until their
contracts are promoted for those versions.

### Four gates between activation and acceleration

Read the runtime evidence as four separate events:

1. **Activation:** a replacement was installed in the namespace for a bounded
   scope.
2. **Target intersection:** the workflow actually called the replaced upstream
   symbol.
3. **Guard accepted:** the operation, input class, arguments, and dependency
   versions matched the validated fast-path envelope.
4. **Fast-path hit:** runtime evidence confirmed that the accelerated kernel,
   rather than its fallback, executed.

Failure at any later gate makes an earlier `active = TRUE` insufficient as an
acceleration claim. JoinLayers and Coralysis are current concrete examples:
their patches can be active and their symbols can intersect, while Shennong's
real default call shapes fail benchmark-specific guards and fall back. There
are also unsafe false-positive cases, not only conservative fallback cases. An
older broad Seurat `FindIntegrationAnchors` replacement lacked a `reduction`
guard, so an RPCA request could enter code written for a CCA-oriented path. The
tradeSeq patch globally replaced `mgcv` behavior without an exact dependency
guard.

Shennong's defense is layered: source and upstream versions are checked in
strict mode, Seurat activation is operation-specific, and incompatible calls
run inside a temporary disabled scope even if a user activated a broader patch
earlier. A patch is promoted from explicit-only only after the direct upstream,
Shennong-off, and Shennong-on arms agree for the actual workflow call shape.

## Automatic workflow boundary

Benchmark coverage and automatic activation are different claims. The current
automatic patch set is exactly `cellchat`, `clusterprofiler`, `lisi`,
`nichenetr`, `scdblfinder`, `seurat`, `seurat_merge`, `soupx`, and `ucell`.
Within that set, clusterProfiler covers annotation caching rather than the
ORA/GSEA statistical kernel, and the broad Seurat patch is admitted
automatically only for NormalizeData. Default scDblFinder, owned Assay5 merge,
SoupX, UCell, LISI, and the guarded communication paths retain their own input
and version envelopes.

Coralysis, standalone decontX, broad Seurat targets beyond NormalizeData,
JoinLayers, tradeSeq, and WGCNA are explicit-only. In particular, the
JoinLayers timing above does not make `seurat_joinlayers` an automatic
Shennong patch. `sn_check_autozyme()` reports `automatic_policy`, eligibility,
workflow intersection, and fast-path status separately.

## Compatibility boundaries

The benchmark does not widen the patches' verified scope:

- `seurat_merge` replaces only `SeuratObject::merge.Assay5`; top-level Seurat
  merge benefits only through normal dispatch. Unsupported classes and call
  shapes use captured upstream behavior.
- `seurat_joinlayers` accelerates an explicit counts-to-counts join on a plain
  `Assay5`. Unsupported layers, renamed outputs, custom subclasses, or mixed
  backends fall back upstream.
- `scdblfinder` is limited to a target-function-fingerprint-compatible default
  call on exact in-memory `dgCMatrix` counts with at most 33,000 cells. Grouped
  and BPCells-backed Shennong workflows remain outside this fast path.

Check eligibility before a run:

```{r check-patches}
Shennong::sn_check_autozyme(c(
  "seurat_merge",
  "seurat_joinlayers",
  "scdblfinder"
))[, c(
  "patch", "automatic_policy", "installed_version", "version_match",
  "workflow_intersection", "bundled_by_shennong", "eligible", "active",
  "fast_path_status", "fast_path_hit"
)]
```

`fast_path_hit` remains `NA` when the provider cannot expose a runtime hit
counter; activation or eligibility alone is never rewritten as a confirmed hit.

The generated public-API parameter inventory plus clustering and enrichment
method matrices are static admission plans. They ensure that selectors,
dispatch cells and required cases are not
silently omitted, but a matrix row is not a successful benchmark or backend
conformance execution. Only the fresh-process direct/off/on evidence reported
above supports the corresponding performance and parity claims.

## Reproduce the benchmark

The heavy benchmark is deliberately not executed during pkgdown builds. Point
the script at the locally prepared fixture and choose an output path:

```{r reproduce, eval=FALSE}
system2(
  file.path(R.home("bin"), "Rscript"),
  c(
    "scripts/real-data/benchmark-single-cell-autozyme.R",
    "--root", "data-local/pkgdown-real",
    "--output", "inst/benchmarks/single-cell-autozyme-benchmark.json",
    "--repetitions", "3"
  )
)
```

The script validates the fixture manifest before running, launches every
condition in a new process with GNU `time`, checks exact output fingerprints,
checks patch rollback, and writes both JSON evidence and a compact CSV summary.

To reproduce the three-arm 30,000-cell audit, first prepare the governed local
fixture and then run the workflow patch runner:

```{r reproduce-workflow, eval=FALSE}
system2(
  file.path(R.home("bin"), "Rscript"),
  c("scripts/real-data/prepare-autozyme-pbmc30k.R")
)

system2(
  file.path(R.home("bin"), "Rscript"),
  c(
    "scripts/real-data/benchmark-autozyme-workflow-patches.R",
    "--input", "data-local/autozyme-benchmark/pbmc-captures-30k.qs2",
    "--output", "data-local/autozyme-benchmark/pbmc30k.json",
    "--sample-by", "capture",
    "--repetitions", "3"
  )
)
```

Benchmark generated at `r benchmark$generated_at_utc` on
`r benchmark$environment$cpu`, using R `r benchmark$environment$r`, Shennong
`r benchmark$environment$shennong`, AutoZyme
`r benchmark$environment$autozyme`, SeuratObject
`r benchmark$environment$seurat_object`, and scDblFinder
`r benchmark$environment$scdblfinder`.
