---
title: "Bulk Deconvolution Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bulk Deconvolution Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(Shennong)
library(Seurat)
```

This workflow shows how to combine a single-cell reference with bulk RNA-seq
mixtures for cell-composition inference.

Shennong currently supports two backends:

- `BayesPrism` for local Bayesian deconvolution
- `CIBERSORTx` for local container-based deconvolution

## 1. Build a single-cell reference

```{r}
data("pbmc_small", package = "Shennong")
ref <- pbmc_small
ref$cell_type <- rep(c("T_like", "B_like", "Myeloid_like"), length.out = ncol(ref))
```

## 2. Create a mock bulk matrix

```{r}
counts <- SeuratObject::LayerData(ref, assay = "RNA", layer = "counts")
bulk <- cbind(
  sample_a = rowSums(counts[, 1:40, drop = FALSE]),
  sample_b = rowSums(counts[, 41:ncol(counts), drop = FALSE])
)
bulk[1:5, 1:2]
```

## 3. Prepare a local CIBERSORTx run

```{r}
bundle <- sn_deconvolve_bulk(
  ref,
  bulk = bulk,
  method = "cibersortx",
  cell_type_col = "cell_type",
  layer = "counts",
  outdir = tempdir(),
  prefix = "pbmc_demo",
  cibersortx_email = "user@example.org",
  cibersortx_token = "replace-with-real-token",
  cibersortx_dry_run = TRUE,
  return_object = FALSE
)

bundle$files
bundle$artifacts$commands
```

In `dry_run` mode, Shennong writes the local input files and returns the exact
container commands. To run CIBERSORTx locally, provide real credentials and set
`cibersortx_dry_run = FALSE`.

```r
ref <- sn_deconvolve_bulk(
  ref,
  bulk = bulk,
  method = "cibersortx",
  cell_type_col = "cell_type",
  layer = "counts",
  cibersortx_email = Sys.getenv("SHENNONG_CIBERSORTX_EMAIL"),
  cibersortx_token = Sys.getenv("SHENNONG_CIBERSORTX_TOKEN"),
  store_name = "pbmc_cibersortx"
)
```

If you already have a local `CIBERSORTx_*_Results.txt` file, you can still
import it directly through `cibersortx_result = ...`.

## 4. Run BayesPrism locally

BayesPrism is an optional GitHub dependency:

```r
remotes::install_github("Danko-Lab/BayesPrism/BayesPrism")
```

Then run:

```r
ref <- sn_deconvolve_bulk(
  ref,
  bulk = bulk,
  method = "bayesprism",
  cell_type_col = "cell_type",
  layer = "counts",
  store_name = "pbmc_bayesprism"
)
```

## 5. Retrieve stored deconvolution results

```r
sn_list_results(ref)
sn_get_deconvolution_result(ref, "pbmc_cibersortx")
```
