---
title: "Bulk deconvolution from a real Kotliarov PBMC reference"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bulk deconvolution from a real Kotliarov PBMC reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
run_vignette <- identical(tolower(Sys.getenv("SHENNONG_RUN_VIGNETTES")), "true")
real_profile <- tolower(Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core"))
root_candidates <- unique(c(
  Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = ""),
  file.path(getwd(), "data-local", "pkgdown-real"),
  file.path(getwd(), "..", "data-local", "pkgdown-real")
))
root_candidates <- root_candidates[nzchar(root_candidates)]
existing_roots <- root_candidates[dir.exists(root_candidates)]
real_root <- if (length(existing_roots)) existing_roots[[1]] else root_candidates[[1]]
pbmc_path <- file.path(real_root, "single-cell", "kotliarov_pbmc.qs2")
core_ready <- run_vignette && file.exists(pbmc_path)
bayesprism_ready <- core_ready && identical(real_profile, "all") &&
  requireNamespace("BayesPrism", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 5
)
```

This article uses real UMI counts from the public Kotliarov PBMC CITE-seq
cohort. The bulk columns are biological-sample pseudobulks formed by summing
cells within `real_sample`; they are not mixtures assembled from hand-picked
clusters. Because the reference cells and mixtures come from the same cohort,
this is a technical contract test rather than an independent biological
validation.

```{r runtime-status}
knitr::kable(data.frame(
  workflow = c("CIBERSORTx input contract", "BayesPrism deconvolution"),
  mode = c("core dry run", "extended backend"),
  status = c(
    if (core_ready) "executed below; no backend result claimed" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture missing",
    if (bayesprism_ready) "executed below" else if (!identical(real_profile, "all")) "disabled: requires SHENNONG_REAL_PROFILE=all" else if (!requireNamespace("BayesPrism", quietly = TRUE)) "dependency missing: BayesPrism" else "fixture missing"
  ),
  check.names = FALSE
))
```

## Build a real reference and biological-sample pseudobulk

```{r prepare-deconvolution, eval=core_ready}
library(Shennong)
library(Seurat)

pbmc <- qs2::qs_read(pbmc_path)
pbmc <- sn_run_cluster(
  pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.5,
  species = "human",
  verbose = FALSE
)
pbmc$cell_state <- paste0("cluster_", pbmc$seurat_clusters)

counts <- SeuratObject::LayerData(pbmc, assay = "RNA", layer = "counts")
sample_cells <- split(colnames(pbmc), pbmc$real_sample)
pseudobulk_counts <- vapply(
  sample_cells,
  function(cells) Matrix::rowSums(counts[, cells, drop = FALSE]),
  numeric(nrow(counts))
)
rownames(pseudobulk_counts) <- rownames(counts)

pbmc_metadata <- pbmc[[]]
sample_design <- do.call(rbind, lapply(
  split(seq_len(nrow(pbmc_metadata)), pbmc_metadata$real_sample),
  function(index) data.frame(
    real_sample = pbmc_metadata$real_sample[index[[1]]],
    real_response = unique(pbmc_metadata$real_response[index]),
    observed_batches = paste(sort(unique(pbmc_metadata$real_batch[index])), collapse = "+"),
    cells = length(index),
    row.names = pbmc_metadata$real_sample[index[[1]]]
  )
))
sample_design <- sample_design[colnames(pseudobulk_counts), , drop = FALSE]

data.frame(
  samples = ncol(pseudobulk_counts),
  genes = nrow(pseudobulk_counts),
  cells = ncol(pbmc),
  clusters = length(unique(pbmc$cell_state)),
  response_groups = paste(sort(unique(sample_design$real_response)), collapse = " / ")
)
```

## Validate a CIBERSORTx export without claiming fractions

The dry run writes only temporary input files and returns redacted container
commands. It proves that the real reference labels and 20 pseudobulk columns
satisfy the interface. It does **not** create or import a deconvolution result.

```{r cibersortx-dry-run, eval=core_ready}
cibersortx_bundle <- sn_deconvolve_bulk(
  x = pbmc,
  bulk = pseudobulk_counts,
  method = "cibersortx",
  cell_type_by = "cell_state",
  layer = "counts",
  outdir = file.path(tempdir(), "kotliarov-cibersortx"),
  prefix = "kotliarov_pseudobulk",
  cibersortx_email = "documentation@example.org",
  cibersortx_token = "documentation-only-token",
  cibersortx_dry_run = TRUE,
  return_object = FALSE
)

data.frame(
  method = cibersortx_bundle$method,
  backend_launched = FALSE,
  reference_file = basename(cibersortx_bundle$files$single_cell_reference),
  mixture_file = basename(cibersortx_bundle$files$mixture),
  commands_redacted = cibersortx_bundle$artifacts$commands_redacted
)
cibersortx_bundle$command
```

## Extended: run BayesPrism locally

For documentation runtime, the reference is deterministically reduced to 800
high-abundance genes and at most 80 cells per real cluster. Every retained value
still comes from the public count matrix, and all 20 mixtures remain biological
sample pseudobulks. `update_gibbs = FALSE` reports BayesPrism's first-stage
fractions and keeps the extended build tractable.

```{r bayesprism, eval=bayesprism_ready}
gene_order <- order(Matrix::rowSums(counts), decreasing = TRUE)
deconvolution_genes <- rownames(counts)[head(gene_order, 800)]
cells_by_state <- split(colnames(pbmc), pbmc$cell_state)
reference_cells <- unlist(lapply(cells_by_state, function(cells) {
  head(sort(cells), 80)
}), use.names = FALSE)
reference <- subset(
  pbmc,
  cells = reference_cells,
  features = deconvolution_genes
)
bulk_for_deconvolution <- pseudobulk_counts[
  deconvolution_genes, , drop = FALSE
]

bayesprism_attempt <- tryCatch(
  sn_deconvolve_bulk(
    x = reference,
    bulk = bulk_for_deconvolution,
    method = "bayesprism",
    cell_type_by = "cell_state",
    cell_state_by = "cell_state",
    layer = "counts",
    n_cores = 2,
    update_gibbs = FALSE,
    return_object = FALSE
  ),
  error = identity
)

if (inherits(bayesprism_attempt, "error")) {
  knitr::kable(data.frame(
    backend = "BayesPrism", status = "failed",
    detail = conditionMessage(bayesprism_attempt), check.names = FALSE
  ))
} else {
  bayesprism_result <- bayesprism_attempt
  print(head(bayesprism_result$table, 20))
  print(ggplot2::ggplot(
    bayesprism_result$table,
    ggplot2::aes(x = .data$sample, y = .data$fraction, fill = .data$cell_type)
  ) +
    ggplot2::geom_col() +
    ggplot2::coord_flip() +
    ggplot2::labs(x = NULL, y = "Estimated fraction", fill = "Cluster") +
    ggplot2::theme_bw())
}
```

CIBERSORTx production runs remain credentialed and containerized. The article
does not synthesize a fraction table when that backend is unavailable.
