---
title: "Bulk transcriptomics from QC to clinical association"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bulk transcriptomics from QC to clinical association}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(Shennong)
```

Shennong's standalone bulk workflow keeps samples as the inferential unit. The
same feature-by-sample matrix and aligned sample metadata can be supplied as a
matrix plus data frame, a list, or a `SummarizedExperiment`.

```{r fixture}
set.seed(83)
sample_data <- data.frame(
  condition = factor(rep(c("normal", "tumor"), each = 4),
                     levels = c("normal", "tumor")),
  batch = factor(rep(1:2, 4)),
  time = seq(10, 45, by = 5),
  event = rep(c(1, 0), 4),
  row.names = paste0("sample_", 1:8)
)
counts <- matrix(
  rpois(80 * 8, 30), nrow = 80,
  dimnames = list(paste0("gene_", 1:80), rownames(sample_data))
)
counts[1:8, sample_data$condition == "tumor"] <-
  counts[1:8, sample_data$condition == "tumor"] + 35
```

## Audit samples before modeling

`sn_assess_bulk_qc()` retains sample metrics, expression quantiles, PCA,
correlation, and robust outlier flags in one validated result.

```{r qc}
qc <- sn_assess_bulk_qc(counts, sample_data)
qc$tables$samples[, c("sample", "library_size", "outlier")]
sn_plot_bulk_pca(qc, sample_data, color_by = "condition")
```

## Validate the design and contrast

The contrast contract is always `c(variable, numerator, denominator)`. With
integer counts, `method = "auto"` chooses edgeR; normalized continuous input
chooses limma; a mixed-effects formula selects dream. Explicit DESeq2, edgeR,
limma, and dream choices remain available.

```{r de, eval=FALSE}
de <- sn_find_de(
  counts,
  metadata = sample_data,
  design = ~ batch + condition,
  contrast = c("condition", "tumor", "normal"),
  method = "auto"
)

# Discover and retrieve the standardized evidence table.
names(de$tables)
head(de$tables$differential_expression)
sn_plot_bulk_de(de)
```

`sn_find_de()` detects Seurat input as single-cell and matrix/list/
`SummarizedExperiment` input as bulk. Set `modality = "single_cell"` or
`modality = "bulk"` only when dispatch must be explicit. The older
`sn_find_bulk_de()` entry point remains a compatibility wrapper around the same
bulk implementation.

## Pathways, networks, and clinical models

Pathway scoring reports gene-set coverage before returning sample scores.
WGCNA returns modules, eigengenes, power diagnostics, and module-trait
associations. Cox models and general clinical associations can use either genes
from the expression matrix or numeric columns from sample metadata.

```{r downstream, eval=FALSE}
pathways <- sn_score_bulk_pathways(
  counts,
  signatures = list(inflammation = paste0("gene_", 1:10)),
  metadata = sample_data
)
pathways$tables$coverage
pathways$tables$scores

network <- sn_run_wgcna(counts, sample_data, traits = "condition")
network$tables$modules
network$tables$trait_associations

survival <- sn_run_survival(
  counts, time = "time", event = "event",
  features = c("gene_1", "gene_2"), metadata = sample_data
)
survival$tables$survival
sn_plot_survival(survival)
```

`sn_run_bulk()` dispatches the same workflows through `workflow = "qc"`,
`"de"`, `"pathway"`, `"network"`, or `"survival"`. The explicit functions are
preferable in scripts because their required inputs are easier to review.
