---
title: "Bulk deconvolution from a PBMC3k reference"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bulk deconvolution from a PBMC3k reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Bulk deconvolution starts with one clear contract: a single-cell reference,
labels on the reference cells, and a bulk matrix with overlapping genes.
Shennong wraps that contract for CIBERSORTx and BayesPrism while keeping
credentialed or containerized steps explicit.

This article uses PBMC3k clusters as a teaching reference.

## Build a small reference and mock bulk matrix

```{r setup-deconvolution}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- sn_load_data("pbmc3k")

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

pbmc$cell_type <- paste0("cluster_", pbmc$seurat_clusters)

counts <- SeuratObject::LayerData(pbmc, assay = "RNA", layer = "counts")

bulk <- cbind(
  bulk_a = Matrix::rowSums(counts[, pbmc$cell_type %in% c("cluster_0", "cluster_1"), drop = FALSE]),
  bulk_b = Matrix::rowSums(counts[, pbmc$cell_type %in% c("cluster_2", "cluster_3"), drop = FALSE])
)
```

The mock bulk samples are sums of selected single-cell clusters. Real bulk data
should be normalized and gene-aligned according to the backend requirements.

## Prepare a CIBERSORTx run without launching the container

Use `cibersortx_dry_run = TRUE` to validate exported files and the command
bundle locally. This is the right mode for examples, tests, and documentation.

```{r cibersortx-dry-run}
bundle <- sn_deconvolve_bulk(
  x = pbmc,
  bulk = bulk,
  method = "cibersortx",
  cell_type_by = "cell_type",
  outdir = file.path(tempdir(), "pbmc3k-cibersortx"),
  prefix = "pbmc3k_demo",
  cibersortx_email = "demo@example.org",
  cibersortx_token = "fake-token",
  cibersortx_dry_run = TRUE,
  return_object = FALSE
)

names(bundle)
bundle$command
bundle$files
```

For an actual CIBERSORTx run, store credentials locally once and omit the
placeholder values.

```{r cibersortx-credentials, eval = FALSE}
sn_set_cibersortx_credentials(
  email = Sys.getenv("CIBERSORTX_EMAIL"),
  token = Sys.getenv("CIBERSORTX_TOKEN")
)

result <- sn_deconvolve_bulk(
  x = pbmc,
  bulk = bulk,
  method = "cibersortx",
  cell_type_by = "cell_type",
  outdir = "results/deconvolution/cibersortx",
  prefix = "pbmc3k"
)
```

## Import or store completed fractions

When CIBERSORTx has already produced a fraction table, import it through
`cibersortx_result` or store the table explicitly. Keeping results on the
reference object makes downstream retrieval predictable.

```{r store-result}
fractions <- data.frame(
  sample = c("bulk_a", "bulk_b"),
  cluster_0 = c(0.45, 0.10),
  cluster_1 = c(0.35, 0.05),
  cluster_2 = c(0.10, 0.50),
  cluster_3 = c(0.10, 0.35),
  check.names = FALSE
)

pbmc <- sn_store_deconvolution(
  object = pbmc,
  result = fractions,
  store_name = "pbmc3k_mock_bulk",
  method = "cibersortx",
  bulk_samples = fractions$sample,
  reference_label = "cell_type",
  artifacts = bundle$files,
  return_object = TRUE
)

sn_get_deconvolution_result(
  pbmc,
  deconvolution_name = "pbmc3k_mock_bulk"
)
```

## BayesPrism is a local optional backend

BayesPrism runs inside R and is useful when you want a Bayesian local workflow.
It is optional because it is much heavier than the dry-run CIBERSORTx export.

```{r bayesprism, eval = FALSE}
bayesprism_result <- sn_deconvolve_bulk(
  x = pbmc,
  bulk = bulk,
  method = "bayesprism",
  cell_type_by = "cell_type",
  key = NULL,
  n_cores = 2,
  store_name = "pbmc3k_bayesprism"
)
```

The rule of thumb is: use dry-run mode to verify file contracts, then run the
backend deliberately in the environment where its dependencies and credentials
are configured.
