---
title: "Bulk transcriptomics from real pseudobulk to survival"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bulk transcriptomics from real pseudobulk to survival}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
run_vignette <- identical(tolower(Sys.getenv("SHENNONG_RUN_VIGNETTES")), "true")
real_profile <- tolower(Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core"))
root_candidates <- unique(c(
  Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = ""),
  file.path(getwd(), "data-local", "pkgdown-real"),
  file.path(getwd(), "..", "data-local", "pkgdown-real")
))
root_candidates <- root_candidates[nzchar(root_candidates)]
existing_roots <- root_candidates[dir.exists(root_candidates)]
real_root <- if (length(existing_roots)) existing_roots[[1]] else root_candidates[[1]]
pbmc_path <- file.path(real_root, "single-cell", "kotliarov_pbmc.qs2")
tcga_path <- file.path(real_root, "bulk", "tcga_skcm.qs2")
core_ready <- run_vignette && file.exists(pbmc_path) && file.exists(tcga_path) &&
  requireNamespace("edgeR", quietly = TRUE) &&
  requireNamespace("survival", quietly = TRUE)
wgcna_ready <- core_ready && identical(real_profile, "all") &&
  requireNamespace("WGCNA", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 5
)
```

Two public datasets cover different contracts without changing their value
semantics:

- Kotliarov RNA UMI counts are summed within each of 20 `real_sample` values
  for count-based QC and high-versus-low response differential expression.
- TCGA-SKCM supplies RSEM TPM and upper-quartile log2-normalized RSEM TPM plus
  real follow-up. Neither TCGA assay is described or modeled as raw counts.

```{r runtime-status}
knitr::kable(data.frame(
  workflow = c(
    "pseudobulk QC and count DE", "pathway and clinical association",
    "survival", "WGCNA"
  ),
  mode = c("core", "core", "core", "extended"),
  status = c(
    if (core_ready) "executed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture or edgeR/survival dependency missing",
    if (core_ready) "executed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture or core dependency missing",
    if (core_ready) "executed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture or survival dependency missing",
    if (wgcna_ready) "executed below" else if (!identical(real_profile, "all")) "disabled: requires SHENNONG_REAL_PROFILE=all" else if (!requireNamespace("WGCNA", quietly = TRUE)) "dependency missing: WGCNA" else "fixture missing"
  ),
  check.names = FALSE
))
```

## Construct biological-sample pseudobulk counts

```{r prepare-bulk, eval=core_ready}
library(Shennong)

pbmc <- qs2::qs_read(pbmc_path)
counts <- SeuratObject::LayerData(pbmc, assay = "RNA", layer = "counts")
sample_cells <- split(colnames(pbmc), pbmc$real_sample)
pseudobulk_counts <- vapply(
  sample_cells,
  function(cells) Matrix::rowSums(counts[, cells, drop = FALSE]),
  numeric(nrow(counts))
)
rownames(pseudobulk_counts) <- rownames(counts)

pbmc_metadata <- pbmc[[]]
sample_data <- do.call(rbind, lapply(
  split(seq_len(nrow(pbmc_metadata)), pbmc_metadata$real_sample),
  function(index) data.frame(
    real_sample = pbmc_metadata$real_sample[index[[1]]],
    real_response = unique(pbmc_metadata$real_response[index]),
    observed_batches = paste(sort(unique(pbmc_metadata$real_batch[index])), collapse = "+"),
    cells = length(index),
    row.names = pbmc_metadata$real_sample[index[[1]]]
  )
))
sample_data <- sample_data[colnames(pseudobulk_counts), , drop = FALSE]
sample_data$response <- factor(
  sub("^d0 ", "", sample_data$real_response),
  levels = c("low", "high")
)

keep <- Matrix::rowSums(pseudobulk_counts >= 10) >= 3
pseudobulk_counts <- pseudobulk_counts[keep, , drop = FALSE]

tcga <- qs2::qs_read(tcga_path)
tcga_expression <- as.matrix(tcga$log2_uq_rsem_tpm)
complete_tcga_features <- rowSums(!is.finite(tcga_expression)) == 0
excluded_tcga_features <- sum(!complete_tcga_features)
tcga_expression <- tcga_expression[complete_tcga_features, , drop = FALSE]
tcga_metadata <- tcga$sample_data[colnames(tcga_expression), , drop = FALSE]

data.frame(
  dataset = c("Kotliarov pseudobulk counts", "TCGA-SKCM log2 UQ RSEM TPM"),
  features = c(nrow(pseudobulk_counts), nrow(tcga_expression)),
  samples = c(ncol(pseudobulk_counts), ncol(tcga_expression)),
  inferential_unit = "biological sample",
  excluded_nonfinite_features = c(0, excluded_tcga_features)
)
```

## Audit real pseudobulk samples through the unified dispatcher

The dedicated QC API and `sn_run_bulk()` dispatcher both run on the same 20
biological-sample count matrix. Their canonical sample tables are checked for
identity before the dedicated plotting functions consume the result.

```{r bulk-qc, eval=core_ready}
qc <- sn_assess_bulk_qc(pseudobulk_counts, sample_data)
qc_dispatch <- sn_run_bulk(
  pseudobulk_counts,
  workflow = "qc",
  metadata = sample_data
)
stopifnot(identical(qc$tables$samples, qc_dispatch$tables$samples))
data.frame(
  entry_point = c("sn_assess_bulk_qc", "sn_run_bulk(workflow = 'qc')"),
  samples = c(nrow(qc$tables$samples), nrow(qc_dispatch$tables$samples)),
  canonical_tables_identical = TRUE
)
qc$tables$samples[, c(
  "sample", "library_size", "detected_features", "mean_correlation", "outlier"
)]
sn_plot_bulk_qc(qc, metric = "library_size")
sn_plot_bulk_pca(qc, sample_data, color_by = "response")
sn_plot_sample_correlation(qc)
```

## Count differential expression with real samples

The contrast contract is `c(variable, numerator, denominator)`. Since this
matrix contains integer UMI pseudobulk counts, `method = "auto"` selects edgeR.
One donor spans both acquisition batches, so batch is reported in the design
table but is not coerced into an invalid single-valued sample covariate.

```{r bulk-de, eval=core_ready}
de <- sn_find_bulk_de(
  pseudobulk_counts,
  metadata = sample_data,
  design = ~ response,
  contrast = c("response", "high", "low"),
  method = "auto"
)

head(de$tables$differential_expression[
  order(de$tables$differential_expression$adjusted_p_value),
], 15)
sn_plot_bulk_de(de, log2_fold_change = 0.5)
```

The generic result-aware figure API consumes the same fitted edgeR result.
Unlike cluster-marker tables restricted to positive markers, this real
20-sample contrast retains both directions and is therefore suitable for a
volcano plot and a signed pathway ranking.

```{r bulk-result-figures, eval=core_ready}
sn_plot_de(
  de,
  type = "volcano",
  adjusted_p_value = 0.05,
  log2_fold_change = 0.5
)

signed_de <- de$tables$differential_expression
signed_de <- signed_de[
  is.finite(signed_de$log2_fold_change) &
    !is.na(signed_de$gene) & nzchar(signed_de$gene),
  ,
  drop = FALSE
]

gsea_result <- sn_enrich(
  x = signed_de,
  gene_clusters = gene ~ log2_fold_change,
  analysis = "gsea",
  species = "human",
  database = "GOBP",
  pvalue_cutoff = 1
)
gsea_table <- as.data.frame(gsea_result)
stopifnot(nrow(gsea_table) > 0L)

sn_plot_enrichment(gsea_result, type = "dot", n = 15)
sn_plot_gsea(gsea_result)
```

## Pathways and clinical associations in TCGA-SKCM

These signatures are intersected with the 3,000 retained real genes before
scoring. Clinical models test expression against the curated sample class and
sex; each TCGA patient remains one observation.

```{r pathways-clinical, eval=core_ready}
candidate_signatures <- list(
  T_cell = c("CD3D", "CD3E", "CD8A", "LCK", "TRAC"),
  interferon = c("STAT1", "IRF1", "CXCL9", "CXCL10", "GBP1"),
  melanoma_identity = c("MITF", "MLANA", "PMEL", "TYR", "SOX10")
)
signatures <- lapply(candidate_signatures, intersect, y = rownames(tcga_expression))
signatures <- signatures[lengths(signatures) >= 2]

pathways <- sn_score_bulk_pathways(
  tcga_expression,
  signatures = signatures,
  method = "mean",
  metadata = tcga_metadata
)
pathways$tables$coverage
head(pathways$tables$scores)

clinical_features <- intersect(
  c("MITF", "CD8A", "PDCD1", "B2M"),
  rownames(tcga_expression)
)
clinical <- sn_run_clinical_association(
  tcga_expression,
  features = clinical_features,
  clinical_vars = c("sample_type", "sex"),
  metadata = tcga_metadata
)
clinical$tables$associations
```

## Overall survival and diagnostics

Follow-up uses `time > 0` days and `event` encoded as 0/1 in the local curated
artifact. The same fitted result drives the forest plot, Kaplan-Meier curve,
risk table, and proportional-hazards diagnostics.

```{r survival, eval=core_ready}
survival_features <- intersect(
  c("MITF", "CD8A", "PDCD1", "B2M"),
  rownames(tcga_expression)
)
survival_result <- sn_run_survival(
  tcga_expression,
  time = "time",
  event = "event",
  features = survival_features,
  covariates = "sex",
  metadata = tcga_metadata,
  group_method = "median"
)

survival_result$tables$survival
survival_result$tables$log_rank
survival_result$tables$proportional_hazards
survival_result$tables$performance
sn_plot_survival(survival_result)
sn_plot_survival(survival_result, view = "km", feature = survival_features[[1]])
sn_plot_survival(survival_result, view = "risk_table", feature = survival_features[[1]])
sn_plot_survival(survival_result, view = "ph_test", feature = survival_features[[1]])
```

## Extended: real WGCNA network

WGCNA runs only in the `all` profile with the package installed. The 500 most
variable TCGA genes are selected deterministically from the real normalized
matrix. A failure is reported as such rather than replaced with a precomputed
module assignment.

```{r wgcna, eval=wgcna_ready}
gene_variance <- apply(tcga_expression, 1, stats::var)
wgcna_genes <- names(head(sort(gene_variance, decreasing = TRUE), 500))
wgcna_attempt <- tryCatch(
  sn_run_wgcna(
    tcga_expression[wgcna_genes, , drop = FALSE],
    metadata = tcga_metadata,
    traits = c("event", "sample_type"),
    power = 6,
    min_module_size = 20,
    backend_control = list(blockwise = list(maxBlockSize = 600))
  ),
  error = identity
)

if (inherits(wgcna_attempt, "error")) {
  knitr::kable(data.frame(
    backend = "WGCNA", status = "failed",
    detail = conditionMessage(wgcna_attempt), check.names = FALSE
  ))
} else {
  network <- wgcna_attempt
  print(head(network$tables$modules, 20))
  print(network$tables$trait_associations)
  print(sn_plot_wgcna(network, type = "modules"))
  print(sn_plot_wgcna(network, type = "traits"))
}
```
