---
title: "Clustering and integration with PBMC3k"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering and integration with PBMC3k}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Seurat teaches users a sequence of steps: normalize, find variable genes, run
PCA, build neighbors, cluster, and embed. Shennong keeps that logic, but exposes
it through one workflow function so the choices are visible in one call.

This article uses PBMC3k. For speed, the website shows code without running the
analysis unless `SHENNONG_RUN_VIGNETTES=true`.

## One function for the standard clustering path

The minimal clustering call is intentionally short. The arguments that usually
matter for a report are visible: normalization method, number of features, PC
range, resolution, clustering algorithm, and species.

```{r cluster-pbmc3k}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- sn_load_data("pbmc3k")

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 2000,
  npcs = 30,
  dims = 1:20,
  resolution = 0.6,
  cluster_algorithm = "leiden",
  species = "human",
  verbose = FALSE
)

pbmc
```

When the input already contains `sn_run_cluster()` stage metadata, the function
reuses matching stages by default. This makes resolution tuning cheap:
`sn_run_cluster(pbmc, resolution = 1.0)` reuses normalization, HVGs, PCA, the
neighbor graph, and UMAP, then recomputes only the cluster labels. Use
`rerun_from = "integration"` after changing an integration backend, or
`reuse = FALSE` when a completely fresh run is needed.

The result is a regular Seurat object. You can use Shennong plotting helpers or
drop down to Seurat whenever needed.

```{r plot-clusters}
sn_plot_dim(
  object = pbmc,
  reduction = "umap",
  group_by = "seurat_clusters",
  label = TRUE,
  title = "PBMC3k clusters"
)
```

## Block unhelpful feature families

Single-cell clustering can be dominated by mitochondrial, ribosomal,
immunoglobulin, or heat-shock genes. `block_genes` makes that choice explicit
instead of hiding it inside a custom variable-feature script.

```{r block-genes}
pbmc_blocked <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 2000,
  dims = 1:20,
  resolution = 0.6,
  species = "human",
  block_genes = c("mito", "ribo", "immunoglobulins"),
  verbose = FALSE
)
```

Use this when a technical or lineage-specific gene family is overwhelming the
structure you want to resolve.

## Rare-aware feature selection

Small populations can disappear when variable genes are selected only by global
variance. Shennong can append rare-aware features before PCA.

```{r rare-aware}
pbmc_rare <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  rare_feature_method = c("gini", "local_markers"),
  rare_feature_group_by = "seurat_clusters",
  rare_feature_n = 50,
  rare_feature_control = list(group_max_fraction = 0.05),
  dims = 1:20,
  resolution = 0.8,
  species = "human",
  verbose = FALSE
)
```

The key idea is simple: keep the global HVGs, then add a small number of genes
that help preserve rare groups. The rare-cell scoring helper can be used
independently to inspect which cells look unusual.

```{r rare-cells}
rare_tbl <- sn_detect_rare_cells(
  pbmc,
  method = "gini",
  nfeatures = 200
)

head(rare_tbl[order(rare_tbl$rare_score, decreasing = TRUE), ])
```

## Force known marker genes into the PCA feature set

Sometimes you already know the marker genes that separate a rare population,
but those genes are too sparse to rank among the global HVGs. Use
`hvg_features` to merge those genes into the final ScaleData/PCA feature set
after Shennong validates that they exist in the object.

```{r manual-hvg}
pbmc_manual <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  hvg_features = c("IL7R", "CCR7", "NKG7", "MS4A1"),
  dims = 1:20,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

pbmc_manual@misc$hvg_selection$user_features
```

This is different from `rare_feature_method`: `rare_feature_method` asks
Shennong to discover rare-aware features, while `hvg_features` records a
user-provided feature prior.

## Choose resolution empirically

Instead of treating `resolution = 0.8` as a magic number,
`sn_sweep_cluster_resolution()` reruns clustering over a small grid and reports
diagnostics such as cluster count, silhouette, and graph connectivity.

```{r resolution-sweep}
sweep <- sn_sweep_cluster_resolution(
  pbmc,
  resolutions = seq(0.2, 1.0, by = 0.2),
  reduction = "pca",
  dims = 1:20,
  metrics = c("n_clusters", "mean_silhouette", "graph_connectivity"),
  max_cells = 1500
)

sweep$summary
sweep$recommended_resolution
```

The recommendation is not a replacement for biology; it is a compact way to
find resolutions that are numerically stable before looking at marker genes.

## Integration is the same API with a batch column

PBMC3k is one dataset, so the example below creates a teaching-only pseudo
batch column. In a real experiment, `batch` would be a donor, library, chemistry,
or dataset column.

```{r harmony-integration, eval = FALSE}
pbmc$library <- rep(c("library_a", "library_b"), length.out = ncol(pbmc))

pbmc_integrated <- sn_run_cluster(
  object = pbmc,
  batch = "library",
  integration_method = "harmony",
  normalization_method = "seurat",
  hvg_group_by = "library",
  nfeatures = 2000,
  dims = 1:20,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

sn_plot_dim(
  object = pbmc_integrated,
  reduction = "umap",
  group_by = "library",
  title = "Pseudo-batch mixing"
)
```

This keeps the mental model small: single-sample clustering and batch
integration use the same entry point; adding `batch` turns on integration, and
`integration_method` names the backend. Harmony is the default because it is
fast and broadly useful.

The same integration switch can be used with SCTransform normalization:

```{r sctransform-integration, eval = FALSE}
pbmc_sct_integrated <- sn_run_cluster(
  object = pbmc,
  batch = "library",
  integration_method = "harmony",
  normalization_method = "sctransform",
  hvg_features = c("IL7R", "CCR7", "NKG7"),
  dims = 1:20,
  resolution = 0.6,
  verbose = FALSE
)
```

For imbalanced datasets where a rare or unevenly distributed state may be
washed out by conventional correction, Coralysis can be selected explicitly.
Coralysis runs on the log-normalized feature set selected by Shennong and
stores the integrated embedding as the `coralysis` reduction.

```{r coralysis-integration, eval = FALSE}
pbmc_coral <- sn_run_cluster(
  object = pbmc,
  batch = "library",
  integration_method = "coralysis",
  normalization_method = "seurat",
  hvg_group_by = "library",
  nfeatures = 2000,
  dims = 1:20,
  resolution = 0.6,
  integration_control = list(
    store_sce = TRUE,
    icp_args = list(L = 25, threads = 2),
    pca_args = list(p = 20, return.model = TRUE)
  ),
  verbose = FALSE
)
```

Seurat layer integration methods are also available when users want to compare
against familiar anchor-based workflows:

```{r seurat-layer-integration, eval = FALSE}
pbmc_cca <- sn_run_cluster(
  object = pbmc,
  batch = "library",
  integration_method = "seurat_cca",
  normalization_method = "seurat",
  nfeatures = 2000,
  dims = 1:20,
  verbose = FALSE
)

pbmc_rpca <- sn_run_cluster(
  object = pbmc,
  batch = "library",
  integration_method = "seurat_rpca",
  normalization_method = "seurat",
  nfeatures = 2000,
  dims = 1:20,
  verbose = FALSE
)
```

Python integration methods can be run through a pixi-managed scverse project
kept under the user-level Shennong runtime directory, `~/.shennong/`. When
`integration_method = "scvi"`, Shennong exports the selected count matrix and
metadata, renders `~/.shennong/pixi/scvi/pixi.toml` from the package-bundled
`inst/pixi/scvi/pixi.toml` template when needed, runs
the Python backend, imports `latent.csv` as the `scvi` reduction, and then
continues with Seurat neighbors, clusters, and UMAP.

The R package does not vendor the Python environments. It ships only the
backend runner script and pixi config templates; the pixi workspaces and
environment prefixes are created under `~/.shennong/pixi/`, matching the
package's `~/.shennong/data/` convention. Use `sn_pixi_paths("scvi")` to inspect the
layout before a run. If pixi is not already available, Shennong can install the
standalone pixi binary through `sn_ensure_pixi()` and then use the Shennong
`PIXI_HOME` under `~/.shennong/pixi/home` for runtime configuration. By default
`accelerator = "auto"` uses a CUDA pixi environment when `nvidia-smi` reports a
CUDA-capable GPU and otherwise uses the CPU environment. China mirror handling
can be enabled with `mirror = "auto"` or a specific mirror such as `"tuna"`.
Other bundled Python environment configs can be inspected with
`sn_list_pixi_environments()` and materialized with
`sn_prepare_pixi_environment()`. Method aliases share environments when the
underlying software stack is the same: scANVI uses the `scvi` environment, and
scPoli uses the `scarches` environment. Spatial tools are represented by
specific environment names such as `cell2location`, `tangram`, `squidpy`,
`spatialdata`, and `stlearn`.

For method wrappers that define an object-level contract, use the `object =`
form. Runner scripts are stored with each pixi family under
`inst/pixi/<family>/scripts/`. For example, inferCNVpy can be run directly on a
Seurat object and will import cell-level CNV metadata and CNV reductions back
into that object:

```{r infercnvpy-object, eval = FALSE}
tumor <- sn_run_infercnvpy(
  object = tumor,
  assay = "RNA",
  layer = "data",
  species = "human",
  reference_by = "cell_type",
  reference_cat = c("T cell", "Myeloid", "Endothelial")
)
```

Other Python tools follow the same Seurat-first pattern. Tools that need
additional biological inputs make those inputs explicit:

```{r python-object-tools, eval = FALSE}
object <- sn_run_cellphonedb(object, group_by = "cell_type")

spatial <- sn_run_tangram(
  object = spatial,
  reference_object = sc_reference,
  cell_type_by = "cell_type",
  spatial_cols = c("x", "y")
)

spatial <- sn_run_cell2location(
  object = spatial,
  reference_signatures = signature_matrix,
  spatial_cols = c("x", "y"),
  method_control = list(max_epochs = 30000)
)

spatial <- sn_run_squidpy(object = spatial, spatial_cols = c("x", "y"))
```

```{r scvi-integration, eval = FALSE}
sn_list_pixi_environments()
sn_pixi_paths("scvi")
sn_ensure_pixi()

pbmc_scvi <- sn_run_cluster(
  object = pbmc,
  batch = "library",
  integration_method = "scvi",
  normalization_method = "seurat",
  hvg_group_by = "library",
  nfeatures = 3000,
  dims = 1:30,
  integration_control = list(
    accelerator = "auto",
    mirror = "auto",
    n_latent = 30,
    max_epochs = 100
  ),
  verbose = FALSE
)

pbmc_scanvi <- sn_run_cluster(
  object = pbmc,
  batch = "library",
  integration_method = "scanvi",
  normalization_method = "seurat",
  hvg_group_by = "library",
  nfeatures = 3000,
  dims = 1:30,
  integration_control = list(
    label_by = "reference_label",
    unlabeled_category = "Unknown",
    n_latent = 30,
    max_epochs = 100
  ),
  verbose = FALSE
)
```
