---
title: "Clustering and integration with real PBMC CITE-seq data"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering and integration with real PBMC CITE-seq data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_file <- file.path(
  Sys.getenv("SHENNONG_REAL_DATA_DIR"),
  "single-cell",
  "kotliarov_pbmc.qs2"
)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_requested <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")
run_vignette <- run_requested &&
  requireNamespace("qs2", quietly = TRUE) &&
  file.exists(real_data_file)
run_extended <- run_vignette && identical(real_profile, "all")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Seurat teaches users a sequence of steps: normalize, find variable genes, run
PCA, build neighbors, cluster, and embed. Shennong keeps that logic, but exposes
it through one workflow function so the choices are visible in one call.

This article uses the public Kotliarov PBMC CITE-seq cohort: 2,000 real cells
from 20 biological samples, two acquisition batches, and aligned RNA and ADT
assays. A real-data pkgdown build executes the core chunks. Ordinary package
checks safely leave them unevaluated when the local fixture is absent.

## One function for the standard clustering path

The minimal clustering call is intentionally short. The arguments that usually
matter for a report are visible: normalization method, number of features, PC
range, resolution, clustering algorithm, and species.

```{r cluster-kotliarov-pbmc}
library(Shennong)
library(Seurat)
library(dplyr)

pixi_environment_ready <- function(environment) {
  if (!nzchar(Sys.which("pixi"))) return(FALSE)
  paths <- sn_pixi_paths(environment)
  file.exists(paths$manifest_path) && dir.exists(paths$workspace_env_dir)
}

pbmc <- qs2::qs_read(real_data_file)

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 2000,
  npcs = 30,
  dims = 1:20,
  resolution = 0.6,
  cluster_algorithm = "leiden",
  species = "human",
  verbose = FALSE
)

data.frame(
  cells = ncol(pbmc),
  clusters = length(unique(pbmc$seurat_clusters)),
  biological_samples = length(unique(pbmc$real_sample)),
  batches = length(unique(pbmc$real_batch))
)
```

Variable-feature and PCA diagnostics are produced from the same real RNA
matrix used for clustering.

```{r hvg-and-elbow-diagnostics, eval = run_vignette}
sn_plot_hvg(pbmc, assay = "RNA", label_n = 10)
sn_plot_elbow(pbmc, reduction = "pca", ndims = 20)
```

When the input already contains `sn_run_cluster()` stage metadata, the function
reuses matching stages by default. This makes resolution tuning cheap:
`sn_run_cluster(pbmc, resolution = 1.0)` reuses normalization, HVGs, PCA, the
neighbor graph, and UMAP, then recomputes only the cluster labels. Use
`rerun_from = "integration"` after changing an integration backend, or
`reuse = FALSE` when a completely fresh run is needed.

UMAP geometry can be tuned independently from clustering with `umap_control`.
This is useful when clusters are biologically separate but overlap visually in
the two-dimensional embedding.

```{r tune-umap, eval = run_vignette}
pbmc_umap <- sn_run_cluster(
  pbmc,
  umap_control = list(
    n.neighbors = 20,
    min.dist = 0.1,
    spread = 1,
    reduction.name = "umap_tuned"
  ),
  rerun_from = "umap"
)

sn_plot_dim(
  pbmc_umap,
  reduction = "umap_tuned",
  group_by = "seurat_clusters",
  label = TRUE,
  title = "UMAP retuned on real PBMC cells"
)
```

The result is a regular Seurat object. You can use Shennong plotting helpers or
drop down to Seurat whenever needed.

```{r plot-clusters}
sn_plot_dim(
  object = pbmc,
  reduction = "umap",
  group_by = "seurat_clusters",
  label = TRUE,
  title = "Kotliarov PBMC clusters"
)
```

## Block unhelpful feature families

Single-cell clustering can be dominated by cell-cycle, mitochondrial,
ribosomal, immunoglobulin, pseudogene, or heat-shock genes. `block_genes`
makes that choice explicit instead of hiding it inside a custom
variable-feature script. The vector can mix bundled signature names, full
signature paths, and custom gene symbols, and is applied to internally selected
HVGs in both log-normalization and SCTransform workflows.

```{r block-genes}
pbmc_blocked <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 2000,
  dims = 1:20,
  resolution = 0.6,
  species = "human",
  block_genes = c("cellCycle.G2M", "cellCycle.G1S", "mito", "ribo", "MYC"),
  verbose = FALSE
)

data.frame(
  selected_hvgs = length(pbmc_blocked@misc$hvg_selection$selected_features),
  blocked_hvgs = length(pbmc_blocked@misc$hvg_selection$blocked_features)
)
head(pbmc_blocked@misc$hvg_selection$blocked_features)
```

Use this when a technical or lineage-specific gene family is overwhelming the
structure you want to resolve.

## Rare-aware feature selection

Small populations can disappear when variable genes are selected only by global
variance. Shennong can append rare-aware features before PCA.

```{r rare-aware}
pbmc_rare <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  rare_feature_method = c("gini", "local_markers"),
  rare_feature_group_by = "seurat_clusters",
  rare_feature_n = 50,
  rare_feature_control = list(group_max_fraction = 0.05),
  dims = 1:20,
  resolution = 0.8,
  species = "human",
  verbose = FALSE
)

head(pbmc_rare@misc$hvg_selection$selected_features)
table(pbmc_rare$seurat_clusters)
```

The key idea is simple: keep the global HVGs, then add a small number of genes
that help preserve rare groups. The rare-cell scoring helper can be used
independently to inspect which cells look unusual.

```{r rare-cells}
rare_tbl <- sn_detect_rare_cells(
  pbmc,
  method = "gini",
  nfeatures = 200
)

head(rare_tbl[order(rare_tbl$rare_score, decreasing = TRUE), ])
```

## Force known marker genes into the PCA feature set

Sometimes you already know the marker genes that separate a rare population,
but those genes are too sparse to rank among the global HVGs. Use
`hvg_features` to merge those genes into the final ScaleData/PCA feature set
after Shennong validates that they exist in the object.

```{r manual-hvg}
pbmc_manual <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  hvg_features = c("IL7R", "CCR7", "NKG7", "MS4A1"),
  dims = 1:20,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

pbmc_manual@misc$hvg_selection$user_features
```

This is different from `rare_feature_method`: `rare_feature_method` asks
Shennong to discover rare-aware features, while `hvg_features` records a
user-provided feature prior.

## CITE-seq RNA and protein workflows

For CITE-seq objects with paired RNA and ADT assays, set
`modality = "cite_seq"`. The default `multimodal_method = "wnn"` runs the
standard RNA PCA path, normalizes the ADT assay with CLR, computes ADT PCA, then
uses Seurat weighted nearest neighbors to cluster on the `wsnn` graph and create
a `wnn.umap` embedding. Here both modalities come from the same real cells.
`sn_run_multimodal(object, method = "wnn")` is the explicit multimodal entry
point and accepts the remaining `sn_run_cluster()` arguments through `...`.

```{r cite-seq-wnn, eval = run_vignette}
pbmc_cite <- qs2::qs_read(real_data_file)
pbmc_cite <- sn_run_multimodal(
  object = pbmc_cite,
  method = "wnn",
  assay = "RNA",
  adt_assay = "ADT",
  normalization_method = "seurat",
  nfeatures = 2000,
  dims = 1:20,
  adt_dims = 1:18,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

table(pbmc_cite$seurat_clusters, pbmc_cite$real_batch)
sn_plot_dim(pbmc_cite, reduction = "wnn.umap", group_by = "seurat_clusters")
```

For batch-aware CITE-seq integration, choose a multimodal backend explicitly.
`multimodal_method = "totalvi"` runs scvi-tools totalVI on RNA counts and ADT
counts. `multimodal_method = "coralysis"` runs native Coralysis on the ADT
assay as a log-normalized protein matrix.
`multimodal_method = "mmochi"` runs MMoCHi ADT landmark registration across
the supplied `batch` column, or in single-sample mode when `batch = NULL`.
It stores the corrected protein matrix as an assay layer when the local Seurat
object supports arbitrary layers and otherwise under
`object@misc$mmochi$corrected_protein`, computes a protein-derived `mmochi`
reduction, and clusters from that reduction. The full MMoCHi hierarchy
classifier remains an advanced Python workflow because it requires an explicit
user-defined hierarchy.

These external or optional backends are evaluated only by the `all` real-data
profile and only when their runtime dependency is already available. The
article never installs a Python environment or changes GPU settings while it is
being built.

```{r cite-seq-totalvi, eval = run_extended && pixi_environment_ready("scvi")}
pbmc_totalvi <- qs2::qs_read(real_data_file)
pbmc_totalvi <- sn_run_cluster(
  object = pbmc_totalvi,
  modality = "cite_seq",
  multimodal_method = "totalvi",
  batch = "real_batch",
  assay = "RNA",
  adt_assay = "ADT",
  normalization_method = "seurat",
  nfeatures = 3000,
  dims = 1:30,
  integration_control = list(max_epochs = 100),
  verbose = FALSE
)

table(pbmc_totalvi$seurat_clusters, pbmc_totalvi$real_batch)
sn_plot_dim(pbmc_totalvi, reduction = "umap", group_by = "seurat_clusters")
```

```{r cite-seq-coralysis-protein, eval = run_extended && requireNamespace("Coralysis", quietly = TRUE)}
pbmc_adt_coral <- qs2::qs_read(real_data_file)
pbmc_adt_coral <- sn_run_cluster(
  object = pbmc_adt_coral,
  modality = "cite_seq",
  multimodal_method = "coralysis",
  batch = "real_batch",
  adt_assay = "ADT",
  adt_dims = 1:18,
  verbose = FALSE
)

table(pbmc_adt_coral$seurat_clusters, pbmc_adt_coral$real_batch)
```

```{r cite-seq-mmochi-protein, eval = run_extended && pixi_environment_ready("mmochi")}
pbmc_adt_mmochi <- qs2::qs_read(real_data_file)
pbmc_adt_mmochi <- sn_run_cluster(
  object = pbmc_adt_mmochi,
  modality = "cite_seq",
  multimodal_method = "mmochi",
  batch = "real_batch",
  adt_assay = "ADT",
  adt_dims = 1:18,
  integration_control = list(
    protein_layer = "data",
    store_corrected_layer = TRUE
  ),
  verbose = FALSE
)

table(pbmc_adt_mmochi$seurat_clusters, pbmc_adt_mmochi$real_batch)
```

## Choose resolution empirically

Instead of treating `resolution = 0.8` as a magic number,
`sn_sweep_cluster_resolution()` reruns clustering over a small grid and reports
diagnostics such as cluster count, silhouette, and graph connectivity.

```{r resolution-sweep}
sweep <- sn_sweep_cluster_resolution(
  pbmc,
  resolutions = seq(0.2, 1.0, by = 0.2),
  reduction = "pca",
  dims = 1:20,
  metrics = c("n_clusters", "mean_silhouette", "graph_connectivity"),
  max_cells = 1500
)

sweep$summary
sweep$recommended_resolution
sn_plot_resolution_sweep(sweep)

resolution_cols <- c("res_0_2", "res_0_6", "res_1_0")
pbmc_tree <- pbmc
for (index in seq_along(resolution_cols)) {
  pbmc_tree <- sn_run_cluster(
    pbmc_tree,
    resolution = c(0.2, 0.6, 1.0)[[index]],
    cluster_name = resolution_cols[[index]],
    reuse = TRUE,
    verbose = FALSE
  )
}

sn_plot_cluster_tree(pbmc_tree, resolution_cols = resolution_cols)
```

The recommendation is not a replacement for biology; it is a compact way to
find resolutions that are numerically stable before looking at marker genes.

## Integration is the same API with a real batch column

`real_batch` records the two acquisition batches in the source cohort. It is a
technical label, whereas `real_sample` remains the biological replicate for
sample-level inference.

```{r harmony-integration, eval = run_vignette && requireNamespace("harmony", quietly = TRUE)}

pbmc_integrated <- sn_run_cluster(
  object = pbmc,
  batch = "real_batch",
  integration_method = "harmony",
  normalization_method = "seurat",
  hvg_group_by = "real_batch",
  nfeatures = 2000,
  dims = 1:20,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

sn_plot_dim(
  object = pbmc_integrated,
  reduction = "umap",
  group_by = "real_batch",
  title = "Mixing of the two real acquisition batches"
)
```

This keeps the mental model small: single-sample clustering and batch
integration use the same entry point; adding `batch` turns on integration, and
`integration_method` names the backend. Harmony is the default because it is
fast and broadly useful.

### Compare several integration methods in one object

Pass a vector to retain an unintegrated baseline and several corrected spaces
without repeating normalization, HVG selection, or PCA. Parameters can be
keyed by method; the historical spelling `"unintergrated"` is accepted with a
deprecation warning, but new code should use `"unintegrated"`.

Scalar-by-contract parameters also accept vectors. Shennong expands their
conditional Cartesian product: for example, two `nfeatures` values and two
`resolution` values create four runs per method, but the two resolutions share
one graph and one UMAP for each method/HVG combination. Natural vectors such as
`dims`, `hvg_features`, `vars_to_regress`, and `block_genes` remain one value.
UMAP is the only projection run by default; set the existing `run_tsne = TRUE`
explicitly when t-SNE is also required.

```{r multi-integration, eval = run_extended && requireNamespace("harmony", quietly = TRUE) && requireNamespace("Coralysis", quietly = TRUE)}
pbmc_comparison <- sn_run_cluster(
  object = pbmc,
  batch = "real_batch",
  integration_method = c("unintegrated", "harmony", "coralysis"),
  integration_control = list(
    harmony = list(theta = 3),
    coralysis = list(
      icp_args = list(L = 25, threads = 1),
      pca_args = list(p = 20),
      store_sce = FALSE
    )
  ),
  normalization_method = "seurat",
  hvg_group_by = "real_batch",
  nfeatures = c(2000, 3000),
  dims = 1:20,
  resolution = c(0.4, 0.8),
  checkpoint_dir = "checkpoints/pbmc-integration-grid",
  verbose = FALSE
)

# Discover the stored result names.
pbmc_comparison@misc$integration_comparison$results

# Inspect the run/embedding/preprocessing map, then retrieve exact components.
comparison_grid <- pbmc_comparison@misc$integration_comparison$grid
first_run <- pbmc_comparison@misc$integration_comparison$results[[1]]
clusters <- pbmc_comparison[[first_run$cluster_column, drop = TRUE]]
umap <- Seurat::Embeddings(pbmc_comparison, first_run$umap_reduction)

# Compare compute cost per run and per unique integration embedding.
pbmc_comparison@misc$integration_comparison$performance
```

For a method-only comparison with scalar parameters, the historical names such
as `results$harmony`, `harmony_clusters`, and `umap.harmony` remain unchanged.
Grid runs use deterministic run IDs. The unintegrated entry points to a stored
PCA branch. Integrated methods
retain their native corrected low-dimensional spaces (for example `harmony` or
`coralysis`) rather than relabeling those coordinates as PCA. The manifest maps
each run to its preprocessing/embedding IDs, reduction, graph names, cluster
column, UMAP, and optional t-SNE so
downstream integration metrics can iterate over one returned object.

When `checkpoint_dir` is supplied, every completed grid entry is saved through
a `.partial` file followed by an atomic rename. A repeated call with the same
object identity, batch labels, package version, grid, and analysis arguments
resumes automatically (`resume = TRUE`) and skips completed entries. Only the
latest complete checkpoint is retained; interrupted `.partial` files are
ignored. Set `checkpoint_compress = TRUE` when disk space matters more than
checkpoint write speed.

The `performance` table contains wall-clock seconds, peak R heap memory, and
integration-specific values. On Linux, pixi/Python backends also capture the
maximum backend process-tree RSS with GNU `time`; GPU device memory is not
included. Resolution-only entries have `reused_embedding = TRUE` and retain the
source embedding's integration cost while recording only the new clustering
work as their workflow time.

### Integration-control templates

Use the executable templates to discover every Shennong-consumed backend
field, including runtime, accelerator, model, training, graph, and
backend-specific arguments:

```{r integration-control-templates}
names(sn_integration_control_template())
sn_integration_control_template("harmony")
sn_integration_control_template("coralysis")
sn_integration_control_template(c("scvi", "scanvi", "scpoli"))
sn_integration_control_template(c("bbknn", "totalvi", "mmochi"))
```

For Seurat CCA/RPCA, fields beyond the displayed template are passed directly
to `Seurat::IntegrateLayers()`. For nested model calls, place native backend
arguments in `model_args`, `train_args`, `scanvi_model_args`,
`scanvi_train_args`, `totalvi_model_args`, `totalvi_train_args`, `bbknn_args`,
or the corresponding Coralysis/MMoCHi argument list.

Use `sn_compare_integrations()` when the object also contains an independent
biological annotation column. The scib-metrics adapter evaluates the native
PCA or integrated latent spaces on exactly the same stratified cells and
features within each preprocessing group; it does not score UMAP or t-SNE
coordinates. A matching unintegrated PCA is required in every preprocessing
group. Each unique embedding is scored once, so resolution-only runs inherit
the same embedding metrics. Supervised methods trained with the
same column as `label_by` are flagged, and graph-only methods such as BBKNN are
listed as non-comparable rather than silently treated as embeddings.
The stored clustering timing and memory columns are joined into the scIB
summary, metric, and ranking tables, enabling quality-versus-compute
comparisons from the benchmark result alone.

```{r scib-integration-benchmark, eval = FALSE}
pbmc_comparison <- sn_compare_integrations(
  object = pbmc_comparison,
  batch_by = "real_batch",
  label_by = "cell_type",
  accelerator = "auto",
  n_jobs = 4
)

# Discover and retrieve the stored benchmark.
sn_list_results(pbmc_comparison, type = "integration_benchmark")
integration_benchmark <- sn_get_result(
  pbmc_comparison,
  type = "integration_benchmark",
  name = "integration_benchmark"
)
integration_benchmark$tables$ranking
integration_benchmark$tables$summary[, c(
  "run_id", "embedding_id", "preprocess_id", "method", "Total",
  "integration_elapsed_seconds", "integration_peak_memory_mb",
  "rank_within_preprocess", "rank_overall"
)]
integration_benchmark$tables$metrics
integration_benchmark$diagnostics$backend_manifest$jax_devices
```

`accelerator = "auto"` selects the managed GPU environment when Shennong
detects a usable NVIDIA CUDA runtime, otherwise it uses CPU. The recorded JAX
device list is the runtime evidence for the device scib-metrics actually used.

The same integration switch can be used with SCTransform normalization. This
heavier optional backend runs in the `all` profile when both Harmony and
`glmGamPoi` are installed:

```{r sctransform-integration, eval = run_extended && requireNamespace("harmony", quietly = TRUE) && requireNamespace("glmGamPoi", quietly = TRUE)}
pbmc_sct_integrated <- sn_run_cluster(
  object = pbmc,
  batch = "real_batch",
  integration_method = "harmony",
  normalization_method = "sctransform",
  hvg_features = c("IL7R", "CCR7", "NKG7"),
  dims = 1:20,
  resolution = 0.6,
  verbose = FALSE
)

table(pbmc_sct_integrated$seurat_clusters, pbmc_sct_integrated$real_batch)
```

For imbalanced datasets where a rare or unevenly distributed state may be
washed out by conventional correction, Coralysis can be selected explicitly.
Coralysis runs on the log-normalized feature set selected by Shennong and
stores the integrated embedding as the `coralysis` reduction. The trained
Coralysis SingleCellExperiment is stored by default under
`object@misc$coralysis`, so the returned object can be used later as a native
Coralysis label-transfer reference. Set
`integration_control = list(store_sce = FALSE)` only for clustering-only runs.
For BPCells-backed assays, Shennong keeps the complete normalized layer on disk
and materializes only the selected integration features as a sparse
`dgCMatrix`, which is the input class required by Coralysis. Coralysis defaults
to one worker because forked workers can each copy this in-memory matrix; set
`integration_control = list(icp_args = list(threads = ...))` only after sizing
the aggregate worker memory against the host.

```{r coralysis-integration, eval = run_extended && requireNamespace("Coralysis", quietly = TRUE)}
pbmc_coral <- sn_run_cluster(
  object = pbmc,
  batch = "real_batch",
  integration_method = "coralysis",
  normalization_method = "seurat",
  hvg_group_by = "real_batch",
  nfeatures = 2000,
  dims = 1:20,
  resolution = 0.6,
  integration_control = list(
    icp_args = list(L = 25, threads = 2),
    pca_args = list(p = 20)
  ),
  umap_control = list(
    n.neighbors = 20,
    min.dist = 0.1
  ),
  verbose = FALSE
)

table(pbmc_coral$seurat_clusters, pbmc_coral$real_batch)
```

Seurat layer integration methods are also available when users want to compare
against familiar anchor-based workflows:

```{r seurat-layer-integration, eval = run_vignette}
pbmc_cca <- sn_run_cluster(
  object = pbmc,
  batch = "real_batch",
  integration_method = "seurat_cca",
  normalization_method = "seurat",
  nfeatures = 2000,
  dims = 1:20,
  verbose = FALSE
)

pbmc_rpca <- sn_run_cluster(
  object = pbmc,
  batch = "real_batch",
  integration_method = "seurat_rpca",
  normalization_method = "seurat",
  nfeatures = 2000,
  dims = 1:20,
  verbose = FALSE
)

data.frame(
  method = c("seurat_cca", "seurat_rpca"),
  clusters = c(
    length(unique(pbmc_cca$seurat_clusters)),
    length(unique(pbmc_rpca$seurat_clusters))
  )
)

sn_plot_dim(
  pbmc_rpca,
  reduction = "umap",
  group_by = "real_batch",
  title = "RPCA integration across real acquisition batches"
)
```

Python integration methods can be run through a pixi-managed scverse project
kept under the user-level Shennong runtime directory, `~/.shennong/`. When
`integration_method = "scvi"`, Shennong exports the count matrix from the
requested `assay` and `layer` together with metadata, renders
`~/.shennong/pixi/scvi/pixi.toml` from the package-bundled
`inst/pixi/scvi/pixi.toml` template when needed, runs
the Python backend, imports `latent.csv` as the `scvi` reduction, and then
continues with Seurat neighbors, clusters, and UMAP.

The R package does not vendor the Python environments. It ships only the
backend runner script and pixi config templates; the pixi workspaces and
environment prefixes are created under `~/.shennong/pixi/`, matching the
package's `~/.shennong/data/` convention. Use `sn_pixi_paths("scvi")` to inspect the
layout before a run. If pixi is not already available, Shennong can install the
standalone pixi binary through `sn_ensure_pixi()` and then use the Shennong
`PIXI_HOME` under `~/.shennong/pixi/home` for runtime configuration. By default
`accelerator = "auto"` uses a CUDA pixi environment when `nvidia-smi` reports a
CUDA-capable GPU and otherwise uses the CPU environment. China mirror handling
can be enabled with `mirror = "auto"` or a specific mirror such as `"tuna"`.
Other bundled Python environment configs can be inspected with
`sn_list_pixi_environments()` and materialized with
`sn_prepare_pixi_environment()`. Method aliases share environments when the
underlying software stack is the same: scANVI uses the `scvi` environment, and
scPoli uses the `scarches` environment. Spatial tools are represented by
specific environment names such as `cell2location`, `tangram`, `squidpy`,
`spatialdata`, and `stlearn`.

The BBKNN, scArches, stLearn, trajectory, and cell2location templates also pin
the Python AutoZyme package from the same `zerostwo/autozyme` revision used by
Shennong's R integration. Only runners with a matching executed hotspot request
a patch. Patchable upstreams are pinned to validated versions and checked again
at runtime; cell2location acceleration is restricted to CPU full-batch training.
Each output manifest distinguishes requested, active, and inactive patches.
scVI/scANVI, scPoli training, Tangram, MMoCHi, infercnvpy, scIB, and
the current Squidpy path remain ordinary upstream execution because the fork
does not provide a patch for the code they call. Set `AUTOZYME_DISABLED=true`
to compare against the unpatched Python path.

The layer contract also applies to scANVI and scPoli. BBKNN consumes the PCA
computed from that same selected layer, imports its batch-balanced connectivity
graph, and uses the graph directly for Seurat clustering and UMAP. Therefore a
comparison on decontaminated counts can keep one explicit input across all
methods instead of silently falling back to `counts`:

The Python runners keep complete expression matrices sparse. scPoli converts
only one bounded neural-network minibatch to a dense tensor during latent
encoding; the complete cells-by-genes matrix is never materialized as dense.
Dense PCA, latent, and UMAP outputs remain low-dimensional by construction.

```{r layer-consistent-python-integration, eval=FALSE}
methods <- c("scvi", "scanvi", "scpoli", "bbknn")
integrated <- lapply(methods, function(method) {
  control <- if (method == "scanvi") list(label_by = "cell_type") else list()
  sn_run_cluster(
    object = pbmc,
    batch = "real_batch",
    assay = "RNA",
    layer = "decontaminated_counts",
    integration_method = method,
    integration_control = control,
    normalization_method = "seurat",
    dims = 1:30,
    verbose = FALSE
  )
})
```

Cancer CNV and spatial Python wrappers require disease-matched references or
spatial coordinates, so they are intentionally exercised in the corresponding
real melanoma and Visium articles rather than against this PBMC object.

```{r scvi-integration, eval = run_extended && pixi_environment_ready("scvi")}
sn_list_pixi_environments()
sn_pixi_paths("scvi")

pbmc_scvi <- qs2::qs_read(real_data_file)
pbmc_scvi <- sn_run_scvi(
  object = pbmc_scvi,
  batch = "real_batch",
  layer = "counts",
  normalization_method = "seurat",
  hvg_group_by = "real_batch",
  nfeatures = 3000,
  dims = 1:30,
  integration_control = list(
    accelerator = "auto",
    mirror = "auto",
    n_latent = 30,
    max_epochs = 100
  ),
  verbose = FALSE
)

table(pbmc_scvi$seurat_clusters, pbmc_scvi$real_batch)
sn_plot_dim(pbmc_scvi, reduction = "umap", group_by = "real_batch")

scanvi_input <- pbmc
held_out_samples <- utils::tail(sort(unique(as.character(pbmc$real_sample))), 5)
scanvi_input$scanvi_label <- as.character(scanvi_input$seurat_clusters)
scanvi_input$scanvi_label[scanvi_input$real_sample %in% held_out_samples] <- "Unknown"

pbmc_scanvi <- sn_run_scanvi(
  object = scanvi_input,
  batch = "real_batch",
  layer = "counts",
  normalization_method = "seurat",
  nfeatures = 3000,
  dims = 1:30,
  integration_control = list(
    label_by = "scanvi_label",
    unlabeled_category = "Unknown",
    accelerator = "auto",
    mirror = "auto",
    n_latent = 30,
    max_epochs = 100
  ),
  verbose = FALSE
)

table(pbmc_scanvi$scanvi_prediction, pbmc_scanvi$real_response)
```

The scPoli object wrapper trains the same real scPoli model used by
`integration_method = "scpoli"`; it uses the real batch column as its condition
and can use RNA clusters as prototype labels. The generic scArches wrapper
remains a custom-runner surface for reference-mapping workflows. Neither is run
unless the shared scArches pixi environment has already been materialized.

```{r scarches-and-scpoli, eval = run_extended && pixi_environment_ready("scarches")}
pbmc_scarches <- sn_run_scarches(
  object = pbmc,
  assay = "RNA",
  layer = "counts",
  batch_by = "real_batch",
  label_by = "seurat_clusters",
  method_control = list(max_epochs = 100)
)

pbmc_scpoli <- sn_run_scpoli(
  object = pbmc,
  assay = "RNA",
  layer = "counts",
  batch_by = "real_batch",
  label_by = "seurat_clusters",
  method_control = list(max_epochs = 100)
)

data.frame(
  method = c("scarches", "scpoli"),
  reductions = c(
    paste(names(pbmc_scarches@reductions), collapse = ", "),
    paste(names(pbmc_scpoli@reductions), collapse = ", ")
  )
)
```
