---
title: "PBMC workflow"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PBMC workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") ||
  identical(Sys.getenv("IN_PKGDOWN"), "true")

required_pkgs <- c(
  "Seurat",
  "hdf5r",
  "harmony",
  "clusterProfiler",
  "org.Hs.eg.db",
  "lisi"
)

run_vignette <- run_vignette && all(vapply(
  required_pkgs,
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  eval = run_vignette
)
```

This article runs a complete PBMC analysis using the built-in `pbmc1k` and
`pbmc3k` example datasets. The workflow covers QC-aware preprocessing,
single-sample clustering, pre/post integration comparison, sample composition
summaries, integration scoring with LISI, marker discovery, dot-plot
visualization, and functional enrichment from cluster markers.

To keep `R CMD check` deterministic, the analysis is only evaluated when
`SHENNONG_RUN_VIGNETTES=true` or during pkgdown builds.

```{r setup}
library(Shennong)
library(dplyr)
library(ggplot2)
library(knitr)
library(Seurat)
```

```{r prepare-pbmc-demo, include = FALSE}
prepare_sample <- function(dataset) {
  object <- sn_load_data(dataset = dataset)
  object$sample <- dataset
  object$study <- "pbmc_demo"
  object <- suppressWarnings(
    sn_filter_cells(
      object,
      features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
      plot = FALSE
    )
  )
  suppressWarnings(sn_filter_genes(object, min_cells = 3, plot = FALSE))
}

pbmc1k <- prepare_sample("pbmc1k")
pbmc3k <- prepare_sample("pbmc3k")

pbmc1k_clustered <- suppressWarnings(
  sn_run_cluster(
    object = pbmc1k,
    normalization_method = "seurat",
    species = "human",
    nfeatures = 800,
    block_genes = NULL,
    npcs = 30,
    dims = 1:20,
    resolution = 0.5,
    verbose = FALSE
  )
)

merged_pbmc <- merge(
  x = pbmc1k,
  y = pbmc3k,
  add.cell.ids = c("pbmc1k", "pbmc3k")
)
merged_pbmc$sample <- ifelse(
  grepl("^pbmc1k_", colnames(merged_pbmc)),
  "pbmc1k",
  "pbmc3k"
)
merged_pbmc$study <- "pbmc_demo"

pbmc_unintegrated <- suppressWarnings(
  sn_run_cluster(
    object = merged_pbmc,
    normalization_method = "seurat",
    hvg_group_by = "sample",
    species = "human",
    nfeatures = 800,
    block_genes = NULL,
    npcs = 30,
    dims = 1:20,
    resolution = 0.5,
    verbose = FALSE
  )
)

pbmc_integrated <- suppressWarnings(
  sn_run_cluster(
    object = merged_pbmc,
    batch = "sample",
    normalization_method = "seurat",
    hvg_group_by = "sample",
    species = "human",
    nfeatures = 800,
    block_genes = NULL,
    npcs = 30,
    dims = 1:20,
    resolution = 0.5,
    verbose = FALSE
  )
)

cell_summary <- tibble::tibble(
  dataset = c("pbmc1k", "pbmc3k", "merged_unintegrated", "merged_integrated"),
  cells = c(ncol(pbmc1k), ncol(pbmc3k), ncol(pbmc_unintegrated), ncol(pbmc_integrated)),
  genes = c(nrow(pbmc1k), nrow(pbmc3k), nrow(pbmc_unintegrated), nrow(pbmc_integrated)),
  clusters = c(
    length(levels(Seurat::Idents(pbmc1k_clustered))),
    NA_integer_,
    length(levels(Seurat::Idents(pbmc_unintegrated))),
    length(levels(Seurat::Idents(pbmc_integrated)))
  )
)

composition_tbl <- sn_calculate_composition(
  pbmc_integrated,
  group_by = "sample",
  variable = "seurat_clusters",
  min_cells = 0
)

lisi_before <- sn_calculate_lisi(
  pbmc_unintegrated,
  reduction = "pca",
  label = "sample"
)

lisi_after <- sn_calculate_lisi(
  pbmc_integrated,
  reduction = "harmony",
  label = "sample"
)

lisi_summary <- dplyr::bind_rows(
  tibble::tibble(state = "before_integration", score = lisi_before$sample),
  tibble::tibble(state = "after_integration", score = lisi_after$sample)
) |>
  dplyr::group_by(state) |>
  dplyr::summarise(
    min = min(score),
    q1 = stats::quantile(score, 0.25),
    median = stats::median(score),
    mean = mean(score),
    q3 = stats::quantile(score, 0.75),
    max = max(score),
    .groups = "drop"
  )

pbmc_integrated <- suppressWarnings(
  sn_find_de(
    pbmc_integrated,
    analysis = "markers",
    group_by = "seurat_clusters",
    assay = "RNA",
    layer = "data",
    logfc_threshold = 0.25,
    min_pct = 0.2,
    store_name = "cluster_markers",
    return_object = TRUE,
    verbose = FALSE
  )
)

markers_full <- pbmc_integrated@misc$de_results$cluster_markers$table

marker_tbl <- markers_full |>
  dplyr::group_by(cluster) |>
  dplyr::slice_max(order_by = avg_log2FC, n = 3, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::select(cluster, gene, avg_log2FC, p_val_adj)

find_enrichment_result <- function(markers) {
  for (current_cluster in unique(as.character(markers$cluster))) {
    marker_subset <- markers |>
      dplyr::filter(.data$cluster == current_cluster) |>
      dplyr::arrange(dplyr::desc(.data$avg_log2FC))

    enrichment <- suppressWarnings(
      sn_enrich(
        marker_subset,
        analysis = "gsea",
        species = "human",
        database = "GOBP",
        gene_col = "gene",
        score_col = "avg_log2FC",
        pvalue_cutoff = 0.1
      )
    )
    enrichment_tbl <- as.data.frame(enrichment)

    if (nrow(enrichment_tbl) > 0) {
      return(list(cluster = current_cluster, table = enrichment_tbl))
    }
  }

  stop("No non-empty enrichment result was produced for the PBMC marker sets.")
}

enrichment_result <- find_enrichment_result(markers_full)
focus_cluster <- enrichment_result$cluster
enrichment_tbl <- enrichment_result$table |>
  dplyr::select(ID, Description, NES, p.adjust) |>
  utils::head(10)
```

## Data loaded for the workflow

```{r}
knitr::kable(cell_summary, digits = 0)
```

The merged analysis is intentionally run twice: once without batch correction
and once with Harmony integration. This makes the before/after comparison
explicit instead of only showing the corrected embedding.

## Single-sample clustering

```{r}
sn_plot_dim(
  object = pbmc1k_clustered,
  reduction = "umap",
  group_by = "seurat_clusters",
  label = TRUE,
  show_legend = FALSE,
  title = "pbmc1k clustered with sn_run_cluster()"
)
```

## Before and after Harmony integration

```{r}
sn_plot_dim(
  object = pbmc_unintegrated,
  reduction = "umap",
  group_by = "sample",
  title = "Before integration: merged object colored by sample"
)
```

```{r}
sn_plot_dim(
  object = pbmc_integrated,
  reduction = "umap",
  group_by = "sample",
  title = "After Harmony integration: merged object colored by sample"
)
```

```{r}
sn_plot_dim(
  object = pbmc_integrated,
  reduction = "umap",
  group_by = "seurat_clusters",
  label = TRUE,
  show_legend = FALSE,
  title = "After Harmony integration: merged object colored by cluster"
)
```

## Cluster composition by sample

```{r}
knitr::kable(dplyr::slice_head(composition_tbl, n = 12), digits = 2)
```

```{r}
sn_plot_barplot(
  composition_tbl,
  x = sample,
  y = proportion,
  fill = seurat_clusters
) +
  labs(
    title = "Cluster composition across pbmc1k and pbmc3k",
    x = NULL,
    y = "Cells (%)",
    fill = "Cluster"
  ) +
  theme_minimal(base_size = 12)
```

## Integration quality summary with LISI

```{r}
knitr::kable(lisi_summary, digits = 3)
```

Higher LISI values indicate better local sample mixing. In this two-sample
example the theoretical maximum is 2, so the post-Harmony distribution should
shift upward relative to the unintegrated baseline.

## Marker genes from the integrated object

```{r}
knitr::kable(marker_tbl, digits = 3)
```

```{r}
sn_plot_dot(
  x = pbmc_integrated,
  features = "top_markers",
  de_name = "cluster_markers",
  n = 3
)
```

## Functional enrichment of marker genes

The table below shows the first non-empty GO Biological Process GSEA result
found among the cluster marker sets. In this run, enrichment was computed from
cluster
`r if (run_vignette) focus_cluster else "<evaluated during pkgdown build>"`.

```{r}
knitr::kable(enrichment_tbl, digits = 4)
```

```{r}
sessioninfo::session_info()
```
