---
title: "CNV, Malignancy, and Metabolism"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CNV, Malignancy, and Metabolism}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
run_vignette <- identical(tolower(Sys.getenv("SHENNONG_RUN_VIGNETTES")), "true")
real_profile <- tolower(Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core"))
root_candidates <- unique(c(
  Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = ""),
  file.path(getwd(), "data-local", "pkgdown-real"),
  file.path(getwd(), "..", "data-local", "pkgdown-real")
))
root_candidates <- root_candidates[nzchar(root_candidates)]
existing_roots <- root_candidates[dir.exists(root_candidates)]
real_root <- if (length(existing_roots)) existing_roots[[1]] else root_candidates[[1]]
melanoma_path <- file.path(real_root, "cancer", "melanoma_gse72056.qs2")
core_ready <- run_vignette && file.exists(melanoma_path)
infercnvpy_paths <- if (requireNamespace("Shennong", quietly = TRUE)) {
  tryCatch(Shennong::sn_pixi_paths("infercnvpy"), error = function(error) NULL)
} else {
  NULL
}
infercnvpy_environment_exists <- !is.null(infercnvpy_paths) &&
  file.exists(infercnvpy_paths$manifest_path) &&
  dir.exists(infercnvpy_paths$workspace_env_dir) &&
  length(list.files(infercnvpy_paths$workspace_env_dir, all.files = FALSE)) > 0
infercnvpy_ready <- core_ready && identical(real_profile, "all") &&
  infercnvpy_environment_exists

knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 6
)
```

The revised GSE72056 matrix contains author-normalized expression and an
author-provided malignancy class derived from inferred CNV. It does not contain
raw UMI counts, so this article never passes those values to CopyKAT as if they
were counts. Curated gene-set metabolism can run directly on the normalized
layer; inferCNVpy runs only when an already-installed audited environment is
present.

```{r runtime-status}
knitr::kable(data.frame(
  workflow = c(
    "author malignancy evidence", "curated metabolism",
    "inferCNVpy", "CopyKAT", "scMetabolism / scFEA / Compass"
  ),
  mode = c("core source evidence", "core", "extended", "extended", "extended"),
  status = c(
    if (core_ready) "displayed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture missing",
    if (core_ready) "executed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture missing",
    if (infercnvpy_ready) "executed below" else if (!identical(real_profile, "all")) "disabled: requires SHENNONG_REAL_PROFILE=all" else "not run: pre-existing inferCNVpy pixi environment absent",
    "not run: source has normalized expression rather than raw counts",
    "not run: optional backend/result not supplied"
  ),
  check.names = FALSE
))
```

## Inspect real malignancy evidence

```{r prepare-melanoma, eval=core_ready}
library(Shennong)

melanoma <- qs2::qs_read(melanoma_path)
malignancy_evidence <- as.data.frame(table(
  malignant_call = melanoma$malignant_call,
  cell_type = melanoma$cell_type
))
malignancy_evidence <- malignancy_evidence[
  malignancy_evidence$Freq > 0, , drop = FALSE
]
malignancy_evidence
data.frame(
  tumors = length(unique(melanoma$tumor)),
  cells = ncol(melanoma),
  normalized_genes = nrow(melanoma),
  expression_layer = paste(SeuratObject::Layers(melanoma[["RNA"]]), collapse = ", ")
)
```

## Curated metabolic activity with biological-sample summaries

The core path uses a sparse-aware mean score to avoid an optional scoring
dependency. Tumor is the sample unit and the study's malignancy class is a
descriptive stratum. No condition p-value is computed because primary versus
metastatic is not a sample-level field in GSE72056.

```{r metabolism, eval=core_ready}
metabolic_signatures <- sn_metabolic_signatures("human")
metabolic_signatures <- metabolic_signatures[c(
  "glycolysis", "tca_cycle", "oxidative_phosphorylation"
)]

metabolism <- sn_run_metabolism(
  melanoma,
  method = "geneset",
  signatures = metabolic_signatures,
  scoring_method = "mean",
  assay = "RNA",
  layer = "data",
  sample_by = "tumor",
  group_by = "malignant_call",
  species = "human",
  min_genes = 3L,
  return_object = FALSE
)

metabolism$tables$coverage
head(metabolism$tables$sample_scores, 30)
metabolism$tables$differential
sn_plot_metabolism(metabolism, type = "activity")
sn_plot_metabolism(metabolism, type = "heatmap")
sn_plot_metabolism(metabolism, type = "sample")
```

## Extended: reference-aware inferCNVpy

This chunk can execute only when `SHENNONG_REAL_PROFILE=all` and the
inferCNVpy pixi environment already exists. `install_pixi = FALSE` prevents the
article from downloading software. Non-malignant immune and stromal cells are
the reference; the author malignancy class is retained for comparison but is
not imported as a backend result. The runner calls `sn_run_infercnvpy()`
explicitly once, then hands its real score, chromosome, cluster, and embedding
outputs to `sn_run_cnv()` for unified summaries and plots.

```{r infercnvpy, eval=infercnvpy_ready}
set.seed(717)
cells_by_call <- split(colnames(melanoma), melanoma$malignant_call)
cnv_cells <- unlist(lapply(cells_by_call, function(cells) {
  if (length(cells) <= 400) cells else sample(cells, 400)
}), use.names = FALSE)
cnv_object <- subset(melanoma, cells = cnv_cells)

article_infercnvpy_runner <- function(object, reference_cells, genome,
                                      store_name, assay, layer, sample_by) {
  object$.article_cnv_reference <- ifelse(
    colnames(object) %in% reference_cells, "reference", "query"
  )
  updated <- sn_run_infercnvpy(
    object,
    assay = assay,
    layer = layer,
    species = genome,
    reference_by = ".article_cnv_reference",
    reference_cat = "reference",
    runtime_dir = infercnvpy_paths$runtime_dir,
    result_name = store_name,
    metadata_prefix = "infercnvpy_",
    run_umap = TRUE,
    n_jobs = 2,
    return_object = TRUE,
    install_pixi = FALSE,
    quiet = TRUE
  )
  updated$.article_cnv_reference <- NULL

  metadata <- updated[[]]
  score_column <- intersect(
    c("infercnvpy_cnv_score", "infercnvpy_score"), colnames(metadata)
  )
  if (length(score_column) == 0L) {
    stop("inferCNVpy completed without an imported CNV score column.")
  }
  cluster_column <- intersect(
    c("infercnvpy_cnv_leiden", "infercnvpy_leiden"), colnames(metadata)
  )
  manifest <- updated@misc$infercnvpy[[store_name]]

  chromosome_path <- file.path(manifest$output_dir, "cnv_chromosome.csv")
  chromosomes <- tibble::tibble()
  if (file.exists(chromosome_path)) {
    chromosome_wide <- utils::read.csv(
      chromosome_path, row.names = 1, check.names = FALSE
    )
    chromosome_wide$cell <- rownames(chromosome_wide)
    chromosomes <- dplyr::bind_rows(lapply(
      setdiff(colnames(chromosome_wide), "cell"),
      function(chromosome) tibble::tibble(
        cell = chromosome_wide$cell,
        chromosome = chromosome,
        cnv = suppressWarnings(as.numeric(chromosome_wide[[chromosome]]))
      )
    ))
  }
  embedding_name <- "infercnvpy_cnv_umap"

  list(
    object = updated,
    scores = stats::setNames(
      as.numeric(metadata[[score_column[[1]]]]), rownames(metadata)
    ),
    chromosomes = chromosomes,
    subclone = if (length(cluster_column)) {
      stats::setNames(metadata[[cluster_column[[1]]]], rownames(metadata))
    } else {
      NULL
    },
    prediction = NULL,
    embedding = if (embedding_name %in% names(updated@reductions)) {
      SeuratObject::Embeddings(updated[[embedding_name]])
    } else {
      NULL
    },
    artifacts = manifest
  )
}

cnv_attempt <- tryCatch(
  sn_run_cnv(
    cnv_object,
    method = "infercnvpy",
    reference_by = "malignant_call",
    reference_cat = "non_malignant",
    sample_by = "tumor",
    assay = "RNA",
    layer = "data",
    association_features = 30L,
    backend_control = list(
      runner = article_infercnvpy_runner
    ),
    return_object = FALSE
  ),
  error = identity
)

if (inherits(cnv_attempt, "error")) {
  knitr::kable(data.frame(
    backend = "inferCNVpy", status = "failed",
    detail = conditionMessage(cnv_attempt), check.names = FALSE
  ))
} else {
  cnv <- cnv_attempt
  print(head(cnv$tables$primary, 20))
  print(cnv$tables$sample_summary)
  print(head(cnv$tables$expression_association, 20))
  print(sn_plot_cnv(cnv, type = "heatmap", n = 20))
  print(sn_plot_cnv(cnv, type = "score"))
  print(sn_plot_cnv(cnv, type = "association"))
}
```

scMetabolism, scFEA, and Compass remain optional external computations. The
pkgdown article does not reinterpret the real mean-score table as output from
those backends.
