---
title: "Comparable cell-cell communication and consensus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparable cell-cell communication and consensus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
run_vignette <- identical(tolower(Sys.getenv("SHENNONG_RUN_VIGNETTES")), "true")
real_profile <- tolower(Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core"))
root_candidates <- unique(c(
  Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = ""),
  file.path(getwd(), "data-local", "pkgdown-real"),
  file.path(getwd(), "..", "data-local", "pkgdown-real")
))
root_candidates <- root_candidates[nzchar(root_candidates)]
existing_roots <- root_candidates[dir.exists(root_candidates)]
real_root <- if (length(existing_roots)) existing_roots[[1]] else root_candidates[[1]]
pbmc_path <- file.path(real_root, "single-cell", "kotliarov_pbmc.qs2")
core_ready <- run_vignette && file.exists(pbmc_path)
consensus_ready <- core_ready && identical(real_profile, "all") &&
  requireNamespace("liana", quietly = TRUE) &&
  requireNamespace("CellChat", quietly = TRUE)
cellphonedb_paths <- if (requireNamespace("Shennong", quietly = TRUE)) {
  tryCatch(Shennong::sn_pixi_paths("cellphonedb"), error = function(error) NULL)
} else {
  NULL
}
cellphonedb_environment_exists <- !is.null(cellphonedb_paths) &&
  file.exists(cellphonedb_paths$manifest_path) &&
  dir.exists(cellphonedb_paths$workspace_env_dir) &&
  length(list.files(cellphonedb_paths$workspace_env_dir, all.files = FALSE)) > 0
cellphonedb_ready <- core_ready && identical(real_profile, "all") &&
  cellphonedb_environment_exists

knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 6
)
```

Shennong keeps backend-native artifacts while mapping communication evidence
to a common table contract. Here, cell groups are clusters fitted from the real
Kotliarov PBMC counts, and all condition comparisons use the 20
`real_sample` values rather than treating cells as replicates.

```{r runtime-status}
knitr::kable(data.frame(
  workflow = c(
    "real PBMC design", "sample-level ligand-target plot",
    "LIANA + CellChat consensus", "CellPhoneDB", "MultiNicheNet"
  ),
  mode = c("core", "core", "extended", "extended", "extended"),
  status = c(
    if (core_ready) "executed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture missing",
    if (core_ready) "executed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture missing",
    if (consensus_ready) "executed below" else if (!identical(real_profile, "all")) "disabled: requires SHENNONG_REAL_PROFILE=all" else if (!requireNamespace("liana", quietly = TRUE) || !requireNamespace("CellChat", quietly = TRUE)) "dependency missing: liana and/or CellChat" else "fixture missing",
    if (cellphonedb_ready) "executed below" else if (!identical(real_profile, "all")) "disabled: requires SHENNONG_REAL_PROFILE=all" else "not run: pre-existing CellPhoneDB pixi environment absent",
    "not run: no ligand-target prior supplied"
  ),
  check.names = FALSE
))
```

## Prepare the real sample-aware design

This core chunk is useful even when no communication backend is installed: it
shows exactly how many cells, clusters, and biological samples would enter the
extended analysis.

```{r prepare-communication, eval=core_ready}
library(Shennong)
library(Seurat)

pbmc <- qs2::qs_read(pbmc_path)
pbmc <- sn_run_cluster(
  pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.5,
  species = "human",
  verbose = FALSE
)
pbmc$cell_state <- paste0("cluster_", pbmc$seurat_clusters)

communication_design <- as.data.frame(table(
  sample = pbmc$real_sample,
  response = pbmc$real_response,
  cluster = pbmc$cell_state
))
communication_design <- communication_design[
  communication_design$Freq > 0, , drop = FALSE
]
head(communication_design, 20)
data.frame(
  cells = ncol(pbmc),
  samples = length(unique(pbmc$real_sample)),
  clusters = length(unique(pbmc$cell_state)),
  response_groups = paste(sort(unique(pbmc$real_response)), collapse = " / ")
)
```

## Core: observed ligand-target co-expression across real samples

`sn_plot_ligand_target()` plots the unified `ligand_targets` table; it does
not require weights to be NicheNet regulatory potentials. Here the weights are
reported explicitly as Spearman correlations between sender-cluster ligand
expression and receiver-cluster target expression across the 20 real samples.
This is observed co-expression, not causal ligand-target evidence. Candidate
ligands are selected by cytokine/chemokine gene-family prefixes, while target
genes are selected from the fitted variable features by between-sample
variance. No interaction score is handwritten or simulated.

```{r ligand-target, eval=core_ready}
expression <- SeuratObject::LayerData(pbmc, assay = "RNA", layer = "data")
state_sizes <- sort(table(pbmc$cell_state), decreasing = TRUE)
sender_state <- names(state_sizes)[[1]]
receiver_state <- names(state_sizes)[[2]]
sample_ids <- sort(unique(as.character(pbmc$real_sample)))

ligand_candidates <- grep(
  "^(CCL|CXCL|IL[0-9]|TNF|TGFB|IFN)",
  rownames(expression), value = TRUE
)
sender_cells <- colnames(pbmc)[pbmc$cell_state == sender_state]
ligand_abundance <- Matrix::rowMeans(
  expression[ligand_candidates, sender_cells, drop = FALSE]
)
ligands <- names(head(sort(ligand_abundance, decreasing = TRUE), 12))

sample_state_means <- function(features, state) {
  means <- vapply(sample_ids, function(sample) {
    cells <- colnames(pbmc)[
      as.character(pbmc$real_sample) == sample & pbmc$cell_state == state
    ]
    if (length(cells) == 0L) {
      rep(NA_real_, length(features))
    } else {
      Matrix::rowMeans(expression[features, cells, drop = FALSE])
    }
  }, numeric(length(features)))
  rownames(means) <- features
  means
}

ligand_by_sample <- sample_state_means(ligands, sender_state)
target_pool <- setdiff(intersect(Seurat::VariableFeatures(pbmc), rownames(expression)), ligands)
target_by_sample <- sample_state_means(target_pool, receiver_state)
target_variance <- apply(target_by_sample, 1, stats::var, na.rm = TRUE)
targets <- names(head(sort(target_variance, decreasing = TRUE), 60))
target_by_sample <- target_by_sample[targets, , drop = FALSE]

ligand_targets <- expand.grid(
  ligand = ligands,
  target_gene = targets,
  stringsAsFactors = FALSE
)
ligand_targets$weight <- vapply(seq_len(nrow(ligand_targets)), function(index) {
  stats::cor(
    ligand_by_sample[ligand_targets$ligand[[index]], ],
    target_by_sample[ligand_targets$target_gene[[index]], ],
    method = "spearman", use = "pairwise.complete.obs"
  )
}, numeric(1))
ligand_targets <- ligand_targets[is.finite(ligand_targets$weight), , drop = FALSE]
ligand_targets <- ligand_targets[
  order(abs(ligand_targets$weight), decreasing = TRUE), , drop = FALSE
]

observed_primary <- data.frame(
  source = sender_state,
  target = receiver_state,
  ligand = ligand_targets$ligand,
  receptor = NA_character_,
  score = abs(ligand_targets$weight),
  method = "sample_spearman",
  stringsAsFactors = FALSE
)
observed_communication <- sn_store_cell_communication(
  pbmc,
  result = observed_primary,
  store_name = "kotliarov_observed_ligand_targets",
  method = "sample_spearman",
  backend = "real_sample_coexpression",
  group_by = "cell_state",
  sender = sender_state,
  receiver = receiver_state,
  ligand_targets = ligand_targets,
  sample_by = "real_sample",
  condition_by = "real_response",
  return_object = FALSE
)

head(observed_communication$tables$ligand_targets, 20)
sn_plot_ligand_target(observed_communication, n = 30)
```

## Extended: run a real cross-method consensus

This computation is intentionally gated behind `SHENNONG_REAL_PROFILE=all`.
Both installed backends and the unified consensus/concordance path must finish;
an error stops the extended render instead of being converted into a status
table or replaced with substitute interactions.

```{r consensus, eval=consensus_ready}
communication <- sn_run_cell_communication(
  pbmc,
  method = c("liana", "cellchat"),
  group_by = "cell_state",
  sample_by = "real_sample",
  condition_by = "real_response",
  contrast = c("d0 high", "d0 low"),
  species = "human",
  consensus = TRUE,
  backend_control = list(
    liana = list(resource = "consensus", method = "natmi"),
    cellchat = list(min_cells = 10)
  ),
  store_name = "kotliarov_consensus",
  return_object = FALSE
)

print(head(communication$tables$primary, 20))
print(head(communication$tables$backend_raw, 20))
print(head(communication$tables$consensus, 20))
print(communication$tables$method_concordance)
print(head(communication$tables$sample_evidence, 20))
print(head(communication$tables$condition_comparison, 20))
print(sn_plot_communication(communication, type = "bubble"))
print(sn_plot_communication(communication, type = "heatmap"))
print(sn_plot_communication(communication, type = "network"))
if (nrow(communication$tables$condition_comparison) > 0) {
  print(sn_plot_communication_comparison(communication))
}
```

## Extended: direct CellPhoneDB object workflow

The direct object adapter runs only when an audited local CellPhoneDB pixi
environment already exists. Its input is a balanced subset of the same real
PBMC counts, its output directory is temporary, and `install_pixi = FALSE`
prevents pkgdown from downloading software.

```{r cellphonedb, eval=cellphonedb_ready}
set.seed(717)
cells_by_state <- split(colnames(pbmc), pbmc$cell_state)
cellphonedb_cells <- unlist(lapply(cells_by_state, function(cells) {
  if (length(cells) <= 120L) cells else sample(cells, 120L)
}), use.names = FALSE)
cellphonedb_object <- subset(pbmc, cells = cellphonedb_cells)

cellphonedb_attempt <- tryCatch(
  sn_run_cellphonedb(
    cellphonedb_object,
    assay = "RNA",
    layer = "counts",
    group_by = "cell_state",
    output_dir = tempfile("shennong-cellphonedb-pkgdown-"),
    runtime_dir = cellphonedb_paths$runtime_dir,
    result_name = "kotliarov_cellphonedb",
    return_object = FALSE,
    method_control = list(iterations = 1000L),
    install_pixi = FALSE,
    quiet = TRUE
  ),
  error = identity
)

if (inherits(cellphonedb_attempt, "error")) {
  knitr::kable(data.frame(
    backend = "CellPhoneDB", status = "failed",
    detail = conditionMessage(cellphonedb_attempt), check.names = FALSE
  ))
} else {
  knitr::kable(data.frame(
    backend = "CellPhoneDB", status = "completed",
    cells = cellphonedb_attempt$n_cells,
    features = cellphonedb_attempt$n_features,
    output_dir = cellphonedb_attempt$output_dir,
    check.names = FALSE
  ))
}
```

CellPhoneDB needs a prepared Python runtime, while MultiNicheNet and NicheNet
need explicit prior networks. Neither is silently installed or replaced with a
handwritten interaction table during pkgdown rendering.
