---
title: "Composition and comparative analysis workflow"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composition and comparative analysis workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") ||
  identical(Sys.getenv("IN_PKGDOWN"), "true")

required_pkgs <- c("Seurat")
run_vignette <- run_vignette && all(vapply(
  required_pkgs,
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  eval = run_vignette
)
```

This workflow covers the analysis tasks that happen after clusters or cell
types already exist and the main question becomes comparative rather than
structural:

- how cell states are distributed across samples
- whether cluster proportions shift between groups
- how QC or score columns differ across categories

```{r setup}
library(Shennong)
library(dplyr)
library(knitr)
library(Seurat)

if (!exists("pbmc_small", inherits = FALSE)) {
  try(data("pbmc_small", package = "Shennong", envir = environment()), silent = TRUE)
}
if (!exists("pbmc_small", inherits = FALSE) && file.exists(file.path("data", "pbmc_small.rda"))) {
  load(file.path("data", "pbmc_small.rda"))
}
```

```{r prepare-composition-demo, include = FALSE}
pbmc_comp <- suppressWarnings(
  sn_run_cluster(
    pbmc_small,
    normalization_method = "seurat",
    nfeatures = 800,
    npcs = 20,
    dims = 1:15,
    resolution = 0.4,
    verbose = FALSE
  )
)
pbmc_comp$sample <- pbmc_small@meta.data[colnames(pbmc_comp), "sample"]
pbmc_comp <- sn_score_cell_cycle(pbmc_comp, species = "human")

composition_tbl <- sn_calculate_composition(
  pbmc_comp,
  group_by = "sample",
  variable = "seurat_clusters",
  min_cells = 0
)

phase_tbl <- sn_calculate_composition(
  pbmc_comp,
  group_by = "sample",
  variable = "Phase",
  min_cells = 0
)

score_tbl <- pbmc_comp@meta.data |>
  dplyr::select(sample, seurat_clusters, S.Score, G2M.Score, nFeature_RNA)
```

## 1. Summarize cell-state proportions

`sn_calculate_composition()` is the main helper for grouped proportion
summaries. The most common pattern is sample-by-cluster composition.

```{r}
knitr::kable(composition_tbl, digits = 2)
```

## 2. Plot proportions across groups

The returned composition table can be fed directly into `sn_plot_barplot()`.

```{r}
sn_plot_barplot(
  composition_tbl,
  x = sample,
  y = proportion,
  fill = seurat_clusters
)
```

## 3. Compare categorical structure beyond clusters

The same composition helper also works for cell-cycle phase, annotation labels,
or any other metadata column.

```{r}
knitr::kable(phase_tbl, digits = 2)
```

## 4. Compare continuous scores across groups

Once a grouped table exists, Shennong's plotting helpers can summarize how QC
or score columns differ by sample or cluster.

```{r}
sn_plot_boxplot(score_tbl, x = sample, y = nFeature_RNA)
```

```{r}
sn_plot_boxplot(score_tbl, x = seurat_clusters, y = S.Score)
```

## 5. Keep comparative analysis downstream of stable labels

Composition and grouped-score analysis become much more interpretable once the
labels are stable. In practice the recommended order is:

1. preprocessing and QC
2. clustering or integration
3. marker or reference-based annotation
4. composition and grouped comparisons

That order reduces the chance of over-interpreting composition shifts that are
actually caused by unstable clustering.
