---
title: "Composition and differential abundance"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composition and differential abundance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_file <- file.path(
  Sys.getenv("SHENNONG_REAL_DATA_DIR"),
  "single-cell",
  "kotliarov_pbmc.qs2"
)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_requested <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")
run_vignette <- run_requested &&
  requireNamespace("qs2", quietly = TRUE) &&
  file.exists(real_data_file)
run_extended <- run_vignette && identical(real_profile, "all")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Composition analysis should not treat every cell as an independent biological
replicate. Shennong separates cell-level counting from sample-level comparison:
first compute proportions, then compare replicate samples.

This article uses the public Kotliarov PBMC CITE-seq subset. The 20
`real_sample` values are biological replicates, and each sample has one
source-study `real_response` value. Cell counts are never treated as replicate
counts for the response comparison.

## Prepare cluster labels and sample metadata

```{r setup-composition}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- qs2::qs_read(real_data_file)

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)
sample_design <- pbmc[[]] |>
  dplyr::distinct(real_sample, real_response) |>
  dplyr::arrange(real_sample)

sample_design
```

The design table above is the source-study metadata used below. Its one-row-per-
sample structure makes the inferential unit explicit.

## Count cells before comparing groups

`sn_calculate_composition()` returns a transparent table with group totals,
category counts, and percentages.

```{r calculate-composition}
composition <- sn_calculate_composition(
  x = pbmc,
  group_by = "real_sample",
  variable = "seurat_clusters",
  min_cells = 10,
  measure = "both"
)

head(composition)
```

The denominator is visible in `group_total`, which makes it easy to audit
whether a proportion is based on enough cells.

Observed-over-expected enrichment is useful when you want to ask which clusters
are enriched or depleted relative to the full contingency table.

```{r calculate-roe}
roe <- sn_calculate_roe(
  x = pbmc,
  group_by = "real_response",
  variable = "seurat_clusters"
)

head(roe)
```

```{r calculate-roe-matrix}
roe_matrix <- sn_calculate_roe(
  x = pbmc,
  group_by = "real_response",
  variable = "seurat_clusters",
  return_matrix = TRUE
)

roe_matrix[, seq_len(min(4, ncol(roe_matrix))), drop = FALSE]
```

```{r plot-composition}
sn_plot_composition(
  composition,
  x = real_sample,
  fill = seurat_clusters,
  y = proportion,
  position = "stack",
  palette = "Paired",
  title = "PBMC cluster composition by real biological sample"
)
```

## Compare sample-level proportions

`sn_compare_composition()` summarizes each sample first, then compares those
sample-level proportions between conditions. That avoids pseudo-replication.

```{r compare-composition}
comparison <- sn_compare_composition(
  x = pbmc,
  sample_by = "real_sample",
  group_by = "real_response",
  variable = "seurat_clusters",
  contrast = c("d0 high", "d0 low"),
  min_cells = 20,
  test = "wilcox"
)

comparison
```

Use the plotting helpers for the summary table when you want a compact
figure-ready view.

```{r plot-comparison}
sn_plot_barplot(
  comparison,
  x = seurat_clusters,
  y = log2_fc,
  fill = change,
  stat = "identity",
  palette = "Set2",
  angle_x = 45,
  y_label = "log2 fold-change",
  title = "Composition shift: high vs low responders"
)
```

## Compare continuous metadata with boxplots

Composition is categorical. For continuous per-cell or per-sample summaries,
summarize to the replicate level first, then plot.

```{r continuous-summaries}
sample_qc <- pbmc[[]] |>
  dplyr::group_by(real_sample, real_response) |>
  dplyr::summarise(
    median_features = median(nFeature_RNA),
    median_mito = median(pctMT),
    .groups = "drop"
  )

sn_plot_boxplot(
  sample_qc,
  x = real_response,
  y = median_features,
  y_label = "Median detected genes",
  title = "Per-sample QC summary by response group"
)
```

## Local differential abundance with miloR

Global proportions can miss local shifts inside an embedding. `sn_run_milo()`
wraps the miloR neighborhood workflow for sample-level differential abundance.
Because miloR is optional, these chunks run only in the `all` profile when the
backend is installed.

```{r run-milo, eval = run_extended && requireNamespace("miloR", quietly = TRUE)}
milo_tbl <- sn_run_milo(
  x = pbmc,
  sample_by = "real_sample",
  group_by = "real_response",
  contrast = c("d0 high", "d0 low"),
  reduction = "pca",
  dims = 1:15,
  annotation_by = "seurat_clusters",
  max_cells = 1500,
  return_object = FALSE
)

head(milo_tbl)
```

The real neighborhood result can be stored and retrieved with the same
object-level convention used for DE and enrichment.

```{r store-milo, eval = run_extended && requireNamespace("miloR", quietly = TRUE)}
pbmc <- sn_store_milo(
  object = pbmc,
  result = milo_tbl,
  store_name = "response_da",
  sample_by = "real_sample",
  group_by = "real_response",
  comparison = "d0_high_vs_d0_low",
  reduction = "pca",
  annotation_by = "seurat_clusters",
  return_object = TRUE
)

head(sn_get_milo_result(pbmc, milo_name = "response_da"))
```

`sn_plot_milo()` then displays the stored neighborhood-level table.

```{r plot-milo, eval = run_extended && requireNamespace("miloR", quietly = TRUE)}
sn_plot_milo(
  pbmc,
  milo_name = "response_da",
  annotation_by = "seurat_clusters",
  title = "Neighborhood differential abundance"
)
```

The recommended reporting pattern is to show both global composition and local
neighborhood abundance when a biological claim depends on changes in cell-state
frequency.
