---
title: "Composition and differential abundance"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composition and differential abundance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Composition analysis should not treat every cell as an independent biological
replicate. Shennong separates cell-level counting from sample-level comparison:
first compute proportions, then compare replicate samples.

This article uses PBMC3k with teaching-only pseudo samples and conditions.

## Prepare cluster labels and sample metadata

```{r setup-composition}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- sn_load_data("pbmc3k")

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

pbmc$sample <- rep(paste0("donor", 1:6), length.out = ncol(pbmc))
pbmc$condition <- ifelse(pbmc$sample %in% paste0("donor", 1:3), "control", "stim")
pbmc$cluster_label <- paste0("cluster_", pbmc$seurat_clusters)
```

The pseudo labels are only for demonstration. In a real analysis, `sample`
should be a biological replicate and `condition` should come from experimental
metadata.

## Count cells before comparing groups

`sn_calculate_composition()` returns a transparent table with group totals,
category counts, and percentages.

```{r calculate-composition}
composition <- sn_calculate_composition(
  x = pbmc,
  group_by = "sample",
  variable = "cluster_label",
  min_cells = 10,
  measure = "both"
)

head(composition)
```

The denominator is visible in `group_total`, which makes it easy to audit
whether a proportion is based on enough cells.

Observed-over-expected enrichment is useful when you want to ask which clusters
are enriched or depleted relative to the full contingency table.

```{r calculate-roe}
roe <- sn_calculate_roe(
  x = pbmc,
  group_by = "condition",
  variable = "cluster_label"
)

head(roe)
```

```{r calculate-roe-matrix}
roe_matrix <- sn_calculate_roe(
  x = pbmc,
  group_by = "condition",
  variable = "cluster_label",
  return_matrix = TRUE
)

roe_matrix[, seq_len(min(4, ncol(roe_matrix))), drop = FALSE]
```

```{r plot-composition}
sn_plot_composition(
  composition,
  x = sample,
  fill = cluster_label,
  y = proportion,
  position = "stack",
  palette = "Paired",
  title = "PBMC3k cluster composition by pseudo donor"
)
```

## Compare sample-level proportions

`sn_compare_composition()` summarizes each sample first, then compares those
sample-level proportions between conditions. That avoids pseudo-replication.

```{r compare-composition}
comparison <- sn_compare_composition(
  x = pbmc,
  sample_by = "sample",
  group_by = "condition",
  variable = "cluster_label",
  contrast = c("stim", "control"),
  min_cells = 20,
  test = "wilcox"
)

comparison
```

Use the plotting helpers for the summary table when you want a compact
figure-ready view.

```{r plot-comparison}
sn_plot_barplot(
  comparison,
  x = cluster_label,
  y = log2_fc,
  fill = change,
  stat = "identity",
  palette = "Set2",
  angle_x = 45,
  y_label = "log2 fold-change",
  title = "Composition shift: stim vs control"
)
```

## Compare continuous metadata with boxplots

Composition is categorical. For continuous per-cell or per-sample summaries,
summarize to the replicate level first, then plot.

```{r continuous-summaries}
sample_qc <- pbmc[[]] |>
  dplyr::group_by(sample, condition) |>
  dplyr::summarise(
    median_features = median(nFeature_RNA),
    median_mito = median(percent.mt),
    .groups = "drop"
  )

sn_plot_boxplot(
  sample_qc,
  x = condition,
  y = median_features,
  y_label = "Median detected genes",
  title = "QC summary by pseudo condition"
)
```

## Local differential abundance with miloR

Global proportions can miss local shifts inside an embedding. `sn_run_milo()`
wraps the miloR neighborhood workflow for sample-level differential abundance.

```{r run-milo, eval = FALSE}
milo_tbl <- sn_run_milo(
  x = pbmc,
  sample_by = "sample",
  group_by = "condition",
  contrast = c("stim", "control"),
  reduction = "pca",
  dims = 1:15,
  annotation_by = "cluster_label",
  max_cells = 1500,
  store_name = "condition_da",
  return_object = FALSE
)

head(milo_tbl)
```

If a milo result was generated elsewhere, store and retrieve it with the same
object-level convention used for DE and enrichment.

```{r store-milo}
mock_milo <- data.frame(
  logFC = c(1.1, -0.7, 0.2),
  SpatialFDR = c(0.03, 0.08, 0.4),
  cluster_label = c("cluster_0", "cluster_1", "cluster_2")
)

pbmc <- sn_store_milo(
  object = pbmc,
  result = mock_milo,
  store_name = "condition_da",
  sample_by = "sample",
  group_by = "condition",
  comparison = "stim_vs_control",
  reduction = "pca",
  annotation_by = "cluster_label",
  return_object = TRUE
)

sn_get_milo_result(pbmc, milo_name = "condition_da")
```

`sn_plot_milo()` then displays the stored neighborhood-level table.

```{r plot-milo}
sn_plot_milo(
  pbmc,
  milo_name = "condition_da",
  annotation_by = "cluster_label",
  title = "Neighborhood differential abundance"
)
```

The recommended reporting pattern is to show both global composition and local
neighborhood abundance when a biological claim depends on changes in cell-state
frequency.
