---
title: "Data input, output, and project setup"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data input, output, and project setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Shennong is meant to make the first step of a single-cell project boring:
find the data, read it with one function, add explicit sample metadata, and
write reusable artifacts without changing APIs for every file type.

This article uses the package PBMC3k example throughout. To keep package checks
and website builds fast, code chunks are shown by default and are evaluated only
when you render with `SHENNONG_RUN_VIGNETTES=true`.

## Start with PBMC3k

The package loader returns a Seurat object by default. If you need the original
10x H5 path instead, set `return_object = FALSE`; this is useful when you want
to show exactly what was read from disk. You can also request several example
datasets at once; filtered matrices are initialized separately and returned as
one merged Seurat object with the source dataset stored in the `sample`
metadata column.

```{r load-pbmc3k}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- sn_load_data(dataset = "pbmc3k")

pbmc_h5 <- sn_load_data(
  dataset = "pbmc3k",
  return_object = FALSE
)

pbmc_h5

pbmc_merged <- sn_load_data(dataset = c("pbmc1k", "pbmc3k"))
table(pbmc_merged$sample)
```

`sn_load_data()` uses the general Zenodo downloader underneath. Public Zenodo
records do not need a token; pass `token = ...` only for restricted or private
records that your account can access.

```{r zenodo-download}
public_h5 <- sn_download_zenodo(
  record_id = "14884845",
  files = "pbmc3k_filtered_feature_bc_matrix.h5"
)

public_h5
```

## Read once, then initialize with metadata

`sn_read()` handles common tabular files, serialized objects, 10x matrices,
H5/H5AD, GMT files, and Shennong's custom dispatchers. For a 10x H5 matrix,
you can read counts directly and then make the metadata decision explicit in
`sn_initialize_seurat_object()`.

```{r read-and-initialize}
counts <- sn_read(pbmc_h5)

pbmc <- sn_initialize_seurat_object(
  x = counts,
  project = "pbmc3k_demo",
  sample_name = "pbmc3k",
  study = "10x_pbmc",
  species = "human"
)

pbmc
```

The important design choice is that provenance is not hidden in the project
name. `sample_name`, `study`, and `species` are separate inputs, so downstream
composition, integration, and interpretation steps can reuse them.

```{r inspect-metadata}
head(pbmc[[]][, c("sample", "study", "nCount_RNA", "nFeature_RNA", "percent.mt")])
```

## Discover 10x folders before reading them

Real projects often start with several Cell Ranger outputs. `sn_list_10x_paths()`
scans a directory tree and returns the paths Shennong can feed back into
`sn_initialize_seurat_object()`.

```{r discover-10x, eval = FALSE}
tenx_paths <- sn_list_10x_paths(
  path = "data/cellranger",
  what = "filtered"
)

pbmc_samples <- sn_initialize_seurat_object(
  x = tenx_paths,
  species = "human",
  study = "pbmc_atlas"
)
```

This pattern keeps import code short while still making the biological metadata
explicit after discovery.

## Write durable intermediate files

Use `sn_write()` and `sn_read()` together for analysis handoffs. The same API
works for tables and serialized R objects; Shennong chooses the writer from the
file extension.

```{r write-read}
outdir <- sn_set_path(file.path(tempdir(), "shennong-pbmc3k-io"))

metadata_path <- file.path(outdir, "pbmc3k_metadata.csv")
object_path <- file.path(outdir, "pbmc3k_initialized.qs")

metadata_export <- cbind(cell = rownames(pbmc[[]]), pbmc[[]])
sn_write(metadata_export, metadata_path, row.names = FALSE)
sn_write(pbmc, object_path)

sn_check_file(c(metadata_path, object_path))

metadata <- sn_read(metadata_path, row_names = "cell")
pbmc_cached <- sn_read(object_path)

dim(metadata)
pbmc_cached
```

`row_names = "cell"` is useful when a CSV carries cell IDs as a regular
column. This avoids ad hoc `read.csv(..., row.names = ...)` calls that are easy
to forget in later scripts.

## Prepare a reusable Zenodo upload

When an intermediate object or reference dataset should be reused by another
project, upload the exact files and a versioned manifest to Zenodo with
`sn_upload_zenodo()`. The default is draft-first: inspect the Zenodo draft, then
rerun with `publish = TRUE` only when the record is ready.

```{r zenodo-dry-run}
zenodo_plan <- sn_upload_zenodo(
  files = c(metadata_path, object_path),
  title = "PBMC3k initialized Shennong example",
  creators = "Duan, Songqi",
  version = "2026.05.04",
  sandbox = TRUE,
  dry_run = TRUE
)

zenodo_plan$files[, c("file", "size", "md5")]
zenodo_plan$manifest_path
```

In a real upload, set `ZENODO_TOKEN` or `ZENODO_SANDBOX_TOKEN` and remove
`dry_run = TRUE`. The uploaded `shennong_zenodo_manifest.json` records the
dataset version, Shennong version, file sizes, and checksums so future analyses
can verify they are reusing the same data release.

## Add metadata exported from AnnData

When a Python workflow exports AnnData metadata or embeddings, use
`sn_add_data_from_anndata()` to merge those files back into the Seurat object
with one explicit import step.

```{r anndata-roundtrip, eval = FALSE}
pbmc <- sn_add_data_from_anndata(
  object = pbmc,
  metadata = "results/scanpy/obs.csv",
  reductions = c(umap = "results/scanpy/X_umap.csv")
)
```

This keeps cross-language handoffs visible in the analysis script instead of
leaving them as undocumented slot edits.

## Initialize a governed analysis project

For a new analysis repository, `sn_initialize_project()` creates a lightweight
project scaffold with Shennong/Codex-oriented governance files. Use it when the
analysis itself needs to be reproducible, not just the R object.

```{r initialize-project, eval = FALSE}
sn_initialize_project(
  path = "~/projects/pbmc3k-shennong-demo",
  project_name = "pbmc3k-shennong-demo"
)
```

Before installing optional backends, inspect what Shennong considers required,
recommended, or optional for your workflow.

```{r dependencies, eval = FALSE}
sn_check_version()

sn_install_shennong(source = "github", ref = "main")

deps <- sn_list_dependencies()
dplyr::count(deps, scope, source)

sn_install_dependencies(scope = "recommended")
```

The package also ships Codex skill assets for Shennong-style analysis projects.
These helpers expose their installed locations without making users search the
package directory manually.

```{r codex-assets, eval = FALSE}
sn_get_codex_skill_path()
sn_install_codex_skill(destination = ".codex/skills")
```
