---
title: "Data input, output, and project setup"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data input, output, and project setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_root <- Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = "")
pbmc_fixture <- if (nzchar(real_data_root)) {
  file.path(real_data_root, "single-cell", "kotliarov_pbmc.qs2")
} else {
  ""
}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") &&
  nzchar(pbmc_fixture) && file.exists(pbmc_fixture)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Shennong begins after data distribution: it reads materialized files, adds
explicit sample metadata, and writes reusable analysis artifacts without
changing APIs for every file type. Dataset discovery, download, caching, and
publication are handled by the separate `ShennongData` package.

This article uses a small, provenance-tracked public PBMC subset materialized
under `SHENNONG_REAL_DATA_DIR`. The data are used locally for validation and
pkgdown output but are intentionally not stored in the Shennong repository.

## Start with materialized public data

The local fixture is a Seurat object prepared from public data. Its acquisition
manifest and checksums live alongside the ignored local data cache, while the
analysis code remains reviewable here.

```{r load-pbmc3k}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- qs2::qs_read(file.path(
  real_data_root,
  "single-cell", "kotliarov_pbmc.qs2"
))

pbmc
table(pbmc$real_sample)
```

## Use ShennongData for discovery and materialization

Shennong does not wrap or re-export the data client. Use qualified
`ShennongData::` calls so the ownership boundary is explicit, inspect the lazy
query, and materialize only the assay/layer required by an analysis.

```{r data-server, eval=FALSE}
connection <- ShennongData::sn_connect(
  url = "https://data.example.org",
  set_default = FALSE
)

remote <- ShennongData::sn_load_data(
  resource = "toil",
  connection = connection
)

rna <- ShennongData::sn_assay(remote, assay = "rna", layer = "expression")
ShennongData::sn_show_query(rna)
materialized <- ShennongData::collect(rna, allow_large = TRUE)
```

Pass `materialized` to `sn_initialize_seurat_object()` or another Shennong
analysis entry point. Data publication and cache policy remain in
`ShennongData`, not the analysis package.

## Read once, then initialize with metadata

`sn_read()` handles common tabular files, serialized objects, 10x matrices,
H5/H5AD, GMT files, and Shennong's custom dispatchers. For a 10x H5 matrix,
you can read counts directly and then make the metadata decision explicit in
`sn_initialize_seurat_object()`.

```{r read-and-initialize}
counts <- SeuratObject::LayerData(pbmc, assay = "RNA", layer = "counts")
metadata <- pbmc[[]]

pbmc <- sn_initialize_seurat_object(
  x = counts,
  metadata = metadata,
  project = "kotliarov_pbmc",
  species = "human"
)

pbmc
```

The important design choice is that provenance is not hidden in the project
name. The materialized metadata retain the real sample and batch fields, while
species is an explicit initialization input.

```{r inspect-metadata}
head(pbmc[[]][, c("real_sample", "real_batch", "nCount_RNA", "nFeature_RNA", "percent.mt")])
```

## Discover 10x folders before reading them

Real projects often start with several Cell Ranger outputs. `sn_list_10x_paths()`
scans a directory tree and returns the paths Shennong can feed back into
`sn_initialize_seurat_object()`.

```{r discover-10x, eval = FALSE}
tenx_paths <- sn_list_10x_paths(
  path = "data/cellranger",
  what = "filtered"
)

pbmc_samples <- sn_initialize_seurat_object(
  x = tenx_paths,
  species = "human",
  study = "pbmc_atlas"
)
```

This pattern keeps import code short while still making the biological metadata
explicit after discovery.

## Write durable intermediate files

Use `sn_write()` and `sn_read()` together for analysis handoffs. The same API
works for tables and serialized R objects; Shennong chooses the writer from the
file extension.

```{r write-read}
outdir <- sn_set_path(file.path(tempdir(), "shennong-kotliarov-io"))

metadata_path <- file.path(outdir, "kotliarov_metadata.csv")
object_path <- file.path(outdir, "kotliarov_initialized.qs2")

metadata_export <- cbind(cell = rownames(pbmc[[]]), pbmc[[]])
sn_write(metadata_export, metadata_path, row.names = FALSE)
sn_write(pbmc, object_path)

sn_check_file(c(metadata_path, object_path))

metadata <- sn_read(metadata_path, row_names = "cell")
pbmc_cached <- sn_read(object_path)

dim(metadata)
pbmc_cached
```

`row_names = "cell"` is useful when a CSV carries cell IDs as a regular
column. This avoids ad hoc `read.csv(..., row.names = ...)` calls that are easy
to forget in later scripts.

For large Seurat objects, keep bulky assay layers on disk with BPCells before
serializing the Seurat object. The returned object keeps the same Seurat layer
interface, but the selected matrices live in BPCells directories.

```{r bpcells-backed-layers, eval = FALSE}
pbmc <- sn_convert_bpcells(
  pbmc,
  directory = file.path(outdir, "pbmc3k_bpcells"),
  layers = c("counts", "data"),
  overwrite = TRUE
)

sn_write(pbmc, file.path(outdir, "pbmc3k_bpcells_bound.qs2"))
```

`sn_initialize_seurat_object()` also accepts a BPCells `IterableMatrix`
directly and preserves that on-disk backend instead of converting the complete
matrix to an in-memory sparse matrix.

```{r initialize-from-bpcells, eval = FALSE}
counts <- BPCells::open_matrix_dir("data/processed/pbmc_counts.bpcells")
pbmc <- sn_initialize_seurat_object(
  x = counts,
  metadata = cell_metadata,
  project = "PBMC"
)
inherits(SeuratObject::LayerData(pbmc, layer = "counts"), "IterableMatrix")
```

## Publish reusable data outside the analysis package

When an intermediate object or reference dataset should be shared, hand the
exact files and their checksums to the `ShennongData` publication workflow.
Shennong intentionally stops at producing validated analysis artifacts; it no
longer owns repository discovery, download, caching, credentials, or remote
publication state.

## Add metadata exported from AnnData

When a Python workflow exports AnnData metadata or embeddings, use
`sn_add_data_from_anndata()` to merge those files back into the Seurat object
with one explicit import step.

```{r anndata-roundtrip, eval = FALSE}
pbmc <- sn_add_data_from_anndata(
  object = pbmc,
  metadata = "results/scanpy/obs.csv",
  reductions = c(umap = "results/scanpy/X_umap.csv")
)
```

This keeps cross-language handoffs visible in the analysis script instead of
leaving them as undocumented slot edits.

## Initialize a governed analysis project

For a new analysis repository, `sn_initialize_project()` creates a lightweight
project scaffold with Shennong/Codex-oriented governance files. Use it when the
analysis itself needs to be reproducible, not just the R object.

```{r initialize-project, eval = FALSE}
sn_initialize_project(
  path = "~/projects/pbmc3k-shennong-demo",
  project_name = "pbmc3k-shennong-demo"
)
```

Before installing optional backends, inspect what Shennong considers required,
recommended, or optional for your workflow.

```{r dependencies, eval = FALSE}
sn_check_version()

sn_install_shennong(source = "github", ref = "main")

deps <- sn_list_dependencies()
dplyr::count(deps, scope, source)

sn_install_dependencies(scope = "recommended")
```

For serialized project objects, prefer `.qs2` files in new workflows. Legacy
`.qs` files remain supported when the archived `qs` package is already present
in the active R library, but `sn_install_dependencies()` does not attempt to
install `qs` on current R releases.

The package also ships Codex skill assets for Shennong-style analysis projects.
These helpers expose their installed locations without making users search the
package directory manually.

```{r codex-assets, eval = FALSE}
sn_get_codex_skill_path()
sn_install_codex_skill(destination = ".codex/skills")
```
