---
title: "Differential expression and enrichment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential expression and enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

Choose the statistical question before choosing a backend. Finding cluster
markers and testing a replicated treatment effect use different sampling units.

| Question | Call | Sampling unit |
|---|---|---|
| Which genes distinguish each cluster? | `sn_find_de(analysis = "markers")` | Cells; exploratory cluster description |
| Which genes differ between two cell groups? | `sn_find_de(analysis = "contrast")` | Cells unless a sample-aware method is selected |
| Which genes change between replicated conditions? | `sn_find_de(analysis = "pseudobulk", sample_by = ...)` | Biological samples |
| Which genes change in a bulk count matrix? | `sn_find_de(counts, metadata = ..., design = ..., contrast = ...)` | Matrix columns / samples |

## Run a complete marker example

This example runs without downloading data. The bundled PBMC object contains
precomputed clusters, which are used only to demonstrate marker extraction.

```{r prepare}
library(Shennong)
data("pbmc_small", package = "SeuratObject")
pbmc <- SeuratObject::CreateSeuratObject(
  SeuratObject::GetAssayData(pbmc_small, assay = "RNA", layer = "counts")
)
pbmc$cluster <- as.character(SeuratObject::Idents(pbmc_small))
pbmc <- sn_normalize_data(pbmc, method = "seurat", verbose = FALSE)

pbmc <- sn_find_de(
  pbmc, analysis = "markers", group_by = "cluster", method = "wilcox",
  assay = "RNA", layer = "data", min_pct = 0.1, logfc_threshold = 0,
  result_id = "markers", verbose = FALSE
)
sn_list_results(pbmc, type = "de")
```

Each cluster is compared with the remaining cells. Keep the full result for
reproducibility, and filter only the table used for reporting:

```{r select}
full <- sn_get_result(pbmc, type = "de", result_id = "markers")
markers <- sn_get_de_result(
  pbmc, "markers", direction = "up", p_adjusted_cutoff = 0.05,
  logfc_threshold = 0.25, top_n = 5, top_scope = "group"
)
head(markers)
full$input[c("assay", "layer", "tested_features_count", "tested_features_source")]
```

Testing filters (`min_pct`, `logfc_threshold` in `sn_find_de`) affect the set
of hypotheses and the tested-gene background. Getter filters affect the output
view only. For enrichment, preserve the tested background rather than using
all genes in the genome or only the significant genes.

## Compare two groups explicitly

`ident_1` is the numerator/target group and `ident_2` is the reference group.
The example compares two existing clusters; it is not a treatment comparison.

```{r contrast}
levels <- sort(unique(pbmc$cluster))
contrast_result <- sn_find_de(
  pbmc, analysis = "contrast", group_by = "cluster",
  ident_1 = levels[[1]], ident_2 = levels[[2]],
  method = "wilcox", layer = "data", min_pct = 0, logfc_threshold = 0,
  result_id = "cluster_contrast", return_object = FALSE, verbose = FALSE
)
head(sn_get_de_result(contrast_result, direction = "up", top_n = 5))
```

Because `return_object = FALSE`, this returns a unified result without adding
it to `pbmc`. Pass that result to the same table getter, or use
`sn_store_result()` to save it later.

## Test replicated conditions with pseudobulk

Use your own multi-sample object for this recipe. It needs raw RNA counts and
metadata columns `sample_id`, `condition`, and `cell_type`. Each `sample_id`
must identify a biological sample with a single condition, and each condition
needs biological replication. Do not invent sample IDs by splitting cells.

```{r pseudobulk, eval=FALSE}
# study is your multi-sample Seurat object, with counts in RNA/counts.
sample_design <- unique(study[[]][, c("sample_id", "condition")])
stopifnot(!anyDuplicated(sample_design$sample_id))
table(sample_design$condition)

study <- sn_find_de(
  study, analysis = "pseudobulk", method = "DESeq2",
  assay = "RNA", layer = "counts",
  group_by = "condition", ident_1 = "treated", ident_2 = "control",
  sample_by = "sample_id", subset_by = "cell_type", subset_levels = "T cell",
  min_cells_per_sample = 20,
  result_id = "T_cell_treated_vs_control"
)
sn_list_results(study, type = "de")
head(sn_get_de_result(study, "T_cell_treated_vs_control", p_adjusted_cutoff = 0.05))
```

The recipe uses DESeq2 through the pseudobulk path. Inspect retained samples
and the design, especially after subsetting rare cell types. For a paired or
covariate-adjusted design, see [bulk transcriptomics](bulk-transcriptomics.html)
and explicitly retain donor/sample metadata. A cell-level Wilcoxon test does
not substitute for a donor-level condition test.

## Choose ORA or GSEA

| Analysis | Input | Interpretation |
|---|---|---|
| ORA | Selected genes and a tested-gene universe | Over-representation among selected genes |
| GSEA | A named numeric ranking or a gene/score table | Enrichment along a ranked list |

`mapping = gene ~ group` describes grouped ORA when `group` is categorical.
`mapping = gene ~ score` selects GSEA when the right side is numeric.
`mapping = gene ~ score | group` describes grouped GSEA. If group labels are
numeric codes, explicitly set `analysis = "ora"` to avoid treating them as
scores. Duplicate genes are rejected by default; choose a documented collapse
rule only when it matches the analysis.

### ORA from stored markers

The following recipes use the `pbmc` created above. They require
**clusterProfiler** and **msigdbr**, and MSigDB may need a first-use download.
They are shown but not executed during ordinary documentation builds. The
small marker example may yield no significant pathways.

```{r ora, eval=FALSE}
pbmc <- sn_run_enrichment(
  pbmc, source_de_result_id = "markers", analysis = "ora",
  species = "human", database = "H",
  de_direction = "up", de_p_adjusted_cutoff = 0.05, de_logfc_threshold = 0.25,
  min_gs_size = 10,
  result_id = "marker_hallmark_ora"
)
sn_list_results(pbmc, type = "enrichment")
terms <- sn_get_enrichment_result(
  pbmc, "marker_hallmark_ora", p_adjusted_cutoff = 0.05, top_n = 10
)
head(terms)
```

`de_*` options select genes from the stored DE analysis before ORA. The tested
gene universe retained by DE is used per comparison when available. An
explicit `universe` replaces that background; supply one only when justified.
Enrichment `pvalue_cutoff` and `qvalue_cutoff` are backend filters, while the
getter applies an adjusted-p-value filter to returned terms.

### GSEA from a complete ranking

Use all tested genes with a meaningful signed ranking statistic, not just
the top markers selected for a figure. This demonstration uses log-fold change;
choose and document the ranking statistic for your study.

```{r gsea, eval=FALSE}
ranking <- contrast_result$tables$primary
ranked <- sn_run_enrichment(
  ranking, mapping = gene ~ avg_log2FC, analysis = "gsea",
  species = "human", database = "H",
  result_id = "contrast_hallmark_gsea", return_object = FALSE
)
head(sn_get_enrichment_result(ranked, top_n = 10))
ranked$parameters
```

For a marker table with one ranking per cluster:

```{r grouped-gsea, eval=FALSE}
grouped <- sn_run_enrichment(
  full$tables$primary, mapping = gene ~ avg_log2FC | cluster,
  analysis = "gsea", species = "human", database = "H",
  return_object = FALSE
)
head(sn_get_enrichment_result(grouped, top_n = 5, top_scope = "group"))
```

### Several databases and their return values

```{r multiple-databases, eval=FALSE}
combined <- sn_run_enrichment(
  ranking, mapping = gene ~ avg_log2FC, analysis = "gsea",
  species = "human", database = c("H", "C2:CP:REACTOME"),
  return_object = FALSE
)
head(combined$tables$primary)       # includes a database column
names(combined$models$database_results)
names(combined$models$backend_results)
```

With `return_object = FALSE`, the combined result contains a primary table
and individual unified results in `models$database_results`; native backend
objects are in `models$backend_results`. With a Seurat input and
`return_object = TRUE`, the workflow stores one result per database. Discover
those IDs with `sn_list_results(pbmc, type = "enrichment")` before retrieving
one. Never infer a database from list position.

## Common problems

| Symptom | Check |
|---|---|
| No genes or terms returned | Inspect the complete table and tested-gene coverage; relax display filters only for inspection |
| Multiple stored results error | List the IDs, then pass the intended `result_id` |
| Result already exists | Use a new ID, or explicitly remove the reviewed old result before rerunning |
| Missing grouping/ranking column | Inspect `names(result$tables$primary)` and the formula mapping |
| Sparse or split Seurat layers | Select the intended assay/layer; see [layer-aware workflows](layered-workflows.html) |
| Unexpected enrichment background | Inspect the DE tested-gene provenance and any explicit `universe` |

Continue with [annotation and pathways](annotation-pathways.html) for feature
classes, reference annotation, plotting, and scoring on a larger real dataset.
