---
title: "Get started: counts to markers and scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started: counts to markers and scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                     warning = FALSE, fig.width = 7, fig.height = 4.5)
```

Run this page from top to bottom to create a Seurat object, cluster cells,
find markers, score a gene set, and retrieve the saved results. It uses the
small PBMC dataset included in **SeuratObject**, so no data download, Python
runtime, API key, or local fixture is needed. The tiny dataset demonstrates
usage; it is not suitable for condition-level biological conclusions.

## 1. Install and load

Run installation once. Shennong is under development; the documentation tracks
GitHub `main`.

```{r install, eval=FALSE}
install.packages(c("remotes", "Seurat"))
remotes::install_github("zerostwo/shennong")
```

```{r load}
library(Shennong)
data("pbmc_small", package = "SeuratObject")
```

## 2. Start from counts

Expression matrices have **genes in rows and cells in columns**. Cell names
must match the row names of any metadata you add. Recreate the example from
counts so the steps below do not reuse its precomputed analysis.

```{r counts}
counts <- SeuratObject::GetAssayData(pbmc_small, assay = "RNA", layer = "counts")
pbmc <- SeuratObject::CreateSeuratObject(counts = counts, project = "quickstart")
dim(pbmc)
head(pbmc[[]])
```

For your own data, replace `counts` with an imported matrix. For example,
`counts <- sn_read("path/to/filtered_feature_bc_matrix")` reads a 10x directory.
For multimodal 10x input, select the gene-expression matrix from the returned
list before creating an RNA object. See [data input and output](data-io-projects.html)
for metadata alignment, backed matrices, and file formats.

## 3. Normalize, reduce dimensions, and cluster

`sn_run_cluster()` performs normalization, variable-feature selection, PCA,
neighbors, clustering, and UMAP. Set `integration_method = "unintegrated"`
for a baseline analysis. Here the feature and dimension counts are deliberately
small because the bundled example is small.

```{r cluster}
pbmc <- sn_run_cluster(
  pbmc,
  assay = "RNA", layer = "counts",
  normalization_method = "seurat",
  integration_method = "unintegrated",
  nfeatures = 100, npcs = 10, dims = 1:10,
  resolution = 1.2, seed = 717,
  verbose = FALSE
)
table(pbmc$seurat_clusters)
```

`assay` chooses the modality; `layer` chooses the expression matrix within it.
`npcs` controls how many PCs are computed and `dims` selects those used
downstream. Cluster labels are computational groups, not cell-type annotations.

```{r plot-clusters}
sn_plot_dim(pbmc, reduction = "umap", group_by = "seurat_clusters")
```

For multiple batches, first inspect a baseline and the experimental design.
Then add `batch_by = "batch"` and choose an integration method. See
[clustering and integration](clustering.html) for batch-aware examples and
[parameter conventions](parameters-and-results.html#choose-the-input-and-grouping)
for the distinct roles of `batch_by`, `group_by`, and `sample_by`.

## 4. Find markers and select a table

The analysis stores all tested results. The getter selects the rows you want
to display. These two steps use separate filters.

```{r markers}
pbmc <- sn_find_de(
  pbmc, analysis = "markers", group_by = "seurat_clusters",
  assay = "RNA", layer = "data", method = "wilcox",
  min_pct = 0.1, logfc_threshold = 0,
  result_id = "cluster_markers", verbose = FALSE
)

markers <- sn_get_de_result(
  pbmc, result_id = "cluster_markers",
  direction = "up", top_n = 3, top_scope = "group"
)
head(markers)
```

This selects up to three positive markers **per cluster**. Add
`p_adjusted_cutoff = 0.05` and `logfc_threshold = 0.25` to the getter to select
significant effects for a report. Use `top_scope = "all"` for one top-N list
across the selected groups. A significance filter may return zero rows on a
small dataset; that is a valid result.

## 5. Score a gene set

A signature is a named list of gene vectors. This example computes the mean
normalized expression of an illustrative B-cell marker panel. Scoring a panel
does not by itself establish a cell identity.

```{r scores}
signatures <- list(B_cell = c("MS4A1", "CD79A", "CD79B"))
pbmc <- sn_score_programs(
  pbmc, signatures = signatures, method = "mean",
  assay = "RNA", layer = "data", seed = 717,
  result_id = "marker_panel"
)

scores <- sn_get_result(pbmc, type = "program_scoring", result_id = "marker_panel")
head(scores$tables$primary)
scores$tables$coverage
```

Inspect `tables$coverage` to see which genes were measured. Missing genes are
not negative expression evidence. For rank-based UCell scores, install UCell
and use `method = "ucell"`. Group summaries require an explicit order of
operations; see [group scoring](parameters-and-results.html#score-cells-or-group-profiles).

## 6. Discover, retrieve, and save

```{r retrieve}
sn_list_results(pbmc)
full_de <- sn_get_result(pbmc, type = "de", result_id = "cluster_markers")
full_de$parameters
head(full_de$tables$primary)
```

`sn_list_results()` lists stored analyses, `sn_get_result()` returns one complete
result, and table getters such as `sn_get_de_result()` return a data frame.
Always assign the returned object when you want to retain a new analysis.

```{r save, eval=FALSE}
saveRDS(pbmc, "pbmc-analysed.rds")
pbmc <- readRDS("pbmc-analysed.rds")
sn_list_results(pbmc)
```

| Next task | Read next |
|---|---|
| Select layers, parameters, result IDs, or overwrite behavior | [Parameters and results](parameters-and-results.html) |
| Test markers or replicated conditions, then enrich pathways | [Differential expression and enrichment](differential-expression.html) |
| Filter cells, normalize counts, or correct ambient RNA | [Preprocessing and QC](preprocessing-qc.html) |
| Integrate batches and assess the embedding | [Clustering](clustering.html) and [diagnostics](metrics-diagnostics.html) |
| Make figures from expression or stored results | [Visualization](visualization.html) |
| Choose a spatial, bulk, or dynamics workflow | [All workflows](index.html) |
