---
title: "Interpretation and reporting"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpretation and reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Shennong's interpretation layer is intentionally downstream of analysis. It
does not replace markers, enrichment, QC, or clustering; it packages those
results into structured evidence, then builds prompts or calls a provider only
when the user asks.

This article uses PBMC3k markers as the evidence source. Provider calls are
shown but not executed by default.

## Create analysis evidence first

```{r setup-evidence}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- sn_load_data("pbmc3k")

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

pbmc <- sn_find_de(
  object = pbmc,
  analysis = "markers",
  group_by = "seurat_clusters",
  layer = "data",
  store_name = "cluster_markers",
  return_object = TRUE,
  verbose = FALSE
)
```

The interpretation functions expect stored results. That keeps prompts
reproducible: a report can say which DE result, enrichment result, and cluster
column were used.

```{r list-stored-results}
sn_list_results(pbmc)
```

## Prepare evidence before building prompts

Evidence helpers return ordinary R lists and tables. Inspect them before using
an LLM; this is the step that keeps interpretation transparent.

```{r prepare-evidence}
annotation_evidence <- sn_prepare_annotation_evidence(
  object = pbmc,
  de_name = "cluster_markers",
  cluster_by = "seurat_clusters",
  n_markers = 8,
  include_qc = TRUE,
  reduction = "umap"
)

de_evidence <- sn_prepare_de_evidence(
  object = pbmc,
  de_name = "cluster_markers",
  n_genes = 10
)

results_evidence <- sn_prepare_results_evidence(
  object = pbmc,
  cluster_de_name = "cluster_markers",
  cluster_by = "seurat_clusters",
  n_markers = 5
)

names(annotation_evidence)
head(annotation_evidence$cluster_summary)
```

If enrichment has been run, add pathway evidence the same way.

```{r enrichment-evidence, eval = FALSE}
pbmc <- sn_enrich(
  x = pbmc,
  source_de_name = "cluster_markers",
  species = "human",
  database = "GOBP",
  store_name = "cluster_gobp",
  return_object = TRUE
)

enrichment_evidence <- sn_prepare_enrichment_evidence(
  object = pbmc,
  enrichment_name = "cluster_gobp",
  n_terms = 8
)
```

## Build a prompt without calling a model

`sn_build_prompt()` is useful even when you do not plan to call a model from R.
It gives you a model-ready or human-readable representation of the exact
evidence bundle.

```{r build-prompt}
annotation_prompt <- sn_build_prompt(
  evidence = annotation_evidence,
  task = "annotation",
  audience = "scientist",
  language = "en",
  background = "PBMC3k single-cell RNA-seq clustered with Shennong.",
  output_format = "llm",
  include_json_schema = TRUE
)

names(annotation_prompt)
annotation_prompt$task
```

The high-level helper can return the same prompt directly.

```{r interpret-annotation-prompt}
annotation_prompt <- sn_interpret_annotation(
  object = pbmc,
  de_name = "cluster_markers",
  cluster_by = "seurat_clusters",
  return_prompt = TRUE,
  output_format = "llm",
  background = "Annotate PBMC3k clusters conservatively."
)

annotation_prompt$task
```

## Provider calls are opt-in

Shennong uses provider functions at the package boundary. You can create an
`ellmer` provider from environment variables, test it, then pass it to the
interpretation helpers.

```{r provider, eval = FALSE}
provider <- sn_make_ellmer_provider(
  model = Sys.getenv("OPENAI_MODEL", "gpt-5.1"),
  reasoning_effort = "medium"
)

sn_test_llm_provider(provider = provider)

response <- sn_run_llm(
  messages = annotation_prompt$messages,
  provider = provider,
  model = Sys.getenv("OPENAI_MODEL", "gpt-5.1")
)
```

## Interpret specific evidence types

The DE and enrichment helpers use the same pattern: return a prompt by default
for review, or pass a provider to store a response.

```{r interpret-de-enrichment}
de_prompt <- sn_interpret_de(
  object = pbmc,
  de_name = "cluster_markers",
  n_genes = 10,
  return_prompt = TRUE,
  background = "Summarize PBMC3k cluster marker programs."
)

de_prompt$task
```

```{r interpret-enrichment, eval = FALSE}
enrichment_prompt <- sn_interpret_enrichment(
  object = pbmc,
  enrichment_name = "cluster_gobp",
  n_terms = 8,
  return_prompt = TRUE,
  background = "Summarize pathway programs by cluster."
)
```

## Write results, legends, and presentation summaries

The reporting helpers assemble multiple evidence sources into audience-specific
outputs. They can return prompts for manual review or store generated responses
under `object@misc$interpretation_results`.

```{r reporting-prompts}
results_prompt <- sn_write_results(
  object = pbmc,
  cluster_de_name = "cluster_markers",
  cluster_by = "seurat_clusters",
  return_prompt = TRUE,
  background = "Write a concise Results paragraph for PBMC3k clustering."
)

legend_prompt <- sn_write_figure_legend(
  object = pbmc,
  cluster_de_name = "cluster_markers",
  cluster_by = "seurat_clusters",
  return_prompt = TRUE,
  background = "Describe a UMAP and marker dot plot."
)

summary_prompt <- sn_write_presentation_summary(
  object = pbmc,
  cluster_de_name = "cluster_markers",
  cluster_by = "seurat_clusters",
  return_prompt = TRUE,
  background = "Prepare a short collaborator-facing PBMC3k summary."
)

c(results_prompt$task, legend_prompt$task, summary_prompt$task)
```

When a provider is used, retrieve stored outputs by name.

```{r retrieve-interpretation, eval = FALSE}
pbmc <- sn_interpret_de(
  object = pbmc,
  de_name = "cluster_markers",
  provider = provider,
  store_name = "cluster_marker_summary",
  return_object = TRUE
)

sn_get_interpretation_result(
  pbmc,
  interpretation_name = "cluster_marker_summary"
)
```

The practical workflow is: compute evidence, inspect evidence, build a prompt,
then call a provider only when the prompt is good enough to send.
