---
title: "Interpretation and reporting workflow"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpretation and reporting workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") ||
  identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  eval = run_vignette
)
```

This article shows how the new interpretation layer fits on top of the current
Shennong analysis workflow. The key idea is that analysis stays deterministic,
while interpretation reuses stored DE and enrichment results to build evidence
bundles, prompts, and optional LLM-style summaries.

The workflow below demonstrates:

- marker discovery with `sn_find_de()`
- enrichment storage with `sn_enrich()` / `sn_store_enrichment()`
- result discovery with `sn_list_results()`
- direct retrieval with `sn_get_de_result()` and `sn_get_enrichment_result()`
- evidence preparation for annotation, DE, and writing tasks
- prompt generation with `sn_build_prompt()`
- provider-based interpretation with `sn_interpret_annotation()`

```{r setup}
library(Shennong)
library(dplyr)
library(knitr)
library(Seurat)
```

```{r prepare-interpretation-demo, include = FALSE}
pbmc <- sn_load_data("pbmc1k")

pbmc <- suppressWarnings(
  sn_run_cluster(
    object = pbmc,
    normalization_method = "seurat",
    species = "human",
    nfeatures = 800,
    block_genes = NULL,
    npcs = 20,
    dims = 1:15,
    resolution = 0.5,
    verbose = FALSE
  )
)

pbmc <- suppressWarnings(
  sn_find_de(
    pbmc,
    analysis = "markers",
    group_by = "seurat_clusters",
    layer = "data",
    logfc_threshold = 0.25,
    min_pct = 0.2,
    store_name = "cluster_markers",
    return_object = TRUE,
    verbose = FALSE
  )
)

cluster_markers <- sn_get_de_result(
  pbmc,
  de_name = "cluster_markers",
  with_metadata = TRUE
)$table

cluster_zero_markers <- cluster_markers |>
  dplyr::filter(.data$cluster == unique(.data$cluster)[1]) |>
  dplyr::arrange(dplyr::desc(.data$avg_log2FC))

real_enrichment <- tryCatch(
  sn_enrich(
    cluster_zero_markers,
    gene_clusters = gene ~ avg_log2FC,
    species = "human",
    database = "GOBP",
    pvalue_cutoff = 1
  ),
  error = function(...) NULL
)

enrichment_tbl <- if (!is.null(real_enrichment) && nrow(as.data.frame(real_enrichment)) > 0) {
  as.data.frame(real_enrichment) |>
    dplyr::select(ID, Description, NES, p.adjust)
} else {
  tibble::tibble(
    ID = c("GO:0006955", "GO:0002250"),
    Description = c("immune response", "adaptive immune response"),
    NES = c(1.9, 1.6),
    p.adjust = c(0.02, 0.05)
  )
}

pbmc <- sn_store_enrichment(
  object = pbmc,
  result = enrichment_tbl,
  store_name = "cluster0_gsea",
  analysis = "gsea",
  database = "GOBP",
  species = "human",
  source_de_name = "cluster_markers",
  return_object = TRUE
)

annotation_evidence <- sn_prepare_annotation_evidence(
  object = pbmc,
  de_name = "cluster_markers",
  cluster_col = "seurat_clusters",
  n_markers = 5
)

de_evidence <- sn_prepare_de_evidence(
  object = pbmc,
  de_name = "cluster_markers",
  n_genes = 5
)

enrichment_evidence <- sn_prepare_enrichment_evidence(
  object = pbmc,
  enrichment_name = "cluster0_gsea",
  n_terms = 5
)

results_evidence <- sn_prepare_results_evidence(
  object = pbmc,
  cluster_de_name = "cluster_markers",
  enrichment_name = "cluster0_gsea",
  cluster_col = "seurat_clusters",
  n_markers = 3,
  n_terms = 5
)

annotation_prompt <- sn_build_prompt(
  evidence = annotation_evidence,
  task = "annotation",
  style = "concise annotation note",
  audience = "scientist"
)

results_prompt <- sn_write_results(
  object = pbmc,
  cluster_de_name = "cluster_markers",
  enrichment_name = "cluster0_gsea",
  cluster_col = "seurat_clusters",
  return_prompt = TRUE
)

mock_provider <- function(messages, model = NULL, ...) {
  list(
    text = paste0(
      '{"cluster_annotations":[',
      '{"cluster":"0","primary_label":"T Cell","broad_label":"Lymphocyte","confidence":"high","status":"confident",',
      '"alternatives":["Activated T Cell"],"supporting_markers":["CD3D","TRAC"],',
      '"supporting_functions":["T cell activation"],"risk_flags":["transitional_state"],',
      '"note":"Canonical T-cell markers dominate this cluster.","recommended_checks":["Inspect IL7R and LTB"]}',
      '],"narrative_summary":"Mock structured annotation generated for the vignette."}'
    )
  )
}

cfg_dir <- file.path(tempdir(), ".shennong")
sn_configure_llm_provider(
  name = "demo-local",
  base_url = "https://api.example.invalid",
  api_key = "demo-key",
  default_model = "gpt-5.4",
  wire_api = "responses",
  config_dir = cfg_dir
)

knitr::kable(sn_list_llm_providers(config_dir = cfg_dir))

sn_test_llm_provider(provider = mock_provider)

pbmc_interpreted <- sn_interpret_annotation(
  object = pbmc,
  de_name = "cluster_markers",
  enrichment_name = "cluster0_gsea",
  cluster_col = "seurat_clusters",
  n_markers = 5,
  provider = mock_provider,
  model = "demo-model",
  store_name = "cluster_annotation_note",
  metadata_prefix = "shennong_celltype",
  return_object = TRUE
)

annotation_response <- pbmc_interpreted@misc$interpretation_results$cluster_annotation_note$response$text
annotation_table <- pbmc_interpreted@misc$interpretation_results$cluster_annotation_note$annotation_table
result_index <- sn_list_results(pbmc_interpreted)
top_cluster_markers <- sn_get_de_result(
  pbmc_interpreted,
  de_name = "cluster_markers",
  top_n = 3,
  groups = unique(cluster_markers$cluster)[1]
)
top_pathways <- sn_get_enrichment_result(
  pbmc_interpreted,
  enrichment_name = "cluster0_gsea",
  top_n = 5
)
```

## Marker discovery remains the analysis layer

Interpretation starts from stored DE results rather than recomputing
statistics. Here, `sn_find_de()` has already saved cluster markers into the
Seurat object, and we retrieve them through `sn_get_de_result()`.

```{r}
knitr::kable(top_cluster_markers |> dplyr::select(cluster, gene, avg_log2FC, p_val_adj))
```

## Store enrichment for later interpretation

The interpretation layer expects enrichment results to be stored in the Seurat
object. `sn_enrich()` can now store them directly when you pass a Seurat object
through `x` or `object =`, and its `gene_clusters` formula now covers both
grouped ORA inputs such as `gene ~ cluster` and ranked GSEA inputs such as

When `sn_interpret_annotation()` receives a structured JSON response, it also
stores the parsed cluster-level annotation table and can map the labels back
onto each cell's metadata. In the example above, that means columns such as
`shennong_celltype_label`, `shennong_celltype_confidence`, and
`shennong_celltype_risk_flags` are available directly in `pbmc_interpreted[[]]`.
`gene ~ avg_log2FC`.

```{r}
knitr::kable(top_pathways)
```

## Discover what is currently stored on the object

Use `sn_list_results()` to see which DE, enrichment, and interpretation
artifacts are available for reuse.

```{r}
knitr::kable(result_index)
```

## Prepare structured evidence

`sn_prepare_*_evidence()` turns stored outputs into compact, reusable evidence
bundles. These are the objects that can later be turned into prompts or passed
into writing helpers.

```{r}
knitr::kable(annotation_evidence$cluster_summary)
```

```{r}
knitr::kable(enrichment_evidence$top_terms)
```

## Build prompts without binding to a specific provider

The first-class package object for the LLM layer is a prompt bundle, not a
network call. This keeps analysis and interpretation cleanly separated.

```{r}
cat(substr(annotation_prompt$user, 1, 900))
```

## Generate manuscript-style prompts

High-level helpers such as `sn_write_results()` can return prompts directly.

## Initialize a Codex-ready analysis project

For longer-running projects, it helps to initialize a governed analysis
repository rather than improvising the folder structure. The packaged project
template used by `sn_initialize_project(with_agent = TRUE)` creates
`AGENTS.md`, `.gitignore`, a project `.Rproj`, `memory/`,
`docs/standards/`, `skills/`, `config/`, `data/`, `scripts/`, `notebooks/`,
`runs/`, and `results/`.

```{r}
codex_project <- file.path(tempdir(), "shennong-codex-project")

created <- sn_initialize_project(
  path = codex_project,
  project_name = "PBMC interpretation pilot",
  objective = "Build a reproducible PBMC interpretation workflow with Shennong.",
  with_agent = TRUE,
  overwrite = TRUE
)

knitr::kable(
  tibble::tibble(
    file = names(created)[-1],
    path = unname(unlist(created[-1]))
  )
)
```

The initialized project keeps durable operating rules in `AGENTS.md`, project
state in `memory/`, enforceable directory and naming rules in
`docs/standards/BioinformaticsAnalysisConventions.md`, and reusable governance
procedures in `skills/`.

```{r}
cat(substr(results_prompt$user, 1, 900))
```

## Plug in a provider when needed

If you provide a function that accepts `messages` and returns text, the same
high-level helpers can store the generated interpretation back into the Seurat
object.

For repeated interactive work, you can also register a reusable default
provider under `~/.shennong` and then let `sn_interpret_annotation()` resolve
 it automatically when `provider = NULL`. Shennong stores provider metadata in
 `~/.shennong/llm-providers.json` and, for configured providers, request
 history in `~/.shennong/llm-history/`.

```{r}
annotation_response
```

## Interpretation results are stored alongside analysis results

```{r}
names(pbmc_interpreted@misc$interpretation_results)
```
