---
title: "Layer-aware workflows"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layer-aware workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_file <- file.path(
  Sys.getenv("SHENNONG_REAL_DATA_DIR"),
  "single-cell",
  "kotliarov_pbmc.qs2"
)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_requested <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")
run_vignette <- run_requested &&
  requireNamespace("qs2", quietly = TRUE) &&
  file.exists(real_data_file)
run_extended <- run_vignette && identical(real_profile, "all")
celltypist_model <- Sys.getenv("SHENNONG_CELLTYPIST_MODEL")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Seurat v5 objects can store multiple assay layers. Shennong exposes `assay` and
`layer` in the functions where the count source matters, so workflows can use
raw counts, decontaminated counts, or other matrix layers without rewriting the
analysis.

This article uses real RNA and ADT count layers from the public Kotliarov PBMC
CITE-seq fixture. It does not copy counts into a pretend corrected layer. A
deterministic, sample-stratified subset keeps the layer examples fast while
preserving cells from all 20 biological samples.

## Inspect and subset real assay layers

```{r setup-layer}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- qs2::qs_read(real_data_file)
counts_full <- SeuratObject::LayerData(pbmc, assay = "RNA", layer = "counts")
demo_features <- names(sort(Matrix::rowSums(counts_full), decreasing = TRUE))[
  seq_len(min(1000, nrow(counts_full)))
]
demo_cells <- unlist(lapply(
  split(colnames(pbmc), as.character(pbmc$real_sample)),
  utils::head,
  n = 20
), use.names = FALSE)
pbmc <- subset(pbmc, cells = demo_cells)
pbmc[["RNA"]] <- subset(
  pbmc[["RNA"]],
  features = demo_features,
  cells = demo_cells
)

data.frame(
  assay = c("RNA", "ADT"),
  layers = c(
    paste(SeuratObject::Layers(pbmc[["RNA"]]), collapse = ", "),
    paste(SeuratObject::Layers(pbmc[["ADT"]]), collapse = ", ")
  ),
  features = c(nrow(pbmc[["RNA"]]), nrow(pbmc[["ADT"]])),
  cells = ncol(pbmc)
)
```

The object retains measured RNA and ADT counts. Normalized `data` layers are
created only by the workflows that require them.

## Cluster from a selected layer

`sn_run_cluster()` temporarily uses the requested layer for the workflow and
restores the original count layer before returning the object.

```{r cluster-layer}
pbmc_layered <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  assay = "RNA",
  layer = "counts",
  nfeatures = 500,
  dims = 1:10,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

sn_plot_dim(
  pbmc_layered,
  reduction = "umap",
  group_by = "seurat_clusters",
  label = TRUE,
  title = "Real PBMC cells clustered from the RNA counts layer"
)
```

## Run DE on the right layer

Marker and contrast analyses usually use normalized data; pseudobulk analyses
usually use counts. Shennong keeps that distinction explicit through `layer`.

```{r de-layer}
pbmc_layered <- sn_find_de(
  object = pbmc_layered,
  analysis = "markers",
  group_by = "seurat_clusters",
  assay = "RNA",
  layer = "data",
  store_name = "cluster_markers",
  return_object = TRUE,
  verbose = FALSE
)

sn_get_de_result(
  pbmc_layered,
  de_name = "cluster_markers",
  top_n = 3,
  with_metadata = TRUE
)
```

The stored metadata records the analysis context, which helps downstream
plotting and interpretation recover how the result was made.

## Use the same layer in other workflows

Layer selection appears in functions where using the wrong matrix would change
the answer.

```{r filter-selected-layer, eval = run_vignette}
pbmc_layered_filtered <- sn_filter_genes(
  pbmc_layered,
  min_cells = 5,
  assay = "RNA",
  layer = "counts",
  species = "human",
  plot = FALSE
)

data.frame(
  features_before = nrow(pbmc_layered),
  features_after = nrow(pbmc_layered_filtered)
)
```

CellTypist and ROGUE are optional backends. They run only in the `all` profile
when the executable/model or R dependency is already available.

```{r layer-celltypist, eval = run_extended && nzchar(celltypist_model) && file.exists(celltypist_model) && nzchar(Sys.which("celltypist"))}
pbmc_celltypist <- sn_run_celltypist(
  x = pbmc_layered,
  assay = "RNA",
  layer = "counts",
  model = celltypist_model,
  over_clustering = "seurat_clusters",
  quiet = TRUE
)

grep("celltypist", colnames(pbmc_celltypist[[]]), value = TRUE, ignore.case = TRUE)
```

```{r layer-rogue, eval = run_extended && requireNamespace("ROGUE", quietly = TRUE)}
rogue_tbl <- sn_calculate_rogue(
  pbmc_layered,
  cluster_by = "seurat_clusters",
  sample_by = "real_sample",
  assay = "RNA",
  layer = "counts"
)

rogue_tbl
```

## Simulate from the current object

Simulation is useful when you need a controlled object with the same covariate
structure as a real experiment. `sn_simulate()` keeps the API method-based, so
new simulation engines can be added later. With `method = "scdesign3"`, it
converts the selected Seurat layer to a SingleCellExperiment, calls scDesign3,
and returns a simulated Seurat object by default. It does not merge original
and simulated cells unless `combine_original = TRUE`.

This optional simulation backend runs only in the `all` profile when
`scDesign3` is installed.

```{r scdesign3-layer, eval = run_extended && requireNamespace("scDesign3", quietly = TRUE)}
sim_pbmc <- sn_simulate(
  object = pbmc_layered,
  method = "scdesign3",
  celltype = "seurat_clusters",
  ncell = 100,
  assay = "RNA",
  layer = "counts",
  n_cores = 1,
  return = "seurat"
)

sim_pbmc
sim_pbmc@misc$scdesign3

sim_counts <- sn_simulate_scdesign3(
  object = pbmc_layered,
  celltype = "seurat_clusters",
  ncell = 50,
  assay = "RNA",
  layer = "counts",
  n_cores = 1,
  return = "counts"
)

dim(sim_counts)
summary(Matrix::colSums(sim_counts))
```

## Cache layer-aware artifacts

Layer-aware results are still regular Seurat objects and tables, so the same IO
helpers from the first article apply.

```{r cache-layered}
outdir <- sn_set_path(file.path(tempdir(), "shennong-kotliarov-layered"))

object_path <- file.path(outdir, "kotliarov_layered.qs2")
markers_path <- file.path(outdir, "kotliarov_layered_markers.csv")

sn_write(pbmc_layered, object_path)
sn_write(
  sn_get_de_result(pbmc_layered, "cluster_markers", top_n = 5),
  markers_path
)

sn_check_file(c(object_path, markers_path))
```

The main habit is to name the layer in every step where it matters. That makes
the workflow readable months later and reduces the chance of silently switching
between raw and corrected matrices.
