---
title: "Layer-aware workflows"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layer-aware workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Seurat v5 objects can store multiple assay layers. Shennong exposes `assay` and
`layer` in the functions where the count source matters, so workflows can use
raw counts, decontaminated counts, or other matrix layers without rewriting the
analysis.

This article uses PBMC3k and creates a demonstration layer by copying counts.
In real work, that layer might come from decontX, SoupX, CellBender, or another
preprocessing tool.

## Create a demonstration layer

```{r setup-layer}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- sn_load_data("pbmc3k")

counts <- SeuratObject::LayerData(pbmc, assay = "RNA", layer = "counts")
SeuratObject::LayerData(pbmc, assay = "RNA", layer = "decontaminated_counts") <- counts

SeuratObject::Layers(pbmc[["RNA"]])
```

The object still has its original counts. The new layer is simply another input
matrix that layer-aware Shennong functions can target.

## Cluster from a selected layer

`sn_run_cluster()` temporarily uses the requested layer for the workflow and
restores the original count layer before returning the object.

```{r cluster-layer}
pbmc_layered <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  assay = "RNA",
  layer = "decontaminated_counts",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

sn_plot_dim(
  pbmc_layered,
  reduction = "umap",
  group_by = "seurat_clusters",
  label = TRUE,
  title = "PBMC3k clustered from decontaminated_counts"
)
```

## Run DE on the right layer

Marker and contrast analyses usually use normalized data; pseudobulk analyses
usually use counts. Shennong keeps that distinction explicit through `layer`.

```{r de-layer}
pbmc_layered <- sn_find_de(
  object = pbmc_layered,
  analysis = "markers",
  group_by = "seurat_clusters",
  assay = "RNA",
  layer = "data",
  store_name = "cluster_markers",
  return_object = TRUE,
  verbose = FALSE
)

sn_get_de_result(
  pbmc_layered,
  de_name = "cluster_markers",
  top_n = 3,
  with_metadata = TRUE
)
```

The stored metadata records the analysis context, which helps downstream
plotting and interpretation recover how the result was made.

## Use the same layer in other workflows

Layer selection appears in functions where using the wrong matrix would change
the answer.

```{r layer-aware-functions, eval = FALSE}
pbmc_layered <- sn_filter_genes(
  pbmc_layered,
  min_cells = 5,
  assay = "RNA",
  layer = "decontaminated_counts",
  species = "human"
)

pbmc_layered <- sn_run_celltypist(
  x = pbmc_layered,
  assay = "RNA",
  layer = "decontaminated_counts",
  over_clustering = "seurat_clusters",
  quiet = TRUE
)

rogue_tbl <- sn_calculate_rogue(
  pbmc_layered,
  cluster = "seurat_clusters",
  assay = "RNA",
  layer = "decontaminated_counts"
)
```

## Cache layer-aware artifacts

Layer-aware results are still regular Seurat objects and tables, so the same IO
helpers from the first article apply.

```{r cache-layered}
outdir <- sn_set_path(file.path(tempdir(), "shennong-pbmc3k-layered"))

object_path <- file.path(outdir, "pbmc3k_layered.qs")
markers_path <- file.path(outdir, "pbmc3k_layered_markers.csv")

sn_write(pbmc_layered, object_path)
sn_write(
  sn_get_de_result(pbmc_layered, "cluster_markers", top_n = 5),
  markers_path
)

sn_check_file(c(object_path, markers_path))
```

The main habit is to name the layer in every step where it matters. That makes
the workflow readable months later and reduces the chance of silently switching
between raw and corrected matrices.
