---
title: "Layer-aware workflow"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layer-aware workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") ||
  identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  eval = run_vignette
)
```

This article demonstrates three Shennong behaviors that are important in
modern multi-layer Seurat workflows:

- automatic species inference when `species` is not provided
- running analysis from a non-default count layer
- storing DE metadata in `object@misc$de_results`

```{r setup}
library(Shennong)
library(dplyr)
library(knitr)
library(Seurat)
```

```{r prepare-layered-demo, include = FALSE}
pbmc <- sn_load_data("pbmc1k")
counts <- SeuratObject::LayerData(pbmc, layer = "counts")
inferred_species <- sn_get_species(counts)

pbmc_inferred <- sn_initialize_seurat_object(
  x = counts,
  project = "pbmc1k_inferred"
)

SeuratObject::LayerData(pbmc_inferred, layer = "decontaminated_counts") <- counts

pbmc_layered <- suppressWarnings(
  sn_run_cluster(
    object = pbmc_inferred,
    normalization_method = "seurat",
    layer = "decontaminated_counts",
    nfeatures = 800,
    npcs = 20,
    dims = 1:15,
    resolution = 0.5,
    verbose = FALSE
  )
)

pbmc_layered <- suppressWarnings(
  sn_find_de(
    pbmc_layered,
    analysis = "markers",
    group_by = "seurat_clusters",
    store_name = "cluster_markers",
    return_object = TRUE,
    verbose = FALSE
  )
)

de_store <- pbmc_layered@misc$de_results$cluster_markers
de_summary <- tibble::tibble(
  field = c(
    "schema_version", "package_version", "analysis", "group_by",
    "rank_col", "p_col", "n_genes"
  ),
  value = c(
    de_store$schema_version,
    de_store$package_version,
    de_store$analysis,
    de_store$group_by,
    de_store$rank_col,
    de_store$p_col,
    as.character(de_store$n_genes)
  )
)
```

## Automatic species inference

The count matrix is initialized without an explicit `species` argument, but
Shennong still infers the species from feature names using `hom_genes` and
common mitochondrial patterns.

```{r}
knitr::kable(
  tibble::tibble(
    object = c("raw_counts", "pbmc_inferred"),
    species = c(inferred_species, sn_get_species(pbmc_inferred))
  )
)
```

## Layer-aware clustering

The object below is clustered from the `decontaminated_counts` layer rather
than the default `counts` layer.

```{r}
sn_plot_dim(
  pbmc_layered,
  reduction = "umap",
  group_by = "seurat_clusters",
  label = TRUE,
  show_legend = FALSE,
  title = "pbmc1k clustered from the decontaminated_counts layer"
)
```

## Stored differential expression metadata

`sn_find_de(..., return_object = TRUE)` stores both the marker table and the
analysis metadata required for later reuse in plotting and enrichment.

```{r}
knitr::kable(de_summary)
```

## Reusing stored markers

```{r}
sn_plot_dot(
  pbmc_layered,
  features = "top_markers",
  de_name = "cluster_markers",
  n = 3
)
```
