---
title: "Integration metrics and diagnostics"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration metrics and diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_file <- file.path(
  Sys.getenv("SHENNONG_REAL_DATA_DIR"),
  "single-cell",
  "kotliarov_pbmc.qs2"
)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_requested <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")
run_vignette <- run_requested &&
  requireNamespace("qs2", quietly = TRUE) &&
  file.exists(real_data_file)
run_extended <- run_vignette && identical(real_profile, "all")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Good integration is not just a pretty UMAP. A useful diagnostic layer should
separate three questions:

- Are technical labels mixed?
- Are biological labels still separated?
- Which clusters or groups are difficult enough to inspect manually?

Shennong exposes both one-line aggregate diagnostics and individual metrics.
This article uses the public Kotliarov PBMC CITE-seq subset. `real_batch` is a
source acquisition label, `real_sample` identifies the 20 biological samples,
and RNA-versus-WNN clusters provide a real cross-modality comparison without
inventing reference labels.

## Prepare a diagnostic object

```{r setup-object}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- qs2::qs_read(real_data_file)

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

pbmc$rna_clusters <- as.character(pbmc$seurat_clusters)

pbmc_wnn <- sn_run_multimodal(
  object = qs2::qs_read(real_data_file),
  method = "wnn",
  assay = "RNA",
  adt_assay = "ADT",
  nfeatures = 1500,
  dims = 1:15,
  adt_dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)
pbmc$wnn_clusters <- as.character(pbmc_wnn$seurat_clusters)[
  match(colnames(pbmc), colnames(pbmc_wnn))
]

table(pbmc$rna_clusters, pbmc$wnn_clusters)

sn_plot_dim(
  pbmc,
  reduction = "umap",
  group_by = "real_batch",
  title = "Real acquisition batches before integration"
)
```

## Start with the aggregate assessment

`sn_assess_integration()` runs a compact panel of metrics and returns the
results in predictable sections: `summary`, `per_cell`, `per_group`, and
`parameters`. Here the dependency-free panel evaluates the unintegrated PCA
space; the same call can be pointed at an integrated reduction later.

```{r aggregate-assessment}
assessment <- sn_assess_integration(
  x = pbmc,
  batch_by = "real_batch",
  label_by = "wnn_clusters",
  cluster_by = "rna_clusters",
  reduction = "pca",
  baseline_reduction = NULL,
  dims = 1:15,
  max_cells = 1500,
  metrics = c(
    "batch_silhouette", "label_silhouette", "graph_connectivity",
    "clustering_agreement", "isolated_label_score", "cluster_purity",
    "cluster_entropy", "pcr_batch", "challenging_groups"
  )
)

assessment$summary
names(assessment$per_group)
sn_plot_integration(assessment)
```

Use the wrapper for routine reporting, then call individual metrics when one
part of the assessment needs more detail.

## Inspect mixing and separation directly

LISI asks how diverse each cell's local neighborhood is with respect to a
label. The upstream `lisi` package is optional, so this real-data chunk runs
only in the `all` profile when that package is installed.

```{r lisi, eval = run_extended && requireNamespace("lisi", quietly = TRUE)}
batch_lisi <- sn_calculate_lisi(
  pbmc,
  reduction = "pca",
  label_by = "real_batch",
  dims = 1:15,
  max_cells = 1500
)

cluster_lisi <- sn_calculate_lisi(
  pbmc,
  reduction = "pca",
  label_by = "wnn_clusters",
  dims = 1:15,
  max_cells = 1500
)

head(batch_lisi)
head(cluster_lisi)
```

Silhouette asks whether a label is compact in the embedding. For biological
labels, high silhouette can be good; for batch labels, high silhouette often
means residual batch separation.

```{r silhouette}
cluster_sil <- sn_calculate_silhouette(
  pbmc,
  label_by = "rna_clusters",
  reduction = "pca",
  dims = 1:15,
  max_cells = 1500
)

head(cluster_sil)
```

Graph connectivity asks whether cells with the same label remain connected in
the neighbor graph.

```{r graph-connectivity}
connectivity <- sn_calculate_graph_connectivity(
  pbmc,
  label_by = "rna_clusters",
  reduction = "pca",
  dims = 1:15,
  max_cells = 1500
)

connectivity
```

## Compare clusters to labels when references exist

Purity, entropy, and clustering agreement can also compare independently
computed clusterings. Here RNA-only clusters are compared with WNN clusters
learned from the aligned RNA and ADT assays; neither is mislabeled as ground
truth.

```{r supervised-metrics}
sn_calculate_cluster_purity(
  pbmc,
  cluster_by = "rna_clusters",
  label_by = "wnn_clusters"
)

sn_calculate_cluster_entropy(
  pbmc,
  cluster_by = "rna_clusters",
  label_by = "wnn_clusters"
)

sn_calculate_clustering_agreement(
  pbmc,
  cluster_by = "rna_clusters",
  label_by = "wnn_clusters"
)

sn_calculate_isolated_label_score(
  pbmc,
  label_by = "wnn_clusters",
  reduction = "pca",
  dims = 1:15,
  max_cells = 1500
)
```

When a curated reference label becomes available, repeat the same calls with
that label and report its provenance. The cross-modality comparison above is a
stability diagnostic, not an accuracy claim.

## Quantify residual batch signal

PCR batch scoring measures how much variation in the embedding is explained by
batch. Supplying a baseline reduction lets you compare correction against the
pre-integration PCA space.

```{r pcr-batch}
pcr <- sn_calculate_pcr_batch(
  pbmc,
  batch_by = "real_batch",
  reduction = "pca",
  baseline_reduction = NULL,
  dims = 1:15,
  max_cells = 1500
)

pcr
```

When several metadata variables might drive the same residual structure, rank
them directly with variance-explained scoring. Use the single-variable mode for
a quick screen, then the partial mode when the design is not fully confounded.

```{r variance-explained}
variance_drivers <- sn_calculate_variance_explained(
  pbmc,
  variables = c("real_batch", "real_response", "real_sample"),
  reduction = "pca",
  dims = 1:15,
  max_cells = 1500
)

variance_drivers
```

## Flag difficult groups

`sn_identify_challenging_groups()` combines rarity, connectivity, and local
mixing diagnostics to surface groups that deserve manual review. This is useful
before final annotation.

```{r challenging-groups}
challenging <- sn_identify_challenging_groups(
  pbmc,
  group_by = "rna_clusters",
  reduction = "pca",
  dims = 1:15,
  max_cells = 1500
)

challenging
```

ROGUE is an optional cluster-purity diagnostic that depends on the upstream
`ROGUE` package. Keep it as a targeted check rather than a mandatory build step.

```{r rogue, eval = run_extended && requireNamespace("ROGUE", quietly = TRUE)}
rogue_tbl <- sn_calculate_rogue(
  pbmc,
  cluster_by = "rna_clusters",
  sample_by = "real_sample",
  max_cells = 1500
)

rogue_tbl
```

The intended workflow is: use the aggregate table for a quick read, then drill
into single metrics only when a value changes a downstream decision.
