---
title: "Metrics and diagnostics workflow"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metrics and diagnostics workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") ||
  identical(Sys.getenv("IN_PKGDOWN"), "true")

required_pkgs <- c("Seurat", "harmony")
run_vignette <- run_vignette && all(vapply(
  required_pkgs,
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  eval = run_vignette
)
```

This article focuses on the diagnostic layer that sits after clustering or
integration. In practice, Shennong metrics fall into three scientific groups:

- batch removal: LISI, batch silhouette, PCR batch score
- biological conservation: graph connectivity, isolated-label preservation
- structure diagnostics: challenging groups, cluster entropy, cluster purity

```{r setup}
library(Shennong)
library(dplyr)
library(knitr)
library(Seurat)

if (!exists("pbmc_small", inherits = FALSE)) {
  try(data("pbmc_small", package = "Shennong", envir = environment()), silent = TRUE)
}
if (!exists("pbmc_small", inherits = FALSE) && file.exists(file.path("data", "pbmc_small.rda"))) {
  load(file.path("data", "pbmc_small.rda"))
}
```

```{r prepare-metrics-demo, include = FALSE}
pbmc_unintegrated <- suppressWarnings(
  sn_run_cluster(
    pbmc_small,
    normalization_method = "seurat",
    hvg_group_by = "sample",
    nfeatures = 800,
    npcs = 20,
    dims = 1:15,
    resolution = 0.4,
    verbose = FALSE
  )
)

pbmc_integrated <- suppressWarnings(
  sn_run_cluster(
    pbmc_small,
    batch = "sample",
    normalization_method = "seurat",
    hvg_group_by = "sample",
    nfeatures = 800,
    npcs = 20,
    dims = 1:15,
    resolution = 0.4,
    verbose = FALSE
  )
)

batch_lisi <- tryCatch(
  sn_calculate_lisi(
    pbmc_integrated,
    reduction = "harmony",
    label = "sample"
  ),
  error = function(...) NULL
)

assessment <- sn_assess_integration(
  pbmc_integrated,
  batch = "sample",
  label = "seurat_clusters",
  cluster = "seurat_clusters",
  reduction = "harmony",
  baseline_reduction = "pca"
)

summary_tbl <- assessment$summary |>
  dplyr::filter(
    .data$metric %in% c(
      "batch_silhouette",
      "batch_lisi",
      "pcr_batch",
      "cluster_batch_entropy",
      "well_resolved_group_fraction",
      "overall_integration_score"
    )
  )

challenging_tbl <- assessment$per_group$challenging_groups
isolated_tbl <- assessment$per_group$isolated_label_score
entropy_tbl <- assessment$per_group$cluster_entropy
rare_tbl <- sn_detect_rare_cells(
  pbmc_integrated,
  method = "gini",
  nfeatures = 100,
  max_fraction = 0.05
)
```

## Run the aggregate assessment first

`sn_assess_integration()` is the recommended entry point because it bundles the
main metrics into one object with summary, per-cell, and per-group outputs.
When the question is specifically "which Seurat resolution gives the most
defensible partition?", use `sn_sweep_cluster_resolution()` to compare
candidate resolutions with silhouette width, graph connectivity, optional
reference-label agreement, and optional ROGUE support.

```{r}
knitr::kable(summary_tbl, digits = 3)
```

## Inspect local sample mixing with LISI

Higher sample LISI indicates better local batch mixing. When `lisi` is
installed, Shennong stores the per-cell values so you can examine the full
distribution rather than relying on a single mean.

```{r}
if (!is.null(batch_lisi)) {
  knitr::kable(head(batch_lisi, 10), digits = 3)
}
```

## Surface rare or difficult groups

The challenging-group summary is useful for small populations that do not form
an obvious island in UMAP but still look unstable in neighbor structure or
silhouette space.

```{r}
knitr::kable(challenging_tbl, digits = 3)
```

## Score candidate rare cells directly

`sn_detect_rare_cells()` provides a cell-level complement to the group-level
diagnostics. The native `gini` backend uses rare-gene enrichment to score
cells, while optional backends such as Python-based scCAD can be used when
installed locally.

```{r}
knitr::kable(head(rare_tbl[order(rare_tbl$rare_score, decreasing = TRUE), ], 10), digits = 3)
```

## Inspect isolated-label preservation

`sn_calculate_isolated_label_score()` and the corresponding table inside
`sn_assess_integration()` highlight low-frequency labels or clusters that
remain well-separated after integration.

```{r}
knitr::kable(isolated_tbl, digits = 3)
```

## Quantify batch mixing inside each cluster

Cluster entropy is a practical complement to LISI because it answers a simpler
question: within a given cluster, are the batches actually mixed?

```{r}
knitr::kable(entropy_tbl, digits = 3)
```

## Add supervised metrics when labels are available

If you already have a trusted annotation column such as `cell_type`, you can
extend the same workflow with supervised conservation metrics:

```r
assessment <- sn_assess_integration(
  pbmc_integrated,
  batch = "sample",
  label = "cell_type",
  cluster = "seurat_clusters",
  reduction = "harmony",
  baseline_reduction = "pca"
)
```

That enables `label_lisi`, ARI/NMI, cluster purity against `cell_type`, and a
more interpretable isolated-label summary.
