---
title: "Integration metrics and diagnostics"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration metrics and diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Good integration is not just a pretty UMAP. A useful diagnostic layer should
separate three questions:

- Are technical labels mixed?
- Are biological labels still separated?
- Which clusters or groups are difficult enough to inspect manually?

Shennong exposes both one-line aggregate diagnostics and individual metrics.
This article uses PBMC3k and a teaching-only pseudo-batch column.

## Prepare a diagnostic object

```{r setup-object}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- sn_load_data("pbmc3k")
pbmc$library <- rep(c("library_a", "library_b"), length.out = ncol(pbmc))

pbmc <- sn_run_cluster(
  object = pbmc,
  batch = "library",
  normalization_method = "seurat",
  hvg_group_by = "library",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)
```

## Start with the aggregate assessment

`sn_assess_integration()` runs a compact panel of metrics and returns the
results in predictable sections: `summary`, `per_cell`, `per_group`, and
`parameters`.

```{r aggregate-assessment}
assessment <- sn_assess_integration(
  x = pbmc,
  batch = "library",
  label = "seurat_clusters",
  cluster = "seurat_clusters",
  reduction = "harmony",
  baseline_reduction = "pca",
  dims = 1:15,
  max_cells = 1500
)

assessment$summary
names(assessment$per_group)
```

Use the wrapper for routine reporting, then call individual metrics when one
part of the assessment needs more detail.

## Inspect mixing and separation directly

LISI asks how diverse each cell's local neighborhood is with respect to a
label. Here the same function can measure library mixing or cluster mixing by
changing `label`.

```{r lisi}
library_lisi <- sn_calculate_lisi(
  pbmc,
  reduction = "harmony",
  label = "library",
  dims = 1:15,
  max_cells = 1500
)

cluster_lisi <- sn_calculate_lisi(
  pbmc,
  reduction = "harmony",
  label = "seurat_clusters",
  dims = 1:15,
  max_cells = 1500
)

head(library_lisi)
head(cluster_lisi)
```

Silhouette asks whether a label is compact in the embedding. For biological
labels, high silhouette can be good; for batch labels, high silhouette often
means residual batch separation.

```{r silhouette}
cluster_sil <- sn_calculate_silhouette(
  pbmc,
  label = "seurat_clusters",
  reduction = "harmony",
  dims = 1:15,
  max_cells = 1500
)

head(cluster_sil)
```

Graph connectivity asks whether cells with the same label remain connected in
the neighbor graph.

```{r graph-connectivity}
connectivity <- sn_calculate_graph_connectivity(
  pbmc,
  label = "seurat_clusters",
  reduction = "harmony",
  dims = 1:15,
  max_cells = 1500
)

connectivity
```

## Compare clusters to labels when references exist

When you have a trusted label column, purity, entropy, and clustering agreement
show how the computed clusters relate to that reference.

```{r supervised-metrics}
pbmc$coarse_label <- paste0("cluster_", pbmc$seurat_clusters)

sn_calculate_cluster_purity(
  pbmc,
  cluster = "seurat_clusters",
  label = "coarse_label"
)

sn_calculate_cluster_entropy(
  pbmc,
  cluster = "seurat_clusters",
  label = "coarse_label"
)

sn_calculate_clustering_agreement(
  pbmc,
  cluster = "seurat_clusters",
  label = "coarse_label"
)

sn_calculate_isolated_label_score(
  pbmc,
  label = "coarse_label",
  reduction = "harmony",
  dims = 1:15,
  max_cells = 1500
)
```

The artificial label above is deliberately simple. In a real workflow, replace
it with cell-type annotations, reference-mapping labels, or curated manual
labels.

## Quantify residual batch signal

PCR batch scoring measures how much variation in the embedding is explained by
batch. Supplying a baseline reduction lets you compare correction against the
pre-integration PCA space.

```{r pcr-batch}
pcr <- sn_calculate_pcr_batch(
  pbmc,
  batch = "library",
  reduction = "harmony",
  baseline_reduction = "pca",
  dims = 1:15,
  max_cells = 1500
)

pcr
```

## Flag difficult groups

`sn_identify_challenging_groups()` combines rarity, connectivity, and local
mixing diagnostics to surface groups that deserve manual review. This is useful
before final annotation.

```{r challenging-groups}
challenging <- sn_identify_challenging_groups(
  pbmc,
  group = "seurat_clusters",
  reduction = "harmony",
  dims = 1:15,
  max_cells = 1500
)

challenging
```

ROGUE is an optional cluster-purity diagnostic that depends on the upstream
`ROGUE` package. Keep it as a targeted check rather than a mandatory build step.

```{r rogue, eval = FALSE}
rogue_tbl <- sn_calculate_rogue(
  pbmc,
  cluster = "seurat_clusters",
  sample = "library",
  max_cells = 1500
)

rogue_tbl
```

The intended workflow is: use the aggregate table for a quick read, then drill
into single metrics only when a value changes a downstream decision.
