---
title: "Parameters and results: common patterns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameters and results: common patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

Use this page when deciding where an option belongs, how to retrieve a result,
or how to rerun an analysis. The executable examples use a bundled dataset;
the backend recipes state their additional requirements.

## Prepare the example

```{r prepare}
library(Shennong)
data("pbmc_small", package = "SeuratObject")
pbmc <- SeuratObject::CreateSeuratObject(
  SeuratObject::GetAssayData(pbmc_small, assay = "RNA", layer = "counts")
)
# Preserve the example clustering only to demonstrate grouped API calls.
pbmc$example_group <- as.character(SeuratObject::Idents(pbmc_small))
pbmc <- sn_normalize_data(pbmc, method = "seurat", verbose = FALSE)
signatures <- list(B_cell = c("MS4A1", "CD79A", "CD79B"))
```

## Choose the input and grouping

Not every function accepts every control. Use `args(sn_run_cluster)` or its
reference page for the exact signature.

| Parameter | Meaning | Example |
|---|---|---|
| `assay` | Assay containing the selected modality | `"RNA"`, `"ADT"` |
| `layer` | Matrix within that assay | `"counts"` for count models; `"data"` for normalized scoring |
| `group_by` | Metadata column defining groups to compare or summarize | `"cell_type"`, `"condition"` |
| `sample_by` | Metadata column defining the biological sampling unit | `"sample_id"`; preserve donor identity in paired designs |
| `batch_by` | Technical batch column used for integration | `"library_batch"` |
| `hvg_group_by` | Groups used to balance variable-feature selection | `"sample_id"` |
| `seed` | Random seed for the workflow | `717` |
| `n_workers` | Worker count, on functions that expose parallel execution | `4` |
| `backend_control` | Named list of backend-specific options | See examples below |

These `*_by` arguments usually take **a column name**, not a vector of labels.
Inspect `colnames(pbmc[[]])` first. A batch is not automatically a biological
replicate, and integrating an embedding does not remove the need for a
sample-aware statistical design. If condition and batch are confounded,
integration cannot identify their separate effects.

## Store an analysis or return a result

DE, enrichment, program scoring, and Milo use the following pattern:

| Call | Return value | How to use it |
|---|---|---|
| `return_object = TRUE` | Updated Seurat object | Assign it back to retain the result |
| `return_object = FALSE` | Unified analysis result | Read `tables$primary`, parameters, and diagnostics |
| `sn_get_result(object, type, result_id)` | Complete stored result | Inspect inputs, tables, models, and provenance |
| `sn_get_de_result(object, result_id, ...)` | Selected DE table | Filter direction, significance, groups, and top-N |

```{r standalone}
result <- sn_score_programs(pbmc, signatures, method = "mean",
                            result_id = "panel", return_object = FALSE)
head(result$tables$primary)
result$parameters
```

The original `pbmc` has no stored scoring run yet. Store the returned result
explicitly, or call the workflow with `return_object = TRUE`:

```{r store}
pbmc <- sn_store_result(pbmc, type = "program_scoring",
                        result_id = "panel", result = result)
sn_list_results(pbmc, type = "program_scoring")
restored <- sn_get_result(pbmc, type = "program_scoring", result_id = "panel")
stopifnot(identical(restored$tables$primary, result$tables$primary))
```

Generic storage saves the result envelope. It does not replay a workflow or
add its per-cell metadata columns. Use the object-returning scoring call when
you need those columns for plotting.

## Select IDs and rerun deliberately

`result_id` identifies an analysis **within its result type**. Omit it when
retrieving only if exactly one result of that type is stored. With multiple
candidates, the getter raises an error listing their IDs. Use
`sn_list_results()` to choose the intended analysis; list order does not mean
"latest" or "best".

Scoring and Milo assign unique default IDs (`programs_mean`,
`programs_mean.2`; `milo`, `milo.2`). Automatic enrichment IDs include the
source, analysis, and database. DE defaults to `"default"`; give repeated DE
runs distinct explicit IDs.

```{r rerun}
pbmc <- sn_score_programs(pbmc, signatures, method = "mean", result_id = "panel_v2")
sn_list_results(pbmc, type = "program_scoring")

# Deliberately replace this exact scoring result and its owned score columns.
pbmc <- sn_score_programs(
  pbmc, signatures, method = "mean",
  result_id = "panel_v2", overwrite = TRUE
)
```

Existing results are protected against accidental replacement.
`sn_store_result()` requires `overwrite = TRUE` for an existing type/ID pair.
Workflows exposing `overwrite` include scoring, Milo, and fate; do not pass
this option to a function that does not support it. For DE and other writers,
use a new ID, or explicitly delete a reviewed result before recomputing it:

```{r delete}
pbmc <- sn_delete_result(pbmc, type = "program_scoring", result_id = "panel")
sn_list_results(pbmc, type = "program_scoring")
```

`artifact_id` refers to a runtime or backend artifact, not a stored analysis.
For auditing older objects and exporting result bundles, see
[advanced result management](analysis-results-and-methods.html).

## Filter after testing

```{r de}
pbmc <- sn_find_de(pbmc, analysis = "markers", group_by = "example_group",
                   method = "wilcox", min_pct = 0, logfc_threshold = 0,
                   result_id = "markers", verbose = FALSE)

# All negative effects; direction works without top_n.
down <- sn_get_de_result(pbmc, "markers", direction = "down")

# Up to three positive genes per group, after significance/effect filtering.
selected <- sn_get_de_result(
  pbmc, "markers", direction = "up",
  p_adjusted_cutoff = 0.05, logfc_threshold = 0.25,
  top_n = 3, top_scope = "group"
)
head(selected)

# Three genes total across the selected groups.
head(sn_get_de_result(pbmc, "markers", top_n = 3, top_scope = "all"))
```

On `sn_find_de()`, `min_pct` and `logfc_threshold` determine which genes are
tested. On `sn_get_de_result()`, thresholds select already tested rows.
Filtering a table does not change the stored result or rerun the model.
`direction = "up"` means a positive effect for the selected comparison;
for a contrast, read it relative to `ident_1` versus `ident_2`.

`sn_get_enrichment_result()` uses the same `top_scope`, `groups`, and
`p_adjusted_cutoff` pattern. `top_n` is per group by default. If the result has
no group column, leave `groups = NULL`.

## Score cells or group profiles

Grouping requires `aggregate` because the order of operations matters:

| Setting | Calculation | Use when |
|---|---|---|
| No `group_by` | Score each cell | You need cell-level heterogeneity |
| `aggregate = "scores"` | Score cells, then average within each group | You want the average cell score |
| `aggregate = "expression"` | Average expression, then score each group profile | You want a score of the group expression profile |

```{r aggregation}
mean_cell_scores <- sn_score_programs(
  pbmc, signatures, method = "mean", group_by = "example_group",
  aggregate = "scores", return_object = FALSE
)
group_profile_scores <- sn_score_programs(
  pbmc, signatures, method = "mean", group_by = "example_group",
  aggregate = "expression", return_object = FALSE
)
mean_cell_scores$tables$primary
group_profile_scores$tables$primary
```

For a linear mean score these operations agree. For rank-based or nonlinear
methods such as UCell, AUCell, GSVA, or ssGSEA they generally differ. To compare
conditions, summarize by genuine biological samples and use the appropriate
design; averaging over cells does not create replication.

## Set backend-specific options

Common controls stay at the top level. Backend-specific lists follow the
selected workflow: integration takes a flat list for one method; program
scoring uses a method-named sublist.

```{r backend-scoring, eval=requireNamespace("UCell", quietly=TRUE)}
ucell <- sn_score_programs(
  pbmc, signatures, method = "ucell", seed = 717,
  backend_control = list(ucell = list(maxRank = 200)),
  return_object = FALSE
)
head(ucell$tables$primary)
```

The integration recipe below requires your multi-batch object and the managed
scVI environment; it is not run on the small example.

```{r backend-integration, eval=FALSE}
# multi_batch must contain a real technical-batch column named batch.
integrated <- sn_run_cluster(
  multi_batch, batch_by = "batch", integration_method = "scvi",
  assay = "RNA", layer = "counts", dims = 1:30, seed = 717,
  backend_control = list(max_epochs = 100)
)
# sn_run_scvi() is a shortcut to the same full clustering workflow.
```

For supervised scANVI/scPoli, put the label column in
`backend_control = list(label_by = "cell_type", ...)`. Use
`sn_get_method_status()` and the [clustering guide](clustering.html) for runtime
setup. The raw `sn_call_*` adapters have their own backend argument contracts;
they do not take a Seurat object.

## Update scripts from earlier development versions

These development API changes are intentional breaking changes.

| Earlier usage | Current usage |
|---|---|
| Clustering `batch` | `batch_by` |
| `integration_control`, public `method_control` | `backend_control` |
| Public `ncores`, `n_cores`, `n_jobs` on renamed wrappers | `n_workers`; native backend list keys remain backend-specific |
| `cluster_random_seed` | Top-level `seed` |
| Scoring `backend_control$seed` | Top-level `seed` |
| DE `p_val_cutoff`, `de_logfc` | Getter `p_adjusted_cutoff`, `logfc_threshold` |
| Getter `with_metadata = TRUE` | `sn_get_result()` for the full result |
| Grouped scoring without an aggregation choice | Set `aggregate = "scores"` or `"expression"` |
| Milo `return_intermediate` | `keep_model = TRUE`, then `result$models$milo` |
| Assuming `return_object = FALSE` is a data frame | Use a table getter or `result$tables$primary` |

If a call fails, first check its current signature with `args()`, list the
object metadata with `colnames(object[[]])`, and discover stored analyses with
`sn_list_results(object)`. See [release notes](../news/index.html) for the full
set of breaking changes.
