---
title: "Preprocessing and QC with PBMC3k"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocessing and QC with PBMC3k}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Preprocessing should answer four questions before clustering:

1. What object did we create from the count matrix?
2. Which low-quality cells were flagged, and why?
3. Which genes are retained for downstream analysis?
4. What QC evidence can be stored and revisited later?

Shennong keeps those decisions explicit while keeping the code short. The
examples use PBMC3k and are evaluated only when
`SHENNONG_RUN_VIGNETTES=true`.

## Load PBMC3k and inspect species-aware QC fields

`sn_load_data("pbmc3k")` downloads or reuses the cached 10x H5 file, reads it,
and initializes a Seurat object with human QC metrics.

```{r load-pbmc3k}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- sn_load_data("pbmc3k")
pbmc$sample <- "pbmc3k"

sn_get_species(pbmc)

head(pbmc[[]][, c("sample", "nCount_RNA", "nFeature_RNA", "percent.mt")])
```

If you start from a raw matrix instead, the equivalent explicit form is:

```{r initialize-from-counts, eval = FALSE}
counts <- sn_read("pbmc3k_filtered_feature_bc_matrix.h5")

pbmc <- sn_initialize_seurat_object(
  x = counts,
  project = "pbmc3k_demo",
  sample_name = "pbmc3k",
  study = "10x_pbmc",
  species = "human"
)
```

When gene identifiers or mixed symbol styles are a concern, standardize at
initialization or call `sn_standardize_gene_symbols()` before downstream
analysis. If `HGNChelper` cannot provide an unambiguous replacement for an
otherwise valid symbol, Shennong preserves the original symbol instead of
returning an `NA` feature name.

```{r standardize-symbols, eval = FALSE}
pbmc <- sn_standardize_gene_symbols(
  pbmc,
  species = "human",
  is_gene_id = FALSE
)
```

## Filter cells with inspectable QC flags

`sn_filter_cells()` uses median absolute deviation thresholds by default. The
key point is that filtering and flagging are separated: set `filter = FALSE` to
see which cells would be removed, then set `filter = TRUE` when the rule is
acceptable.

```{r flag-cells}
pbmc_flagged <- sn_filter_cells(
  x = pbmc,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  method = "mad",
  n = c(5, 5, 3),
  plot = FALSE,
  filter = FALSE
)

qc_flag_cols <- grep("^qc_", colnames(pbmc_flagged[[]]), value = TRUE)
head(pbmc_flagged[[]][, qc_flag_cols, drop = FALSE])
```

Once the flags look sensible, apply the same rule:

```{r filter-cells}
pbmc_cells <- sn_filter_cells(
  x = pbmc,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  method = "mad",
  n = c(5, 5, 3),
  plot = FALSE,
  filter = TRUE
)

c(before = ncol(pbmc), after = ncol(pbmc_cells))
```

## Filter genes by expression and annotation

`sn_filter_genes()` keeps the threshold and optional annotation rule in one
place. The expression threshold protects sparse downstream steps, while
`gene_class = "coding"` can remove noncoding genes when your workflow is
protein-coding focused.

```{r filter-genes}
pbmc_genes <- sn_filter_genes(
  x = pbmc_cells,
  min_cells = 5,
  plot = FALSE,
  filter = TRUE,
  species = "human",
  gene_class = "coding"
)

c(before = nrow(pbmc_cells), after = nrow(pbmc_genes))
```

## Normalize and add cell-cycle scores

`sn_normalize_data()` and `sn_score_cell_cycle()` keep preprocessing inside the
same Shennong vocabulary used later by clustering and plotting.

```{r normalize-cycle}
pbmc_norm <- sn_normalize_data(
  object = pbmc_genes,
  normalization_method = "seurat",
  verbose = FALSE
)

pbmc_norm <- sn_score_cell_cycle(pbmc_norm, species = "human")

table(pbmc_norm$Phase)
```

## Optional doublet and ambient-RNA correction

Doublet detection and ambient correction depend on optional backends. Shennong
keeps them as explicit steps rather than hiding them in clustering.

```{r doublets, eval = FALSE}
pbmc_doublets <- sn_find_doublets(pbmc_norm)
table(pbmc_doublets$doublet_class)
```

For ambient RNA, PBMC3k raw counts can be loaded from the same example registry.
The corrected matrix is written to a new layer by default, so the original
counts remain inspectable.

```{r ambient, eval = FALSE}
pbmc_raw <- sn_load_data("pbmc3k", matrix_type = "raw")

pbmc_decont <- sn_remove_ambient_contamination(
  x = pbmc_norm,
  raw = pbmc_raw,
  method = "decontx",
  layer = "decontaminated_counts",
  verbose = FALSE
)
```

## Store a QC assessment

`sn_assess_qc()` summarizes the object after filtering and can compare it to
the pre-filter object. Store the result when you want later reports to recover
the QC decision without rerunning preprocessing.

```{r assess-qc}
qc_report <- sn_assess_qc(
  object = pbmc_norm,
  reference = pbmc,
  sample_col = "sample",
  verbose = FALSE
)

qc_report$overall

pbmc_norm <- sn_assess_qc(
  object = pbmc_norm,
  reference = pbmc,
  sample_col = "sample",
  store_name = "pbmc3k_preprocessing",
  return_object = TRUE,
  verbose = FALSE
)

names(pbmc_norm@misc$qc_assessments)
```

The object is now ready for clustering. The important part is not just that
cells and genes were filtered, but that the rationale and summary are stored in
a place downstream code can find.
