---
title: "Preprocessing and QC with real PBMC CITE-seq data"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocessing and QC with real PBMC CITE-seq data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_file <- file.path(
  Sys.getenv("SHENNONG_REAL_DATA_DIR"),
  "single-cell",
  "kotliarov_pbmc.qs2"
)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_requested <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")
run_vignette <- run_requested &&
  requireNamespace("qs2", quietly = TRUE) &&
  file.exists(real_data_file)
run_extended <- run_vignette && identical(real_profile, "all")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Preprocessing should answer four questions before clustering:

1. What object did we create from the count matrix?
2. Which low-quality cells were flagged, and why?
3. Which genes are retained for downstream analysis?
4. What QC evidence can be stored and revisited later?

Shennong keeps those decisions explicit while keeping the code short. The
examples use 2,000 real cells sampled deterministically across 20 biological
samples from the public Kotliarov PBMC CITE-seq cohort. The RNA and ADT assays
remain aligned. Ordinary package checks do not evaluate the article when the
local fixture is unavailable; a real-data pkgdown build sets
`SHENNONG_RUN_VIGNETTES=true`.

## Load the local PBMC fixture and inspect species-aware QC fields

Data discovery and materialization are handled by `ShennongData`. This article
reads the already materialized, provenance-tracked local fixture; Shennong then
owns the analysis steps.

```{r load-kotliarov-pbmc}
library(Shennong)
library(Seurat)
library(dplyr)

pbmc <- qs2::qs_read(real_data_file)

sn_get_species(pbmc)

head(pbmc[[]][, c(
  "real_sample", "real_response", "real_batch",
  "nCount_RNA", "nFeature_RNA", "pctMT", "nCount_ADT"
)])

data.frame(
  assay = names(pbmc@assays),
  features = vapply(pbmc@assays, nrow, numeric(1)),
  cells = vapply(pbmc@assays, ncol, numeric(1))
)
```

The initialization API can be exercised on the real RNA UMI matrix without
introducing another example dataset:

```{r initialize-from-counts, eval = run_vignette}
counts <- SeuratObject::LayerData(pbmc, assay = "RNA", layer = "counts")

pbmc_from_counts <- sn_initialize_seurat_object(
  x = counts,
  project = "kotliarov_pbmc",
  sample_name = "kotliarov_subset",
  study = "Kotliarov_2020",
  species = "human"
)

data.frame(
  cells = ncol(pbmc_from_counts),
  features = nrow(pbmc_from_counts),
  assay = SeuratObject::DefaultAssay(pbmc_from_counts)
)
```

When gene identifiers or mixed symbol styles are a concern, standardize at
initialization or call `sn_standardize_gene_symbols()` before downstream
analysis. If `HGNChelper` cannot provide an unambiguous replacement for an
otherwise valid symbol, Shennong preserves the original symbol instead of
returning an `NA` feature name.

```{r standardize-symbols, eval = run_vignette}
symbols_before <- rownames(pbmc)
pbmc_standardized <- sn_standardize_gene_symbols(
  pbmc,
  species = "human",
  is_gene_id = FALSE
)

data.frame(
  feature_set = c("input", "standardized"),
  features = c(length(symbols_before), nrow(pbmc_standardized)),
  duplicated_symbols_resolved = c(NA_integer_, length(symbols_before) - nrow(pbmc_standardized))
)
head(rownames(pbmc_standardized))
```

## Filter cells with inspectable QC flags

`sn_filter_cells()` uses median absolute deviation thresholds by default. The
key point is that filtering and flagging are separated: set `filter = FALSE` to
see which cells would be removed, then set `filter = TRUE` when the rule is
acceptable.

```{r flag-cells}
pbmc_flagged <- sn_filter_cells(
  x = pbmc,
  features = c("nFeature_RNA", "nCount_RNA", "pctMT"),
  group_by = "real_sample",
  method = "mad",
  n = c(5, 5, 3),
  plot = FALSE,
  filter = FALSE
)

qc_flag_cols <- grep("_qc$", colnames(pbmc_flagged[[]]), value = TRUE)
qc_flag_summary <- do.call(rbind, lapply(qc_flag_cols, function(column) {
  data.frame(metric = column, status = names(table(pbmc_flagged[[column]][, 1])),
             cells = as.integer(table(pbmc_flagged[[column]][, 1])))
}))
qc_flag_summary

sn_plot_qc_thresholds(
  pbmc_flagged,
  features = c("nFeature_RNA", "nCount_RNA", "pctMT"),
  sample_by = "real_batch"
)
```

Once the flags look sensible, apply the same rule:

```{r filter-cells}
pbmc_cells <- sn_filter_cells(
  x = pbmc,
  features = c("nFeature_RNA", "nCount_RNA", "pctMT"),
  group_by = "real_sample",
  method = "mad",
  n = c(5, 5, 3),
  plot = FALSE,
  filter = TRUE
)

c(before = ncol(pbmc), after = ncol(pbmc_cells))
```

## Filter genes by expression and annotation

`sn_filter_genes()` keeps the threshold and optional annotation rule in one
place. The expression threshold protects sparse downstream steps, while
`gene_class = "coding"` can remove noncoding genes when your workflow is
protein-coding focused.

```{r filter-genes}
pbmc_genes <- sn_filter_genes(
  x = pbmc_cells,
  min_cells = 5,
  plot = FALSE,
  filter = TRUE,
  species = "human",
  gene_class = "coding"
)

c(before = nrow(pbmc_cells), after = nrow(pbmc_genes))
```

## Normalize and add cell-cycle scores

`sn_normalize_data()` and `sn_score_cell_cycle()` keep preprocessing inside the
same Shennong vocabulary used later by clustering and plotting.

```{r normalize-cycle}
pbmc_norm <- sn_normalize_data(
  object = pbmc_genes,
  normalization_method = "seurat",
  verbose = FALSE
)

pbmc_norm <- sn_score_cell_cycle(pbmc_norm, species = "human")

table(pbmc_norm$Phase)
```

The fixture was deterministically prepared from author-demultiplexed singlets.
The following core plot uses that real HTO classification directly; it verifies
the input selection rule rather than pretending that `sn_find_doublets()` was
run. A lightweight embedding is computed only to display those source calls.

```{r plot-author-doublet-calls, eval = run_vignette}
pbmc_qc_embedding <- sn_run_cluster(
  pbmc_norm,
  normalization_method = "seurat",
  nfeatures = 1000,
  dims = 1:15,
  resolution = 0.6,
  cluster_algorithm = "louvain",
  species = "human",
  verbose = FALSE
)

table(pbmc_qc_embedding$hto_classification_global)
sn_plot_doublets(
  pbmc_qc_embedding,
  class_col = "hto_classification_global",
  reduction = "umap"
)
```

## Optional doublet and ambient-RNA correction

Doublet detection and ambient correction depend on optional backends. Shennong
keeps them as explicit steps rather than hiding them in clustering. These
steps run only when their R backend is installed. The doublet re-estimation is
reserved for the `all` profile because the fixture already contains the
authors' singlet calls; the core profile runs DecontX and its before/after plot
when `celda` is available.

```{r doublets, eval = run_extended && requireNamespace("scDblFinder", quietly = TRUE)}
pbmc_doublets <- sn_find_doublets(pbmc_norm)
table(pbmc_doublets$doublet_class)

pbmc_doublets <- sn_run_cluster(
  pbmc_doublets,
  normalization_method = "seurat",
  nfeatures = 1000,
  dims = 1:15,
  resolution = 0.6,
  cluster_algorithm = "louvain",
  species = "human",
  verbose = FALSE
)
sn_plot_doublets(pbmc_doublets, reduction = "umap")
```

For ambient RNA, materialize a matching raw-count resource with
`ShennongData` when the selected backend requires it. The decontX path below
uses the filtered object directly and writes corrected counts to a new layer,
so the original counts remain inspectable.

```{r ambient, eval = run_vignette && requireNamespace("celda", quietly = TRUE)}
pbmc_decont <- sn_remove_ambient_contamination(
  x = pbmc_norm,
  method = "decontx",
  layer = "decontaminated_counts",
  verbose = FALSE
)

SeuratObject::Layers(pbmc_decont[["RNA"]])
sn_plot_ambient_correction(
  pbmc_decont,
  assay = "RNA",
  layer_before = "counts",
  layer_after = "decontaminated_counts"
)
```

## Store a QC assessment

`sn_assess_qc()` summarizes the object after filtering and can compare it to
the pre-filter object. Store the result when you want later reports to recover
the QC decision without rerunning preprocessing.

```{r assess-qc}
qc_report <- sn_assess_qc(
  object = pbmc_norm,
  reference = pbmc,
  sample_by = "real_sample",
  verbose = FALSE
)

qc_report$overall
head(qc_report$by_sample)

sn_plot_qc(qc_report, metric = "retention_fraction")

pbmc_norm <- sn_assess_qc(
  object = pbmc_norm,
  reference = pbmc,
  sample_by = "real_sample",
  store_name = "kotliarov_preprocessing",
  return_object = TRUE,
  verbose = FALSE
)

names(pbmc_norm@misc$qc_assessments)
```

The object is now ready for clustering. The important part is not just that
cells and genes were filtered, but that the rationale and summary are stored in
a place downstream code can find.
