---
title: "Preprocessing and QC workflow"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocessing and QC workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") ||
  identical(Sys.getenv("IN_PKGDOWN"), "true")

required_pkgs <- c("Seurat", "SeuratObject")
run_vignette <- run_vignette && all(vapply(
  required_pkgs,
  requireNamespace,
  quietly = TRUE,
  FUN.VALUE = logical(1)
))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  eval = run_vignette
)
```

This article covers the first stage of a typical Shennong analysis:

- constructing a Seurat object
- inferring species and standardizing feature naming
- filtering poor-quality cells
- filtering genes by expression and bundled GENCODE annotations
- normalizing data
- adding cell-cycle scores for downstream modeling

The examples use the built-in `pbmc_small` and `pbmc_small_raw` objects so the
workflow stays check-safe and does not depend on network downloads.

```{r setup}
library(Shennong)
library(dplyr)
library(knitr)
library(Seurat)

if (!exists("pbmc_small", inherits = FALSE)) {
  try(data("pbmc_small", package = "Shennong", envir = environment()), silent = TRUE)
}
if (!exists("pbmc_small", inherits = FALSE) && file.exists(file.path("data", "pbmc_small.rda"))) {
  load(file.path("data", "pbmc_small.rda"))
}

if (!exists("pbmc_small_raw", inherits = FALSE)) {
  try(data("pbmc_small_raw", package = "Shennong", envir = environment()), silent = TRUE)
}
if (!exists("pbmc_small_raw", inherits = FALSE) && file.exists(file.path("data", "pbmc_small_raw.rda"))) {
  load(file.path("data", "pbmc_small_raw.rda"))
}
```

```{r prepare-preprocessing-demo, include = FALSE}
counts <- SeuratObject::LayerData(pbmc_small, layer = "counts")
meta_cols <- intersect(
  c("sample", "source_dataset", "source_barcode"),
  colnames(pbmc_small@meta.data)
)
meta_tbl <- pbmc_small@meta.data[, meta_cols, drop = FALSE]

pbmc_init <- sn_initialize_seurat_object(
  x = counts,
  project = "pbmc_small_preprocessing"
)
pbmc_init <- Seurat::AddMetaData(pbmc_init, metadata = meta_tbl)

# For multiple 10x samples on disk, one-shot import is also supported:
# tenx_paths <- sn_list_10x_paths("data/")
# pbmc_list <- sn_initialize_seurat_object(x = tenx_paths, species = "human")

pbmc_cells <- suppressWarnings(
  sn_filter_cells(
    pbmc_init,
    features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
    plot = FALSE
  )
)

pbmc_genes <- suppressWarnings(
  sn_filter_genes(
    pbmc_cells,
    min_cells = 3,
    gene_class = "coding",
    plot = FALSE
  )
)

pbmc_norm <- sn_normalize_data(
  pbmc_genes,
  normalization_method = "seurat",
  verbose = FALSE
)

pbmc_cycle <- sn_score_cell_cycle(pbmc_norm, species = "human")

qc_summary <- tibble::tibble(
  stage = c("initialized", "cell_filtered", "gene_filtered", "normalized"),
  cells = c(ncol(pbmc_init), ncol(pbmc_cells), ncol(pbmc_genes), ncol(pbmc_norm)),
  genes = c(nrow(pbmc_init), nrow(pbmc_cells), nrow(pbmc_genes), nrow(pbmc_norm))
)

feature_summary <- tibble::tibble(
  metric = c("species", "median_nFeature", "median_nCount", "median_percent_mt"),
  value = c(
    sn_get_species(pbmc_init),
    round(stats::median(pbmc_cells$nFeature_RNA), 1),
    round(stats::median(pbmc_cells$nCount_RNA), 1),
    round(stats::median(pbmc_cells$percent.mt), 2)
  )
)

cycle_summary <- pbmc_cycle@meta.data |>
  dplyr::count(Phase, name = "cells")
```

## 1. Initialize a Seurat object

Start from a matrix-like object and let Shennong create the Seurat object with
species-aware QC metadata.

```{r}
knitr::kable(feature_summary)
```

## 2. Filter poor-quality cells

`sn_filter_cells()` works on standard QC columns such as `nFeature_RNA`,
`nCount_RNA`, and `percent.mt`. The function can also draw diagnostics, but the
article disables plotting to keep the rendered workflow compact.

```{r}
knitr::kable(qc_summary)
```

## 3. Filter genes with expression and annotation rules

Shennong ships bundled human and mouse GENCODE snapshots. That means
`sn_filter_genes()` can combine expression thresholds with annotation-aware
filters such as `gene_class = "coding"` or exact `gene_type` values.

```r
pbmc_genes <- sn_filter_genes(
  pbmc_cells,
  min_cells = 3,
  gene_class = "coding",
  plot = FALSE
)
```

For finer control you can target exact biotypes:

```r
pbmc_lnc_only <- sn_filter_genes(
  pbmc_cells,
  min_cells = 3,
  gene_type = c("lncRNA", "antisense"),
  plot = FALSE
)
```

## 4. Normalize and add cell-cycle scores

After QC, normalize the object and add cell-cycle scores if they are useful for
regression, cluster interpretation, or doublet diagnostics.

```{r}
knitr::kable(cycle_summary)
```

## 5. Optional QC extensions

Shennong also exposes heavier QC helpers that are usually run only when the
analysis requires them:

```r
# Doublet detection
pbmc_doublets <- sn_find_doublets(pbmc_norm)

# Ambient RNA correction from a raw count matrix
pbmc_decont <- sn_remove_ambient_contamination(
  pbmc_norm,
  raw_counts = pbmc_small_raw
)

# Summarize current QC status or compare a filtered object against a pre-QC reference
qc_report <- sn_assess_qc(pbmc_decont, reference = pbmc_norm)
qc_report$overall
```

## Where this stage leads next

Once preprocessing is complete, the usual next step is `sn_run_cluster()` for
single-sample clustering or Harmony-based integration. The dedicated clustering
article continues from this point.
