---
title: "Program Discovery and Gene Regulatory Networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Program Discovery and Gene Regulatory Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Shennong separates latent program discovery from gene-set scoring and from
regulatory-network inference. All three workflows store versioned results in
the Seurat object, but their evidence is not interchangeable.

## Discover latent programs

The local NMF backend uses several random restarts, retains the best
reconstruction, and stores restart and cross-run stability diagnostics.

```{r discover}
object <- sn_discover_programs(
  object,
  method = "nmf",
  n_programs = 6,
  name = "tumor_programs",
  backend_control = list(nrun = 10, seed = 717)
)

programs <- sn_get_result(object, "program_discovery", "tumor_programs")
head(programs$tables$gene_weights)
programs$tables$fit_diagnostics
programs$tables$stability

sn_plot_discovered_programs(object, "tumor_programs", type = "weights")
sn_plot_discovered_programs(object, "tumor_programs", type = "activity")
```

Use `group_by` only when each stratum can support an independent
factorization. cNMF and Hotspot require an explicit runner or imported result;
their external environment is never installed silently.

## Infer and inspect regulons

GENIE3 is the directly runnable R backend. Supplying a curated regulator list
usually produces a more interpretable network and reduces compute cost.

```{r genie3}
object <- sn_run_grn(
  object,
  method = "genie3",
  regulators = transcription_factors,
  group_by = "cell_type",
  name = "tumor_grn"
)

grn <- sn_get_result(object, "grn", "tumor_grn")
head(grn$tables$edges)
head(grn$tables$regulons)
head(grn$tables$activity)
head(grn$tables$specificity)

sn_plot_regulon(object, "tumor_grn", type = "network")
sn_plot_regulon(object, "tumor_grn", type = "specificity")
```

For pySCENIC, run the motif-pruning workflow in its audited Python environment
and pass the edge/AUCell tables through `backend_control$runner` or
`backend_control$result`. The legacy R SCENIC adapter exists for prior result
exports; it is not presented as the default runtime.

The default derived activity is the mean expression of each regulon's retained
targets. `backend_control = list(activity_method = "ucell")` selects rank-based
activity. When `group_by` is supplied, Shennong reports an eta-squared effect
size alongside group means; it deliberately does not label this statistic as
SCENIC's regulon-specificity score.
