---
title: "Program Discovery and Gene Regulatory Networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Program Discovery and Gene Regulatory Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
run_vignette <- identical(tolower(Sys.getenv("SHENNONG_RUN_VIGNETTES")), "true")
real_profile <- tolower(Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core"))
root_candidates <- unique(c(
  Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = ""),
  file.path(getwd(), "data-local", "pkgdown-real"),
  file.path(getwd(), "..", "data-local", "pkgdown-real")
))
root_candidates <- root_candidates[nzchar(root_candidates)]
existing_roots <- root_candidates[dir.exists(root_candidates)]
real_root <- if (length(existing_roots)) existing_roots[[1]] else root_candidates[[1]]
melanoma_path <- file.path(real_root, "cancer", "melanoma_gse72056.qs2")
core_ready <- run_vignette && file.exists(melanoma_path)
genie3_ready <- core_ready && identical(real_profile, "all") &&
  requireNamespace("GENIE3", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
  fig.width = 7, fig.height = 5
)
```

Program discovery and GRN inference answer different questions. This article
uses author-normalized expression from real malignant cells in GSE72056. The
local NMF implementation is a core workflow; GENIE3 is an optional R backend
and never receives a fabricated edge table when it is absent.

```{r runtime-status}
knitr::kable(data.frame(
  workflow = c("NMF program discovery", "GENIE3 GRN", "pySCENIC / GRNBoost2"),
  mode = c("core", "extended", "external import or runner"),
  status = c(
    if (core_ready) "executed below" else if (!run_vignette) "disabled: set SHENNONG_RUN_VIGNETTES=true" else "fixture missing",
    if (genie3_ready) "executed below" else if (!identical(real_profile, "all")) "disabled: requires SHENNONG_REAL_PROFILE=all" else if (!requireNamespace("GENIE3", quietly = TRUE)) "dependency missing: GENIE3" else "fixture missing",
    "not run: no audited external result supplied"
  ),
  check.names = FALSE
))
```

## Discover latent programs in real malignant cells

The fixture retains the study's inferred-CNV malignancy class. We select the
300 most variable genes among malignant cells and run four NMF programs with
three seeded restarts. Every weight and activity value printed below is fitted
during the article build.

```{r prepare-programs, eval=core_ready}
library(Shennong)

melanoma <- qs2::qs_read(melanoma_path)
malignant <- subset(
  melanoma,
  cells = colnames(melanoma)[melanoma$malignant_call == "malignant"]
)
expression <- SeuratObject::LayerData(
  malignant, assay = "RNA", layer = "data"
)
gene_variance <- Matrix::rowMeans(expression ^ 2) -
  Matrix::rowMeans(expression) ^ 2
program_features <- names(head(sort(gene_variance, decreasing = TRUE), 300))

data.frame(
  tumors = length(unique(malignant$tumor)),
  malignant_cells = ncol(malignant),
  measured_genes = nrow(malignant),
  fitted_genes = length(program_features)
)
```

```{r discover-programs, eval=core_ready}
programs <- sn_discover_programs(
  malignant,
  method = "nmf",
  n_programs = 4,
  name = "melanoma_programs",
  assay = "RNA",
  layer = "data",
  features = program_features,
  backend_control = list(
    nrun = 3L,
    max_iter = 120L,
    top_genes = 25L,
    seed = 717L
  ),
  return_object = FALSE
)

head(programs$tables$gene_weights, 20)
programs$tables$fit_diagnostics
programs$tables$stability
sn_plot_discovered_programs(programs, type = "weights", n = 10)
sn_plot_discovered_programs(programs, type = "activity")
sn_plot_discovered_programs(programs, type = "stability")
```

## Extended: infer a GENIE3 network

When the `all` profile and GENIE3 are available, the backend uses the same
real malignant-cell matrix. The selected regulators must occur among the 300
variable input genes. A backend failure is rendered as a failed status and is
not replaced by imported or simulated edges.

```{r genie3, eval=genie3_ready}
candidate_regulators <- c(
  "MITF", "SOX10", "STAT1", "IRF1", "JUN", "FOS", "CEBPB", "ETS1"
)
regulators <- intersect(candidate_regulators, program_features)

genie3_attempt <- tryCatch(
  sn_run_grn(
    malignant,
    method = "genie3",
    regulators = regulators,
    group_by = "tumor",
    assay = "RNA",
    layer = "data",
    name = "melanoma_genie3",
    backend_control = list(
      n_features = 300L,
      n_trees = 200L,
      n_cores = 2L,
      max_edges = 1000L,
      top_targets = 30L,
      seed = 717L
    ),
    return_object = FALSE
  ),
  error = identity
)

if (inherits(genie3_attempt, "error")) {
  knitr::kable(data.frame(
    backend = "GENIE3", status = "failed",
    detail = conditionMessage(genie3_attempt), check.names = FALSE
  ))
} else {
  grn <- genie3_attempt
  print(head(grn$tables$edges, 20))
  print(head(grn$tables$regulons, 20))
  print(head(grn$tables$specificity, 20))
  print(sn_plot_regulon(grn, type = "network", n = 30))
  print(sn_plot_regulon(grn, type = "specificity"))
}
```

pySCENIC and GRNBoost2 require an explicit audited runner or imported result.
The documentation build does not silently install their Python environments or
present NMF output as a GRN result.
