---
title: "Publication figure specifications and export bundles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Publication figure specifications and export bundles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
real_data_root <- Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = "")
pbmc_fixture <- if (nzchar(real_data_root)) {
  file.path(real_data_root, "single-cell", "kotliarov_pbmc.qs2")
} else {
  ""
}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") &&
  nzchar(pbmc_fixture) && file.exists(pbmc_fixture)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Shennong figures remain native `ggplot`, patchwork, or ComplexHeatmap objects.
The publication engine attaches a lightweight specification instead of
introducing a new plotting class, so ordinary `+ theme(...)` composition keeps
working.

## Profiles and automatic sizing

Generic profiles describe screen, column, page, and slide constraints. They do
not claim to replace current journal artwork instructions; verify those before
submission.

```{r profiles}
library(Shennong)
library(Seurat)

sn_list_figure_profiles()

pbmc <- qs2::qs_read(pbmc_fixture)
pbmc <- sn_run_cluster(
  pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

plot <- sn_plot_dim(
  pbmc,
  reduction = "umap",
  group_by = "real_response",
  split_by = "real_batch",
  title = "Kotliarov PBMC response groups"
)

sn_recommend_figure_size(plot, profile = "double_column")
plot
```

The example above is rendered from 2,000 real CITE-seq cells and retains the
study's responder and acquisition-batch labels. Point size follows a clamped
power rule, and high-density layers are recommended for rasterization while
text and geometry remain vector in PDF/SVG output.

## Inspect and apply a specification

Core Shennong plots attach `shennong_figure_spec` automatically. Any native
plot can also be profiled after construction.

```{r apply}
plot <- sn_apply_figure_profile(plot, "double_column")
sn_figure_spec(plot)$recommended
sn_validate_figure(plot)
plot
```

Validation reports calculated dimensions, text and raster settings, small
panels, category/legend overload, long labels, pagination, extreme aspect
ratios, and dense networks. It predicts likely problems; final visual review is
still required.

## Deterministic export

`sn_save_figure()` supports PDF, SVG, TIFF, and PNG without depending on the
current RStudio graphics device. Explicit dimensions override automatic values.

```{r save}
figure_dir <- file.path(tempdir(), "shennong-real-figures")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)

png_path <- sn_save_figure(
  plot,
  file.path(figure_dir, "kotliarov_response.png"),
  profile = "double_column"
)

pdf_path <- sn_export_figure(
  plot,
  file.path(figure_dir, "kotliarov_response.pdf"),
  profile = "double_column"
)

file.info(c(png_path, pdf_path))[, c("size", "mtime")]
```

For a reproducible handoff, export the rendering, source table, calculated
specification, session information, checksums, and validation result together.

```{r bundle}
manifest <- sn_export_figure_bundle(
  plot,
  path = file.path(figure_dir, "kotliarov_response_bundle"),
  formats = c("pdf", "png"),
  include_data = TRUE,
  include_spec = TRUE,
  include_session = TRUE,
  profile = "single_column"
)
manifest$files
manifest$validation$status
```

Result-aware `sn_plot_de()`, `sn_plot_enrichment()`, and `sn_plot_gsea()` attach
their standardized plotting tables as bundle source data. The QC, clustering,
integration, annotation projection, bulk, dimensional, feature, dot, heatmap,
violin, box, bar, composition, and Milo helpers also expose figure metadata for
automatic sizing and QA.
