---
title: "From samples to evidence: a research workflow map"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From samples to evidence: a research workflow map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  eval = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

::: {.sn-hero-panel}
[Start with the question]{.sn-kicker}

Shennong's functions form an evidence chain rather than a menu of unrelated
methods. Start with the biological unit of replication, preserve the source
assay and layer, compare plausible methods, and finish with a stored table and
an inspectable figure.

<div class="sn-flow" role="list" aria-label="Shennong research evidence chain">
<span role="listitem">observations</span><b aria-hidden="true">→</b>
<span role="listitem">quality control</span><b aria-hidden="true">→</b>
<span role="listitem">representation</span><b aria-hidden="true">→</b>
<span role="listitem">biological comparison</span><b aria-hidden="true">→</b>
<span role="listitem">stored evidence</span><b aria-hidden="true">→</b>
<span role="listitem">figure + provenance</span>
</div>

Use this page to choose a route. The linked articles contain the executable
real-data analysis and its visual checkpoints.
:::

## What the examples do—and do not—claim

The website uses locally materialized subsets of public observations. They are
not bundled in the package and the articles never download them during
`R CMD check`. A normal build shows the code and input contract. A designated
real-data build executes an article only when its fixture and dependencies are
present:

```sh
SHENNONG_RUN_VIGNETTES=true \
SHENNONG_REAL_DATA_DIR=/path/to/pkgdown-real \
Rscript -e 'pkgdown::build_site()'
```

<span class="sn-tier">core</span> uses a bounded local fixture and the default
workflow. <span class="sn-tier">extended</span> additionally requires
`SHENNONG_REAL_PROFILE=all` and optional dependencies. External services,
credentials, model files, or imported backend results remain explicitly
opt-in. A displayed plot is therefore evidence from an executed chunk, not a
placeholder manufactured when a backend is absent.

## Choose a scientific narrative

::: {.sn-module-grid}

::: {.sn-path-card}
### Longitudinal immune cohort

Use the real Kotliarov PBMC CITE-seq subset to ask how RNA and ADT cell states,
batch, and vaccine response vary across 20 biological samples. Begin with
[preprocessing](preprocessing-qc.html), continue through
[integration](clustering.html), [annotation and pathways](annotation-pathways.html),
then test [composition](composition-analysis.html) or
[differential abundance](abundance-priority.html).
:::

::: {.sn-path-card}
### Tumour ecosystem and clinical translation

Use author-normalized malignant cells from GSE72056 for programs, regulatory
networks, malignancy evidence, and metabolism. Connect those hypotheses to
TCGA-SKCM expression and follow-up without relabelling TPM as counts. Follow
[program discovery](program-discovery-grn.html), [CNV and metabolism](cnv-metabolism.html),
and [bulk clinical analysis](bulk-transcriptomics.html).
:::

::: {.sn-path-card}
### Cell-state dynamics

Use the Hermann spermatogenesis fixture when the question is ordered change,
not only discrete clusters. Separate trajectory topology, gene trends,
velocity, and terminal-state probabilities. Follow
[trajectory and dynamic genes](trajectory-dynamics.html), then the optional
[velocity and fate](velocity-fate.html) path.
:::

::: {.sn-path-card}
### Tissue organization

Use the real Visium lymph-node sections to keep expression and coordinates
together. Inspect native neighborhoods and spatial features before introducing
an extended backend or reference mapping. Follow the
[spatial workflow](spatial-workflows.html).
:::

:::

## The core single-cell evidence chain

<ol class="sn-steps">
<li><strong>Define the observation and replicate.</strong> Materialize the
source once, retain sample metadata, and decide whether the scientific replicate
is a donor, capture, section, or bulk sample.</li>
<li><strong>Audit quality before filtering.</strong> Plot distributions and
thresholds, record flags, and retain the pre-filter evidence needed to explain
cell and feature loss.</li>
<li><strong>Choose a representation.</strong> Cluster without integration as a
baseline, then compare batch-aware methods against both mixing and biological
separation metrics.</li>
<li><strong>Name and test states.</strong> Join marker evidence, reference
evidence, signatures, and ontology terms. Test replicate-aware changes instead
of treating cells as independent samples.</li>
<li><strong>Store the result, then plot it.</strong> Retrieve the exact result
name from the object so tables, figures, prompts, and exported source data all
refer to the same analysis.</li>
<li><strong>Report runtime and provenance.</strong> Record backend versions,
parameter fingerprints, and optional acceleration separately from the
scientific result.</li>
</ol>

### 1. Materialize once and retain sample identity

Dataset discovery and publication belong to ShennongData. Shennong begins with
a materialized matrix, file, or object and preserves the metadata that defines
the study design. See [Data input, output, and project setup](data-io-projects.html).

```{r load-real-pbmc}
library(Shennong)

pbmc <- qs2::qs_read(file.path(
  Sys.getenv("SHENNONG_REAL_DATA_DIR"),
  "single-cell",
  "kotliarov_pbmc.qs2"
))

table(pbmc$real_sample, pbmc$real_response)
```

### 2. Make QC decisions visible

The filtering calls return an analysis object; plotting functions expose the
decision boundary and retained fraction. The complete real-data sequence is in
[Preprocessing and QC](preprocessing-qc.html).

```{r qc-checkpoints}
pbmc <- sn_filter_cells(
  pbmc,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  plot = FALSE
)
pbmc <- sn_filter_genes(pbmc, min_cells = 3, plot = FALSE)

sn_plot_qc_thresholds(
  pbmc,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt")
)
```

### 3. Establish a baseline before integration

`sn_run_cluster()` exposes the normalization and integration choices in one
call. Compare batch mixing and biological conservation; a UMAP alone is not an
integration test. See [Clustering and integration](clustering.html),
[layer-aware workflows](layered-workflows.html), and
[integration metrics](metrics-diagnostics.html).

```{r representation-checkpoints}
pbmc <- sn_run_cluster(
  pbmc,
  normalization_method = "seurat",
  integration_method = "harmony",
  batch = "real_batch",
  resolution = 0.6
)

sn_plot_dim(pbmc, reduction = "umap", group_by = "real_batch")
sn_plot_dim(pbmc, reduction = "umap", group_by = "seurat_clusters", label = TRUE)

assessment <- sn_assess_integration(
  pbmc,
  batch_by = "real_batch",
  cluster_by = "seurat_clusters",
  reduction = "harmony",
  baseline_reduction = "pca"
)
sn_plot_integration(assessment)
```

### 4. Connect states to genes and pathways

Store marker results on the object, then let downstream plots and enrichment
retrieve them by name. ORA and GSEA answer different questions: ORA tests a
selected gene set against an explicit universe, whereas GSEA uses the complete
ranked statistic. See [Markers, signatures, pathways, and annotation](annotation-pathways.html).

```{r annotation-checkpoints}
pbmc <- sn_find_de(
  pbmc,
  analysis = "markers",
  group_by = "seurat_clusters",
  store_name = "cluster_markers",
  return_object = TRUE,
  verbose = FALSE
)

sn_plot_dot(pbmc, features = "top_markers", de_name = "cluster_markers")

pbmc <- sn_run_enrichment(
  pbmc,
  source_de_name = "cluster_markers",
  gene_clusters = gene ~ cluster,
  database = c("GOBP", "H"),
  universe = rownames(pbmc),
  store_name = "cluster_pathways",
  return_object = TRUE
)

pathways <- sn_get_enrichment_result(
  pbmc,
  enrichment_name = "cluster_pathways.GOBP"
)
sn_plot_enrichment(pathways, type = "dot")
```

### 5. Return to the sample as the replicate

Cell counts are descriptive until they are summarized within biological
samples. Compare sample-level proportions or use a differential-abundance
model with an explicit design. See [Composition analysis](composition-analysis.html)
and [Differential abundance and state priority](abundance-priority.html).

```{r sample-comparison-checkpoints}
composition <- sn_calculate_composition(
  pbmc,
  group_by = c("real_sample", "real_response"),
  variable = "seurat_clusters"
)

sn_plot_composition(
  pbmc,
  x = real_response,
  fill = seurat_clusters,
  type = "sample_boxplot",
  sample_by = "real_sample"
)
```

### 6. Extend only when the question requires it

The common object and result contract supports several distinct extensions:

- [cell-cell communication](communication-consensus.html) for sender-receiver
  evidence and cross-method consensus;
- [program discovery and GRNs](program-discovery-grn.html) for latent programs
  and regulatory hypotheses;
- [trajectory](trajectory-dynamics.html) and [velocity/fate](velocity-fate.html)
  for directed or ordered change;
- [spatial analysis](spatial-workflows.html) for coordinate-aware expression;
- [bulk deconvolution](bulk-deconvolution.html) and
  [bulk transcriptomics](bulk-transcriptomics.html) for cohort-level translation.

Each extension has its own admissible input. A normalized tumour matrix is not
silently treated as UMI counts, and an absent external backend never produces
a substitute result.

### 7. Retrieve, validate, and publish the same evidence

Discover a stored result before retrieving it, validate its versioned envelope,
then export the plot together with source data and a figure specification. See
[result and method contracts](analysis-results-and-methods.html),
[publication figures](publication-figures.html), and
[interpretation and reporting](interpretation-workflows.html).

```{r evidence-export}
sn_list_results(pbmc)
markers <- sn_get_result(pbmc, type = "de", name = "cluster_markers")
sn_validate_result(markers)

figure <- sn_plot_de(markers, type = "effect")
sn_export_figure_bundle(
  figure,
  path = "figures/cluster-markers"
)
```

## Module directory

| Research decision | Principal entry points | Visual checkpoint | Detailed article |
|---|---|---|---|
| Materialize and govern inputs | `sn_read()`, `sn_write()`, `sn_initialize_project()` | input and metadata audit tables | [Data and projects](data-io-projects.html) |
| Filter and normalize | `sn_filter_cells()`, `sn_filter_genes()`, `sn_normalize_data()` | `sn_plot_qc_thresholds()`, `sn_plot_qc()` | [Preprocessing and QC](preprocessing-qc.html) |
| Cluster or integrate modalities | `sn_run_cluster()`, `sn_run_multimodal()` | `sn_plot_dim()`, `sn_plot_resolution_sweep()` | [Clustering](clustering.html) |
| Respect assay layers | layer-aware arguments across preprocessing, DE, and scoring | embedding and layer audit | [Layered workflows](layered-workflows.html) |
| Validate integration | `sn_assess_integration()`, `sn_compare_integrations()` | `sn_plot_integration()` | [Metrics and diagnostics](metrics-diagnostics.html) |
| Annotate and test features | `sn_run_annotation()`, `sn_find_de()`, `sn_run_enrichment()` | marker dot plots, ORA dots, GSEA curves | [Annotation and pathways](annotation-pathways.html) |
| Score or discover programs | `sn_score_programs()`, `sn_discover_programs()`, `sn_run_grn()` | activity heatmaps and regulon networks | [Programs and GRNs](program-discovery-grn.html) |
| Compare sample composition | `sn_calculate_composition()`, `sn_test_abundance()` | composition bars and abundance effects | [Composition](composition-analysis.html), [abundance](abundance-priority.html) |
| Model communication | `sn_run_cell_communication()` | bubble, heatmap, network, ligand-target plot | [Communication](communication-consensus.html) |
| Follow dynamic states | `sn_run_trajectory()`, `sn_run_velocity()`, `sn_run_fate()` | topology, trends, velocity, fate | [Trajectory](trajectory-dynamics.html), [velocity](velocity-fate.html) |
| Explain tumour states | `sn_run_cnv()`, `sn_run_metabolism()` | CNV heatmap and pathway activity | [CNV and metabolism](cnv-metabolism.html) |
| Keep tissue coordinates | `sn_run_spatial()` and spatial sub-workflows | features, domains, neighborhoods, mapping | [Spatial workflows](spatial-workflows.html) |
| Translate to cohorts | `sn_run_bulk()`, `sn_find_de()`, `sn_run_survival()`, `sn_run_bulk_deconvolution()` | QC, PCA, DE, pathways, survival, fractions | [Bulk](bulk-transcriptomics.html), [deconvolution](bulk-deconvolution.html) |
| Build publication evidence | result retrieval, figure specification, interpretation helpers | exported figure and source-data bundle | [Visualization](visualization.html), [publication](publication-figures.html), [interpretation](interpretation-workflows.html) |
| Profile the workflow | usage tracking and acceleration controls | timing summaries | [Runtime observability](runtime-observability.html) |

## Read acceleration evidence as four separate gates

An active patch is not proof that a Shennong call used a fast kernel. The full
sequence is:

::: {.sn-gate-grid}
::: {.sn-gate data-state="accepted"}
### 1. Activation
The replacement is installed in a namespace for a bounded scope. This reports
availability, not use.
:::
::: {.sn-gate data-state="accepted"}
### 2. Target intersection
The workflow must actually call the replaced upstream symbol. A loaded patch
for an unused backend has no effect.
:::
::: {.sn-gate data-state="blocked"}
### 3. Guard accepted
The real call shape, input class, operation, version, and parameters must match
the validated fast-path guard. Otherwise the replacement must fall back.
:::
::: {.sn-gate data-state="accepted"}
### 4. Fast-path hit
Runtime evidence must show that the guarded kernel executed. Shennong does not
infer this from activation alone.
:::
:::

This distinction prevents scientifically unsafe false hits. The pinned
JoinLayers and Coralysis replacements can be active while Shennong's real call
shape fails their benchmark-specific guards and falls back. An older broad
Seurat `FindIntegrationAnchors` replacement lacked a reduction guard, so an
RPCA request could incorrectly enter a CCA-oriented path. The tradeSeq patch
globally replaced `mgcv` behavior without an exact dependency guard. Shennong
therefore uses strict source/version checks, operation-specific scopes, and
temporary `with_disabled` boundaries around incompatible calls. Those patches
remain explicit-only until a three-arm contract proves the actual Shennong
workflow. See [Analysis results and methods](analysis-results-and-methods.html) for the
measured intersections and quarantine rationale.

## Find the next article from an output

If the output is a table, call `sn_list_results()` and retrieve its exact stored
name before interpretation. If it is a figure, keep the source table beside the
export. If it is only an embedding, continue to diagnostics or biological
testing before treating it as evidence. Runtime statistics belong in a separate
opt-in database and never replace the result stored on the analysis object.
