---
title: "Runtime observability: local outbox and managed DBI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Runtime observability: local outbox and managed DBI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Shennong can record which public APIs are used, their elapsed time,
their sanitized parameter configuration, and whether the call ran in an
explicit development, production, test, or benchmark context. This facility is
opt-in. Loading Shennong never creates a database, starts a tracker, connects
to a remote destination, or sends usage data over the network.

The intended feedback loop is:

```text
tracked public export
        |
        +--> SQLite start row (call number + sanitized parameter hash)
        |
        +--> unchanged scientific function
        |
        +--> SQLite completion (elapsed time + status + warnings)
                         |
                         +--> sn_list_usage_runs()
                         +--> sn_summarize_usage()
```

## Enable tracking explicitly

Choose a local SQLite file and state the execution mode. `"production"` means
a real analytical environment only when the caller explicitly selects it;
Shennong does not infer production from `interactive()`, CI variables, or the
installation path.

```{r enable}
library(Shennong)

usage_database <- file.path(
  "analysis-private",
  "observability",
  "shennong-usage.sqlite"
)

sn_enable_usage_tracking(
  path = usage_database,
  mode = "development",
  display = TRUE,
  nested = TRUE
)

object <- sn_run_cluster(
  object,
  batch = "sample",
  normalization_method = "seurat",
  integration_method = "harmony",
  nfeatures = 3000,
  npcs = 30,
  resolution = 0.6,
  verbose = FALSE
)

sn_disable_usage_tracking()
```

With `display = TRUE`, completion prints a compact line such as:

```text
[Shennong timing] sn_run_cluster #12 (development): 38.417 s [ok]
```

The function result, visibility, warnings, errors, and R random-number stream
are preserved. Tracking uses SQLite-generated run identifiers, so it does not
consume the analytical RNG. Database failures warn once and fail open by
default; use `strict = TRUE` only when a development or test run must assert
that every audit write succeeded.

For a temporary scope, use `sn_with_usage_tracking()`:

```{r scoped}
result <- sn_with_usage_tracking(
  {
    sn_enrich(
      genes,
      analysis = "ora",
      database = "GOBP",
      pvalue_cutoff = 0.05
    )
  },
  path = usage_database,
  mode = "development",
  display = FALSE
)
```

## Consent-gated remote research

Managed production deployments can use a remote DBI destination without making
scientific calls depend on the network. Every call first commits to a local
SQLite outbox; `sn_flush_usage_tracking()` later sends only completed,
sanitized rows. A remote connection is impossible until explicit
`remote_research` consent is supplied.

```{r remote}
consent <- sn_confirm_usage_consent(
  scope = "remote_research",
  policy_id = "my-usage-study",
  policy_version = "1.0",
  purposes = c("usage_research", "performance_research"),
  data_categories = c("api_name", "timing", "safe_parameters"),
  study_id = "public-beta"
)

store <- sn_create_usage_store(
  backend = "dbi",
  path = "runs/shennong-usage-outbox.sqlite",
  connect = function() {
    # Prefer a libpq service, .pgpass, IAM token, or secret manager.
    DBI::dbConnect(
      RPostgres::Postgres(),
      service = "shennong_usage_writer"
    )
  },
  table_prefix = "shennong_usage",
  allow_remote_ddl = FALSE
)

sn_enable_usage_tracking(
  store = store,
  consent = consent,
  mode = "production",
  remote_flush = "manual"
)

# Run ordinary Shennong functions, close the local session, then deliver.
sn_disable_usage_tracking()
sn_flush_usage_tracking(store, consent)

# A reader-authorized factory can query the remote table through the same API.
sn_summarize_usage(
  store = store,
  source = "remote",
  sort_by = "total_seconds"
)
```

`allow_remote_ddl = TRUE` is intended only for administrator-controlled table
initialization and creates uniqueness indexes for session/run IDs. When
`allow_remote_ddl = FALSE`, administrator migrations must provide equivalent
unique constraints. Each flush selects only local sessions carrying the exact,
tamper-checked consent receipt; other consent cohorts remain pending. The
receipt's `data_categories` are field-level gates, so API-only consent omits
method/backend, parameters, timing, package versions, errors and acceleration.
Distributed users should not receive direct PostgreSQL
credentials; a public large-scale study should place an authenticated HTTPS
ingestion service in front of the research database. DBI mode is for managed
environments. Shennong generates no participant or installation identifier;
`study_id` is only a cohort label.

## Inspect calls and optimization priorities

`sn_list_usage_runs()` exposes both the workflow-wide invocation number and the
invocation number for the same sanitized parameter fingerprint. This makes it
possible to distinguish “the twelfth `sn_run_cluster()` call” from “the third
call using this integration/normalization configuration.”

```{r list}
cluster_runs <- sn_list_usage_runs(
  usage_database,
  workflow = "sn_run_cluster",
  mode = "development"
)

cluster_runs[, c(
  "workflow", "method", "status", "elapsed_seconds",
  "invocation_number", "parameter_set_invocation", "params_json"
)]
```

Use `by_parameters = TRUE` to compare configurations and `sort_by` to choose
the optimization question:

```{r summarize}
# Most commonly used workflows
sn_summarize_usage(
  usage_database,
  mode = "production",
  sort_by = "calls"
)

# Workflows consuming the most cumulative wall time
sn_summarize_usage(
  usage_database,
  mode = "production",
  sort_by = "total_seconds"
)

# Separate sn_run_cluster parameter fingerprints
sn_summarize_usage(
  usage_database,
  workflow = "sn_run_cluster",
  mode = "production",
  by_parameters = TRUE,
  sort_by = "median_seconds"
)
```

Summaries count only top-level calls by default. Nested calls are still
available in `sn_list_usage_runs(top_level = FALSE)`, with `parent_run_id`,
`root_run_id`, and `depth`, but excluding them from the default summary avoids
double-counting inclusive elapsed time.

## What is instrumented

The process-local registry covers 257 of the current 267 function exports,
including analysis, plot, get/list, store, IO, validation, installation,
project, rio-adapter, and low-level backend calls. The remaining ten exports
are the usage-control APIs themselves and are excluded to prevent recursive
writes. Use `functions =` to select a smaller exact set when a study does not
need the complete API surface.

The registry is installed only by `sn_enable_usage_tracking()` and the original
function bindings are restored by `sn_disable_usage_tracking()`. Workers do not
implicitly inherit tracking: PSOCK workers start clean, and forked children
detect the parent PID and do not write into the parent's session.

This development implementation replaces namespace bindings when tracking is
enabled. A function reference saved before enablement, or imported into another
package earlier, bypasses that replacement. `sn_check_usage_tracking()` reports
`covers_preexisting_function_references = FALSE`; enable tracking at process
startup before caching function references. A future permanent thin-wrapper
build is required before claiming interception of pre-existing references.

## Static coverage is not executed evidence

`inst/conformance/public-api-parameters.json` inventories every exported
`sn_*` function, its formals, selector values, and static runtime
classification. The clustering matrix separately enumerates normalization and
integration selectors, pairwise axes, and high-risk cases; the enrichment
matrix inventories ORA/GSEA across GO, KEGG, MSigDB and grouped/ungrouped
dispatch. Their tests detect API drift and missing classifications; they do not
execute every listed case and do not establish wrapper-to-upstream parity. A usage row proves that the
namespace wrapper intercepted a call, while a backend conformance pass requires
the independently executed oracle and comparators defined in the corresponding
machine-readable contract.

## Privacy boundary

The recorder uses a conservative parameter policy. Short literal method names,
numeric controls, seeds, booleans, dimensions, and thresholds can be retained.
Allowlisted selector variables can be resolved only to small atomic values;
arbitrary call expressions are deliberately not evaluated.
The following are never stored as values:

- Seurat, SingleCellExperiment, matrix, data-frame, or model content;
- genes, signatures, cells, samples, patients, subjects, or long vectors;
- paths, URLs, endpoints, usernames, email addresses, or environment values;
- credentials, tokens, cookies, passwords, prompts, messages, responses, or
  other free text.

Errors retain the condition class and a short redacted message. `sn_run_llm()`
has a stricter policy: only provider, model, and a scalar temperature may be
recorded; prompts, messages, tool payloads, and responses are excluded. The
database contains no installation identifier and no telemetry upload endpoint.

Each local session records R, Shennong, and installed material backend versions once,
so timing comparisons can be separated when the execution environment changes.
Keep the SQLite file under the same access controls as other local analysis
logs, and use a local disk rather than NFS for WAL-mode concurrency.
Remote rows omit PID and error text. Full package versions, safe parameter JSON,
error class, and acceleration evidence are separate consent data categories.

## AutoZyme interpretation

When a Shennong AutoZyme scope activates during a tracked workflow, the run row
records the patch names with evidence `"scope_activation_only"`. This is
deliberately not a `fast_path_hit` claim: a patch can be active while its input
guard falls back to the upstream implementation. Use the three-arm benchmark
runner described in the AutoZyme article to establish output parity and an
empirical speed difference for a particular fixture and package build.

Automatic activation is deliberately narrower than the patch catalog. The
current automatic set is `cellchat`, `clusterprofiler` (annotation cache only),
`lisi`, `nichenetr`, `scdblfinder`, `seurat` (NormalizeData only),
`seurat_merge`, `soupx`, and `ucell`. Coralysis, standalone decontX, broad
Seurat targets, JoinLayers, tradeSeq, and WGCNA remain explicit-only until
their Shennong call shapes and guards pass the required contracts. A benchmark
or an eligible manual patch does not add it to the automatic set.

## Time any additional expression

`sn_time_call()` is a small utility for an expression outside the automatic
workflow registry. It returns the expression value unchanged and can optionally
join the active local usage session.

```{r time-call}
markers <- sn_time_call(
  Seurat::FindAllMarkers(object),
  label = "exploratory-markers",
  display = TRUE,
  record = TRUE
)
```
