---
title: "Spatial Feature, Domain, and Neighborhood Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Feature, Domain, and Neighborhood Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
real_data_root <- Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = "")
spatial_fixture <- if (nzchar(real_data_root)) {
  file.path(real_data_root, "spatial", "visium_lymph_node.qs2")
} else {
  ""
}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") &&
  nzchar(spatial_fixture) && file.exists(spatial_fixture)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_extended <- run_vignette && identical(real_profile, "all")
run_banksy <- run_vignette &&
  requireNamespace("Banksy", quietly = TRUE) &&
  requireNamespace("SpatialExperiment", quietly = TRUE)
run_nnsvg <- run_extended &&
  requireNamespace("nnSVG", quietly = TRUE) &&
  requireNamespace("SpatialExperiment", quietly = TRUE)
communication_result_path <- Sys.getenv(
  "SHENNONG_SPATIAL_COMMUNICATION_RESULT",
  unset = ""
)
run_spatial_communication <- run_extended && run_banksy &&
  nzchar(communication_result_path) && file.exists(communication_result_path)
external_spatial_enabled <- run_extended && identical(
  Sys.getenv("SHENNONG_RUN_EXTERNAL_BACKENDS", unset = "false"),
  "true"
)
cell2location_signatures_path <- Sys.getenv(
  "SHENNONG_CELL2LOCATION_SIGNATURES",
  unset = ""
)
spatial_reference_path <- Sys.getenv("SHENNONG_SPATIAL_REFERENCE", unset = "")
external_output_root <- file.path(
  real_data_root,
  "results",
  "spatial-external"
)
run_cell2location <- external_spatial_enabled &&
  nzchar(cell2location_signatures_path) &&
  file.exists(cell2location_signatures_path)
run_tangram <- external_spatial_enabled &&
  nzchar(spatial_reference_path) &&
  file.exists(spatial_reference_path)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Shennong's spatial workflow keeps coordinates, graphs, statistics, backend
artifacts, and plots connected through the common result contract. This article
uses the locally materialized two-section human lymph-node Visium fixture under
`SHENNONG_REAL_DATA_DIR` and performs the core analysis on the fresh-frozen
section.

The analysis runs only when the fixture exists and
`SHENNONG_RUN_VIGNETTES=true`. Normal package checks therefore knit the article
without downloading data. Optional R or external adapter backends have
additional, explicit gates and are never replaced with synthetic output.

## Load the real sections and preserve their coordinates

The fixture retains the original Seurat spatial images. Shennong's statistical
workflow consumes two finite metadata columns, so we copy the matching tissue
coordinates into named metadata fields without altering the count matrix or
histology image.

```{r load-spatial-fixture}
library(Shennong)
library(dplyr)

spatial_artifact <- qs2::qs_read(spatial_fixture)
if (!is.list(spatial_artifact$sections) || !length(spatial_artifact$sections)) {
  stop("The Visium fixture does not contain named spatial sections.", call. = FALSE)
}

tibble::tibble(
  section = names(spatial_artifact$sections),
  cells = vapply(spatial_artifact$sections, ncol, numeric(1)),
  features = vapply(spatial_artifact$sections, nrow, numeric(1)),
  platform = vapply(
    spatial_artifact$sections,
    function(section) paste(unique(as.character(section$real_platform)), collapse = ", "),
    character(1)
  )
)

.prepare_spatial_section <- function(section) {
  coordinates <- Seurat::GetTissueCoordinates(section)
  if (!all(colnames(section) %in% rownames(coordinates))) {
    stop("Spatial image coordinates do not align with fixture barcodes.", call. = FALSE)
  }
  coordinate_columns <- if (all(c("x", "y") %in% names(coordinates))) {
    c("x", "y")
  } else if (all(c("imagecol", "imagerow") %in% names(coordinates))) {
    c("imagecol", "imagerow")
  } else {
    numeric_columns <- names(coordinates)[
      vapply(coordinates, is.numeric, logical(1))
    ]
    utils::head(numeric_columns, 2L)
  }
  if (length(coordinate_columns) != 2L) {
    stop("Could not resolve two numeric Visium coordinate columns.", call. = FALSE)
  }
  section$spatial_x <- as.numeric(
    coordinates[colnames(section), coordinate_columns[[1]]]
  )
  section$spatial_y <- as.numeric(
    coordinates[colnames(section), coordinate_columns[[2]]]
  )
  SeuratObject::DefaultAssay(section) <- "Spatial"
  sn_normalize_data(
    section,
    method = "seurat",
    assay = "Spatial",
    layer = "counts",
    verbose = FALSE
  )
}

prepared_sections <- lapply(
  spatial_artifact$sections,
  .prepare_spatial_section
)
section_name <- if ("fresh_frozen" %in% names(prepared_sections)) {
  "fresh_frozen"
} else {
  names(prepared_sections)[[1]]
}
spatial_object <- prepared_sections[[section_name]]

head(spatial_object[[]][, c(
  "real_section", "real_platform", "spatial_x", "spatial_y"
)])
```

## Inspect and integrate both real sections

The unified dispatcher first checks the fresh-frozen coordinate graph. For a
joint two-section view, the code then combines the shared real count features,
computes one joint PCA, and passes a transparent section-centered embedding to
the explicit custom integration adapter. This is a deterministic real-data
adapter demonstration; it is not presented as STAligner or another unavailable
backend.

```{r spatial-qc-integration}
spatial_qc <- sn_run_spatial(
  spatial_object,
  task = "qc",
  spatial_cols = c("spatial_x", "spatial_y"),
  k = 6L
)
spatial_qc$diagnostics

shared_features <- Reduce(intersect, lapply(prepared_sections, rownames))
section_counts <- lapply(prepared_sections, function(section) {
  SeuratObject::LayerData(
    section,
    assay = "Spatial",
    layer = "counts"
  )[shared_features, , drop = FALSE]
})
combined_counts <- do.call(cbind, section_counts)
combined_metadata <- do.call(rbind, unname(lapply(
  prepared_sections,
  function(section) section[[]][colnames(section), , drop = FALSE]
)))
combined_metadata <- combined_metadata[colnames(combined_counts), , drop = FALSE]

integration_object <- SeuratObject::CreateSeuratObject(
  counts = combined_counts,
  assay = "Spatial",
  meta.data = combined_metadata,
  project = "real_lymph_node_sections"
)
integration_object <- sn_normalize_data(
  integration_object,
  method = "seurat",
  assay = "Spatial",
  layer = "counts",
  verbose = FALSE
)
integration_object <- Seurat::FindVariableFeatures(
  integration_object,
  assay = "Spatial",
  nfeatures = min(1000L, nrow(integration_object)),
  verbose = FALSE
)
integration_features <- SeuratObject::VariableFeatures(integration_object)
integration_object <- Seurat::ScaleData(
  integration_object,
  assay = "Spatial",
  features = integration_features,
  verbose = FALSE
)
integration_object <- Seurat::RunPCA(
  integration_object,
  assay = "Spatial",
  features = integration_features,
  npcs = 10L,
  verbose = FALSE
)

joint_pca <- Seurat::Embeddings(integration_object, "pca")
section_labels <- as.character(integration_object$real_section)
global_center <- colMeans(joint_pca)
integrated_pca <- joint_pca
for (section in unique(section_labels)) {
  members <- section_labels == section
  integrated_pca[members, ] <- sweep(
    joint_pca[members, , drop = FALSE],
    2L,
    colMeans(joint_pca[members, , drop = FALSE]),
    "-"
  )
  integrated_pca[members, ] <- sweep(
    integrated_pca[members, , drop = FALSE],
    2L,
    global_center,
    "+"
  )
}
integration_embedding <- data.frame(
  cell = rownames(integrated_pca),
  integrated_pca,
  check.names = FALSE
)

integration_object <- sn_integrate_spatial(
  integration_object,
  method = "custom",
  spatial_cols = c("spatial_x", "spatial_y"),
  store_name = "lymph_node_sections",
  backend_control = list(
    result = list(
      embedding = integration_embedding,
      model = list(transformation = "joint PCA with within-section centering")
    ),
    seed = 717L
  )
)

spatial_integration <- sn_get_result(
  integration_object,
  "spatial_integration",
  "lymph_node_sections"
)
spatial_integration$diagnostics
head(spatial_integration$tables$primary)
sn_plot_spatial(
  integration_object,
  group_by = "real_section",
  spatial_cols = c("spatial_x", "spatial_y")
)
```

## Native spatial neighborhoods from real expression groups

The core neighborhood implementation does not require BANKSY. To give it
non-circular, data-derived labels, we cluster the fresh-frozen spots in the
joint real-expression PCA and then test those labels on the spatial KNN graph.
The expression grouping and spatial enrichment therefore use independent
feature and coordinate evidence from the same real spots.

```{r core-spatial-neighborhoods}
set.seed(717)
section_pca <- integrated_pca[
  colnames(spatial_object),
  seq_len(min(5L, ncol(integrated_pca))),
  drop = FALSE
]
section_groups <- stats::kmeans(
  section_pca,
  centers = 4L,
  nstart = 20L,
  iter.max = 100L
)
spatial_object$core_expression_group <- paste0(
  "expression_",
  section_groups$cluster
)

spatial_object <- sn_run_spatial_neighborhood(
  spatial_object,
  method = "knn",
  group_by = "core_expression_group",
  spatial_cols = c("spatial_x", "spatial_y"),
  store_name = "expression_neighborhood",
  backend_control = list(
    k = 6L,
    n_permutations = 99L,
    distance_bins = 4L,
    seed = 717L
  )
)
core_neighborhood <- sn_get_result(
  spatial_object,
  "spatial_neighborhood",
  "expression_neighborhood"
)
head(core_neighborhood$tables$enrichment, 10L)
head(core_neighborhood$tables$cooccurrence, 10L)
core_neighborhood$diagnostics
sn_plot_spatial_neighborhood(core_neighborhood, type = "enrichment")
sn_plot_spatial_neighborhood(core_neighborhood, type = "cooccurrence")
```

## Spatially variable features with Moran's I

The core backend builds a memory-bounded spatial KNN graph and retains the
observed Moran's I, permutation null, adjusted P value, rank, and graph edges.
The feature cap and permutation count are explicit so the real site build stays
auditable and bounded.

```{r spatial-features}
spatial_object <- sn_run_spatial(
  spatial_object,
  task = "svg",
  method = "morans_i",
  spatial_cols = c("spatial_x", "spatial_y"),
  assay = "Spatial",
  layer = "data",
  store_name = "lymph_node_morans",
  backend_control = list(
    k = 6L,
    n_permutations = 99L,
    max_features = 300L,
    seed = 717L
  )
)

svg <- sn_get_result(
  spatial_object,
  "spatial_features",
  "lymph_node_morans"
)
head(svg$tables$features, 10L)
head(svg$graphs$spatial)
svg$diagnostics

sn_plot_spatial_svg(svg, n = 20L)
sn_plot_spatial_feature(
  spatial_object,
  features = utils::head(svg$tables$features$feature, 4L),
  spatial_cols = c("spatial_x", "spatial_y"),
  assay = "Spatial",
  layer = "data"
)
```

## Domains and neighborhoods

BANKSY is a direct optional R backend. It is used only when both BANKSY and
SpatialExperiment are already installed; the article does not install them and
does not manufacture domain labels when they are unavailable.

```{r banksy-status}
tibble::tibble(
  backend = "BANKSY",
  Banksy_installed = requireNamespace("Banksy", quietly = TRUE),
  SpatialExperiment_installed = requireNamespace("SpatialExperiment", quietly = TRUE),
  will_run = run_banksy
)
```

```{r spatial-domains, eval=run_banksy}
spatial_object <- sn_find_spatial_domains(
  spatial_object,
  method = "banksy",
  spatial_cols = c("spatial_x", "spatial_y"),
  assay = "Spatial",
  layer = "data",
  store_name = "banksy_domains",
  backend_control = list(
    lambda = 0.8,
    resolution = 0.8,
    npcs = 20L,
    seed = 717L
  )
)

domains <- sn_get_result(
  spatial_object,
  "spatial_domains",
  "banksy_domains"
)
head(domains$tables$domains)
domains$diagnostics
sn_plot_spatial_domain(domains)
```

The neighborhood test uses the actual BANKSY assignments written to metadata.
Its enrichment P values come from label permutations on the retained KNN graph;
co-occurrence summarizes graph edges by distance bin.

```{r spatial-neighborhoods, eval=run_banksy}
spatial_object <- sn_run_spatial_neighborhood(
  spatial_object,
  method = "knn",
  group_by = "banksy_domains",
  spatial_cols = c("spatial_x", "spatial_y"),
  store_name = "banksy_neighborhood",
  backend_control = list(
    k = 6L,
    n_permutations = 99L,
    distance_bins = 4L,
    seed = 717L
  )
)

neighborhood <- sn_get_result(
  spatial_object,
  "spatial_neighborhood",
  "banksy_neighborhood"
)
head(neighborhood$tables$enrichment, 10L)
head(neighborhood$tables$cooccurrence, 10L)
neighborhood$diagnostics

sn_plot_spatial_neighborhood(neighborhood, type = "enrichment")
sn_plot_spatial_neighborhood(neighborhood, type = "cooccurrence")
```

## Extended spatial-feature backend

nnSVG runs only in the `all` profile and only when its two R dependencies are
already installed. SPARK-X, stLearn, BayesSpace, CellCharter, and Squidpy use
explicit runner/result boundaries; no placeholder result is created here.

```{r nnsvg-status}
tibble::tibble(
  profile = real_profile,
  nnSVG_installed = requireNamespace("nnSVG", quietly = TRUE),
  SpatialExperiment_installed = requireNamespace("SpatialExperiment", quietly = TRUE),
  will_run_nnSVG = run_nnsvg
)
```

```{r nnsvg, eval=run_nnsvg}
nnsvg <- sn_find_spatial_features(
  spatial_object,
  method = "nnsvg",
  spatial_cols = c("spatial_x", "spatial_y"),
  assay = "Spatial",
  layer = "data",
  features = utils::head(svg$tables$features$feature, 100L),
  store_name = "lymph_node_nnsvg",
  return_object = FALSE
)

head(nnsvg$tables$features, 10L)
nnsvg$diagnostics
sn_plot_spatial_svg(nnsvg, n = 20L)
```

## Distance-aware communication needs independent interaction evidence

Spatial proximity cannot create ligand-receptor evidence. This step becomes
eligible only in the `all` profile, after BANKSY has produced real groups, and
when `SHENNONG_SPATIAL_COMMUNICATION_RESULT` points to a real Shennong
cell-communication result serialized with `qs2`. The source and target labels
in that result must match the domain labels used here.

```{r spatial-communication-status}
tibble::tibble(
  result_file = if (nzchar(communication_result_path)) {
    basename(communication_result_path)
  } else {
    NA_character_
  },
  result_available = nzchar(communication_result_path) &&
    file.exists(communication_result_path),
  will_run = run_spatial_communication
)
```

```{r spatial-communication, eval=run_spatial_communication}
communication <- qs2::qs_read(communication_result_path)
spatial_communication <- sn_run_spatial_communication(
  spatial_object,
  communication = communication,
  group_by = "banksy_domains",
  spatial_cols = c("spatial_x", "spatial_y"),
  store_name = "domain_spatial_communication",
  return_object = FALSE
)

head(spatial_communication$tables$primary, 10L)
head(spatial_communication$tables$group_distances, 10L)
spatial_communication$diagnostics
sn_plot_spatial_communication(spatial_communication)
```

## External spatial adapters require explicit real inputs

The `all` profile still does not authorize environment creation or downloads.
Python object workflows run only when `SHENNONG_RUN_EXTERNAL_BACKENDS=true` is
set deliberately for already-prepared local environments. Cell2location also
requires a real signature CSV, and Tangram requires a real, biologically matched
single-cell reference serialized with `qs2`. The status table therefore makes
every unmet gate visible instead of reporting an optional backend as successful.

```{r external-spatial-status}
tibble::tibble(
  workflow = c(
    "cell2location", "Tangram", "Squidpy", "SpatialData", "stLearn"
  ),
  profile_all = run_extended,
  external_execution_enabled = external_spatial_enabled,
  required_real_input_available = c(
    nzchar(cell2location_signatures_path) &&
      file.exists(cell2location_signatures_path),
    nzchar(spatial_reference_path) && file.exists(spatial_reference_path),
    TRUE,
    TRUE,
    TRUE
  ),
  will_run = c(
    run_cell2location,
    run_tangram,
    rep(external_spatial_enabled, 3L)
  )
)
```

The two deconvolution entry points are aliases. The environment variable
`SHENNONG_SPATIAL_DECONVOLUTION_ENTRYPOINT` can select `direct` for
`sn_run_cell2location()`; the default uses the unified
`sn_run_spatial_deconvolution()` name. Only the metadata columns imported from
the real posterior are reshaped for plotting.

```{r run-cell2location, eval=run_cell2location}
deconvolution_entrypoint <- Sys.getenv(
  "SHENNONG_SPATIAL_DECONVOLUTION_ENTRYPOINT",
  unset = "unified"
)
if (identical(deconvolution_entrypoint, "direct")) {
  deconvolved_object <- sn_run_cell2location(
    spatial_object,
    assay = "Spatial",
    layer = "counts",
    reference_signatures = cell2location_signatures_path,
    spatial_cols = c("spatial_x", "spatial_y"),
    output_dir = file.path(external_output_root, "cell2location"),
    result_name = "lymph_node_cell2location"
  )
} else {
  deconvolved_object <- sn_run_spatial_deconvolution(
    spatial_object,
    assay = "Spatial",
    layer = "counts",
    reference_signatures = cell2location_signatures_path,
    spatial_cols = c("spatial_x", "spatial_y"),
    output_dir = file.path(external_output_root, "cell2location"),
    result_name = "lymph_node_cell2location"
  )
}

deconvolution_manifest <- deconvolved_object@misc$cell2location[[
  "lymph_node_cell2location"
]]
abundance_columns <- deconvolution_manifest$imported_metadata
if (!length(abundance_columns)) {
  stop("The real cell2location run imported no abundance columns.", call. = FALSE)
}
deconvolution_table <- dplyr::bind_rows(lapply(abundance_columns, function(column) {
  tibble::tibble(
    cell = colnames(deconvolved_object),
    spatial_x = deconvolved_object$spatial_x,
    spatial_y = deconvolved_object$spatial_y,
    cell_type = sub("^cell2location_", "", column),
    proportion = as.numeric(deconvolved_object[[column, drop = TRUE]])
  )
}))
sn_plot_spatial_deconvolution(deconvolution_table)
```

Tangram likewise exposes a direct and a unified mapping entry point. The
selected call consumes the real local reference object; it never substitutes a
fixture-derived pseudo-reference.

```{r run-tangram, eval=run_tangram}
spatial_reference <- qs2::qs_read(spatial_reference_path)
if (!inherits(spatial_reference, "Seurat")) {
  stop("SHENNONG_SPATIAL_REFERENCE must contain a Seurat object.", call. = FALSE)
}
mapping_entrypoint <- Sys.getenv(
  "SHENNONG_SPATIAL_MAPPING_ENTRYPOINT",
  unset = "unified"
)
if (identical(mapping_entrypoint, "direct")) {
  mapped_object <- sn_run_tangram(
    spatial_object,
    reference_object = spatial_reference,
    assay = "Spatial",
    layer = "counts",
    spatial_cols = c("spatial_x", "spatial_y"),
    cell_type_by = "cell_type",
    output_dir = file.path(external_output_root, "tangram"),
    result_name = "lymph_node_tangram"
  )
} else {
  mapped_object <- sn_run_spatial_mapping(
    spatial_object,
    reference_object = spatial_reference,
    assay = "Spatial",
    layer = "counts",
    spatial_cols = c("spatial_x", "spatial_y"),
    cell_type_by = "cell_type",
    output_dir = file.path(external_output_root, "tangram"),
    result_name = "lymph_node_tangram"
  )
}
mapped_object@misc$tangram$lymph_node_tangram
```

Squidpy, SpatialData, and stLearn need no additional biological reference, but
they remain behind the same explicit already-prepared-environment gate.

```{r run-external-spatial-objects, eval=external_spatial_enabled}
squidpy_object <- sn_run_squidpy(
  spatial_object,
  assay = "Spatial",
  layer = "data",
  spatial_cols = c("spatial_x", "spatial_y"),
  output_dir = file.path(external_output_root, "squidpy"),
  result_name = "lymph_node_squidpy"
)
spatialdata_object <- sn_run_spatialdata(
  spatial_object,
  assay = "Spatial",
  layer = "data",
  spatial_cols = c("spatial_x", "spatial_y"),
  output_dir = file.path(external_output_root, "spatialdata"),
  result_name = "lymph_node_spatialdata"
)
stlearn_object <- sn_run_stlearn(
  spatial_object,
  assay = "Spatial",
  layer = "data",
  spatial_cols = c("spatial_x", "spatial_y"),
  output_dir = file.path(external_output_root, "stlearn"),
  result_name = "lymph_node_stlearn"
)

tibble::tibble(
  backend = c("Squidpy", "SpatialData", "stLearn"),
  result_recorded = c(
    !is.null(squidpy_object@misc$squidpy$lymph_node_squidpy),
    !is.null(spatialdata_object@misc$spatialdata$lymph_node_spatialdata),
    !is.null(stlearn_object@misc$stlearn$lymph_node_stlearn)
  )
)
```
