---
title: "Spatial Feature, Domain, and Neighborhood Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Feature, Domain, and Neighborhood Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Shennong's spatial workflow keeps coordinates, graphs, statistics, backend
artifacts, and plots connected through the common result contract. Coordinate
columns may be supplied explicitly or discovered from common metadata names.

## Spatial features

```{r svg}
object <- sn_find_spatial_features(
  object,
  method = "morans_i",
  spatial_cols = c("spatial_x", "spatial_y"),
  backend_control = list(k = 6, n_permutations = 999, seed = 717)
)

svg <- sn_get_result(object, "spatial_features", "spatial_features")
head(svg$tables$features)
head(svg$graphs$spatial)
sn_plot_spatial_svg(svg)
sn_plot_spatial_feature(object, svg$tables$features$feature[1:4])
```

Use `method = "nnsvg"` for the optional nearest-neighbor Gaussian-process
backend. SPARK-X enters through an explicit runner/result adapter.

## Domains and neighborhoods

```{r domains-neighborhoods}
object <- sn_find_spatial_domains(
  object,
  method = "banksy",
  store_name = "banksy_domains",
  backend_control = list(lambda = 0.8, resolution = 1, seed = 717)
)
sn_plot_spatial_domain(object, "banksy_domains")

object <- sn_run_spatial_neighborhood(
  object,
  group_by = "cell_type",
  backend_control = list(k = 6, n_permutations = 999)
)
neighborhood <- sn_get_result(object, "spatial_neighborhood", "spatial_neighborhood")
sn_plot_spatial_neighborhood(neighborhood, type = "enrichment")
sn_plot_spatial_neighborhood(neighborhood, type = "cooccurrence")
```

The local graph avoids allocating a full all-by-all distance matrix. Squidpy is
available as the heavier backend when its graph ecosystem is needed.

## Distance-aware communication

Run `sn_run_cell_communication()` first. Then augment that evidence with
spatial distances:

```{r communication}
object <- sn_run_spatial_communication(
  object,
  communication_name = "communication",
  group_by = "cell_type",
  max_distance = 200,
  store_name = "spatial_communication"
)

spatial_communication <- sn_get_result(
  object, "spatial_communication", "spatial_communication"
)
head(spatial_communication$tables$group_distances)
sn_plot_spatial_communication(spatial_communication)
```

Distance filtering does not create ligand-receptor evidence; it only adds a
proximity constraint to an already auditable communication result.
