---
title: "Trajectory, Pseudotime, and Dynamic Genes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trajectory, Pseudotime, and Dynamic Genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Shennong uses Slingshot as the default cluster-aware trajectory backend and
tradeSeq for negative-binomial smooth models. The workflow stores a versioned
result instead of only adding a single pseudotime vector to cell metadata.

## Infer and store a trajectory

```{r, eval=FALSE}
library(Shennong)

object <- sn_run_trajectory(
  object,
  reduction = "pca",
  cluster_by = "seurat_clusters",
  start = "naive",
  end = c("effector", "memory"),
  store_name = "t_cell_development"
)
```

`start` and `end` are cluster labels. For an expected topology, provide named
cluster paths with `lineages`; Shennong constrains the endpoints and records a
warning if Slingshot does not recover a path exactly.

## Discover and retrieve the stored result

```{r, eval=FALSE}
sn_list_results(object, type = "trajectory")

trajectory <- sn_get_result(
  object,
  type = "trajectory",
  name = "t_cell_development"
)

trajectory$tables$cells
trajectory$tables$terminal_states
trajectory$graphs$lineages
trajectory$diagnostics
```

The cell table contains primary pseudotime plus a pseudotime and probability
for every lineage. The object metadata also contains
`t_cell_development_pseudotime` and `t_cell_development_lineage` for direct use
with Seurat.

## Inspect dynamic genes and branches

```{r, eval=FALSE}
trajectory$tables$dynamic_genes |>
  dplyr::arrange(adjusted_p_value)

trajectory$tables$branch_genes |>
  dplyr::filter(test == "pattern") |>
  dplyr::arrange(adjusted_p_value)

trajectory$tables$fitted_trends
trajectory$tables$convergence
```

By default, tradeSeq tests variable features when available, capped at 2,000.
Use `dynamic_features` to define an explicit auditable set. Raw counts remain
sparse at the Shennong boundary; tradeSeq controls any backend conversion.

## Plot the result

```{r, eval=FALSE}
sn_plot_trajectory(object, "t_cell_development")
sn_plot_pseudotime(object, "t_cell_development", lineage = "Lineage1")
sn_plot_lineage_probability(object, "t_cell_development", lineage = "Lineage1")
sn_plot_dynamic_heatmap(object, "t_cell_development")
sn_plot_gene_trend(object, "t_cell_development", c("GZMB", "TCF7"))
sn_plot_branch_comparison(object, "t_cell_development", test = "pattern")
```

`test_dynamic = FALSE` runs trajectory inference without tradeSeq. This is
useful for topology review before committing compute to gene-wise models.
