---
title: "Trajectory, Pseudotime, and Dynamic Genes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trajectory, Pseudotime, and Dynamic Genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
real_data_root <- Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = "")
trajectory_fixture <- if (nzchar(real_data_root)) {
  file.path(real_data_root, "trajectory", "hermann_spermatogenesis.qs2")
} else {
  ""
}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") &&
  nzchar(trajectory_fixture) && file.exists(trajectory_fixture)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_extended <- run_vignette && identical(real_profile, "all")
run_monocle3 <- run_extended && requireNamespace("monocle3", quietly = TRUE)
palantir_result_path <- Sys.getenv("SHENNONG_PALANTIR_RESULT", unset = "")
run_palantir <- run_extended && nzchar(palantir_result_path) &&
  file.exists(palantir_result_path)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Shennong uses Slingshot as the default cluster-aware trajectory backend and
tradeSeq for negative-binomial smooth models. The workflow stores a versioned
result instead of only adding a single pseudotime vector to cell metadata.

This article reads the locally materialized Hermann mouse spermatogenesis
fixture from `SHENNONG_REAL_DATA_DIR`. The analysis runs only when that file is
present and `SHENNONG_RUN_VIGNETTES=true`; a normal package check therefore
knits the article without downloading data or executing the analysis.

## Load and prepare the real trajectory fixture

The fixture retains total counts plus the original `spliced` and `unspliced`
layers. We remove only cells whose source annotation is missing, then use
Shennong's standard clustering workflow to create the PCA and UMAP reductions
needed by downstream trajectory plots.

```{r load-trajectory-fixture}
library(Shennong)
library(dplyr)

trajectory_object <- qs2::qs_read(trajectory_fixture)
annotated <- !is.na(trajectory_object$real_cell_type) &
  nzchar(as.character(trajectory_object$real_cell_type))
trajectory_object <- subset(
  trajectory_object,
  cells = colnames(trajectory_object)[annotated]
)

tibble::tibble(
  fixture_file = basename(trajectory_fixture),
  cells = ncol(trajectory_object),
  features = nrow(trajectory_object),
  layers = paste(SeuratObject::Layers(trajectory_object[["RNA"]]), collapse = ", ")
)

sort(table(trajectory_object$real_cell_type), decreasing = TRUE)

trajectory_object <- sn_run_cluster(
  object = trajectory_object,
  normalization_method = "seurat",
  nfeatures = min(1500L, nrow(trajectory_object)),
  npcs = 30,
  dims = 1:20,
  resolution = 0.6,
  cluster_algorithm = "louvain",
  cluster_name = "trajectory_cluster",
  species = "mouse",
  verbose = FALSE
)

sn_plot_dim(
  trajectory_object,
  reduction = "umap",
  group_by = "real_cell_type",
  title = "Hermann spermatogenesis cell types"
)
```

## Infer and store the core trajectory

Trajectory orientation is not inferred from the embedding. The code resolves
an early spermatogonial label and a late spermatid label from the source
annotations, then maps each biological anchor to its dominant unsupervised
cluster. This preserves the real source labels as endpoint evidence while
avoiding unstable lineage geometry from very small source-label groups. If
either anchor cannot be resolved, the article stops rather than guessing an
orientation.

```{r run-trajectory}
first_matching_label <- function(labels, patterns, role) {
  for (pattern in patterns) {
    hit <- labels[grepl(pattern, labels, ignore.case = TRUE)]
    if (length(hit)) return(hit[[1]])
  }
  stop(
    "Could not resolve the ", role, " trajectory label from: ",
    paste(labels, collapse = ", "),
    call. = FALSE
  )
}

source_labels <- sort(unique(as.character(trajectory_object$real_cell_type)))
start_label <- first_matching_label(
  source_labels,
  c("spermatogon", "undifferentiated", "^spg", "stem"),
  "start"
)
end_label <- first_matching_label(
  source_labels,
  c(
    "late.*round.*spermatid", "elongat.*spermatid",
    "round.*spermatid", "spermatid", "^spt"
  ),
  "end"
)
if (identical(start_label, end_label)) {
  stop("Resolved trajectory start and end labels are identical.", call. = FALSE)
}

dominant_cluster <- function(source_label) {
  members <- as.character(trajectory_object$real_cell_type) == source_label
  counts <- sort(
    table(as.character(trajectory_object$trajectory_cluster[members])),
    decreasing = TRUE
  )
  if (!length(counts)) {
    stop("No unsupervised cluster represents source label `", source_label, "`.", call. = FALSE)
  }
  names(counts)[[1]]
}

start_cluster <- dominant_cluster(start_label)
end_cluster <- dominant_cluster(end_label)
if (identical(start_cluster, end_cluster)) {
  stop(
    "Biological trajectory endpoints map to the same unsupervised cluster `",
    start_cluster, "`.",
    call. = FALSE
  )
}

counts <- SeuratObject::LayerData(
  trajectory_object,
  assay = "RNA",
  layer = "counts"
)
detected_cells <- Matrix::rowSums(counts > 0)
stable_hvgs <- SeuratObject::VariableFeatures(trajectory_object)
stable_hvgs <- stable_hvgs[detected_cells[stable_hvgs] >= 50L]
dynamic_features <- utils::head(stable_hvgs, 10L)
if (length(dynamic_features) < 4L) {
  stop(
    "Fewer than four variable genes are detected in at least 50 real cells.",
    call. = FALSE
  )
}
trend_features <- utils::head(dynamic_features, 4L)

tibble::tibble(
  endpoint = c("start", "end"),
  source_label = c(start_label, end_label),
  dominant_cluster = c(start_cluster, end_cluster),
  dynamic_features = length(dynamic_features)
)

trajectory_object <- sn_run_trajectory(
  trajectory_object,
  method = "slingshot",
  reduction = "pca",
  cluster_by = "trajectory_cluster",
  start = start_cluster,
  end = end_cluster,
  store_name = "spermatogenesis",
  dynamic_features = dynamic_features,
  max_dynamic_features = length(dynamic_features),
  nknots = 3L,
  trend_features = trend_features,
  trend_points = 40L,
  backend_control = list(seed = 717L)
)
```

## Inspect the retained evidence

The cell table retains pseudotime and lineage probability for every inferred
lineage. Dynamic-gene, fitted-trend, convergence, and terminal-state tables are
kept with the topology rather than recomputed inside plotting code.

```{r inspect-trajectory}
sn_list_results(trajectory_object, type = "trajectory")

trajectory <- sn_get_result(
  trajectory_object,
  type = "trajectory",
  name = "spermatogenesis"
)

head(trajectory$tables$cells)
trajectory$tables$terminal_states
trajectory$graphs$lineages
trajectory$diagnostics

trajectory$tables$dynamic_genes |>
  dplyr::arrange(.data$adjusted_p_value) |>
  head(10L)

trajectory$tables$convergence |>
  head(10L)
```

## Plot topology, pseudotime, and gene dynamics

The lineage name is taken from the stored graph, so the probability plot stays
aligned with the backend result instead of assuming a hard-coded name.

```{r plot-trajectory}
lineage_name <- names(trajectory$graphs$lineages)[[1]]

sn_plot_trajectory(trajectory)
sn_plot_pseudotime(trajectory, lineage = lineage_name)
sn_plot_lineage_probability(trajectory, lineage = lineage_name)
sn_plot_dynamic_heatmap(trajectory)
sn_plot_gene_trend(trajectory, features = trend_features)
```

```{r plot-trajectory-branches, eval=run_vignette && trajectory$diagnostics$n_lineages > 1L}
sn_plot_branch_comparison(trajectory, test = "pattern")
```

## Extended backends are explicit

The `all` profile may run Monocle 3 only when its R dependency is already
installed. The article does not install packages while knitting.

```{r extended-trajectory-status}
tibble::tibble(
  backend = c("monocle3", "palantir"),
  profile_all = run_extended,
  dependency_or_result_available = c(
    requireNamespace("monocle3", quietly = TRUE),
    nzchar(palantir_result_path) && file.exists(palantir_result_path)
  ),
  will_run = c(run_monocle3, run_palantir)
)
```

```{r run-monocle3, eval=run_monocle3}
monocle <- sn_run_trajectory(
  trajectory_object,
  method = "monocle3",
  reduction = "umap",
  cluster_by = "real_cell_type",
  start = start_label,
  test_dynamic = FALSE,
  return_object = FALSE
)

head(monocle$tables$cells)
sn_plot_pseudotime(monocle)
```

Palantir is an explicit adapter boundary. It runs only when the `all` profile
is active and `SHENNONG_PALANTIR_RESULT` points to a real, named per-cell result
serialized with `qs2`; this article never fabricates adapter output.

```{r run-palantir, eval=run_palantir}
palantir_result <- qs2::qs_read(palantir_result_path)
palantir <- sn_run_trajectory(
  trajectory_object,
  method = "palantir",
  reduction = "pca",
  cluster_by = "real_cell_type",
  start = start_label,
  end = end_label,
  test_dynamic = FALSE,
  backend_control = list(result = palantir_result),
  return_object = FALSE
)

sn_validate_result(palantir)
head(palantir$tables$cells)
sn_plot_trajectory(palantir)
```
