---
title: "RNA Velocity and Fate Mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RNA Velocity and Fate Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
real_data_root <- Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = "")
velocity_fixture <- if (nzchar(real_data_root)) {
  file.path(real_data_root, "trajectory", "hermann_spermatogenesis.qs2")
} else {
  ""
}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") &&
  nzchar(velocity_fixture) && file.exists(velocity_fixture)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_extended <- run_vignette && identical(real_profile, "all")
pixi_info <- list(installed = FALSE, path = NA_character_)
trajectory_environment <- ""
run_velocity_backend <- FALSE
regvelo_prior <- Sys.getenv("SHENNONG_REGVELO_PRIOR", unset = "")
run_regvelo <- FALSE

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Shennong keeps velocity and fate mapping as separate stored analyses. Velocity
estimates a directed transition process from spliced and unspliced counts;
CellRank then analyzes the retained transition process with GPCCA.

This article uses the real Hermann mouse spermatogenesis fixture under
`SHENNONG_REAL_DATA_DIR`. The core profile loads and checks the real layers,
runs Shennong clustering, and plots the biological annotations. Managed Python
backends run only in the `all` profile and only when an existing pixi binary and
an already materialized trajectory environment are both available.

## Verify the real layers and embedding

The fixture was prepared from source spliced and unspliced assays. No layer is
synthesized inside the article. Cells without a source annotation are removed
before clustering so every displayed group is traceable to the source data.

```{r load-velocity-fixture}
library(Shennong)
library(dplyr)

velocity_object <- qs2::qs_read(velocity_fixture)
annotated <- !is.na(velocity_object$real_cell_type) &
  nzchar(as.character(velocity_object$real_cell_type))
velocity_object <- subset(
  velocity_object,
  cells = colnames(velocity_object)[annotated]
)

required_layers <- c("counts", "spliced", "unspliced")
available_layers <- SeuratObject::Layers(velocity_object[["RNA"]])
if (!all(required_layers %in% available_layers)) {
  stop(
    "The Hermann fixture is missing required RNA layers: ",
    paste(setdiff(required_layers, available_layers), collapse = ", "),
    call. = FALSE
  )
}

dplyr::bind_rows(lapply(required_layers, function(layer) {
  matrix <- SeuratObject::LayerData(velocity_object, assay = "RNA", layer = layer)
  tibble::tibble(layer = layer, features = nrow(matrix), cells = ncol(matrix))
}))

sort(table(velocity_object$real_cell_type), decreasing = TRUE)

velocity_object <- sn_run_cluster(
  object = velocity_object,
  normalization_method = "seurat",
  nfeatures = min(1500L, nrow(velocity_object)),
  npcs = 30,
  dims = 1:20,
  resolution = 0.6,
  cluster_algorithm = "louvain",
  cluster_name = "velocity_cluster",
  species = "mouse",
  verbose = FALSE
)

sn_plot_dim(
  velocity_object,
  reduction = "umap",
  group_by = "real_cell_type",
  title = "Hermann spermatogenesis before velocity inference"
)
```

## Check the managed backend without installing it

`sn_run_velocity()` invokes a managed Python environment. To keep site builds
reproducible, this article never calls an installer. The extended chunks become
eligible only when `SHENNONG_REAL_PROFILE=all`, pixi is discoverable, and the
trajectory environment already exists on disk.

```{r velocity-backend-status}
pixi_info <- sn_check_pixi(quiet = TRUE)
trajectory_paths <- sn_get_pixi_paths("trajectory")
trajectory_environment <- file.path(
  trajectory_paths$workspace_env_dir,
  "default"
)
run_velocity_backend <- run_extended && isTRUE(pixi_info$installed) &&
  dir.exists(trajectory_environment)

tibble::tibble(
  profile = real_profile,
  pixi_installed = isTRUE(pixi_info$installed),
  environment = "trajectory/default",
  environment_ready = dir.exists(trajectory_environment),
  will_run_velocity = run_velocity_backend
)

pixi_control <- list(
  pixi = if (isTRUE(pixi_info$installed)) pixi_info$path else NULL,
  pixi_environment = "default",
  install_pixi = FALSE,
  quiet = TRUE
)
```

## Estimate velocity in the extended profile

The source spliced and unspliced matrices contain finite, non-negative
fractional molecule estimates. `enforce_normalization = TRUE` prevents scVelo
from mistaking those non-integer values for already log-transformed data; the
normalized expression is then log1p-transformed before Seurat-flavor HVG
selection. Both choices are retained in the unified result parameters and
backend manifest. This fixture uses scVelo's explicit deterministic model;
stochastic inference remains available through `backend_control` but is not
silently substituted if its generalized least-squares backend fails.

```{r velocity, eval=run_velocity_backend}
velocity_object <- sn_run_velocity(
  velocity_object,
  method = "scvelo",
  spliced_assay = "RNA",
  spliced_layer = "spliced",
  unspliced_assay = "RNA",
  unspliced_layer = "unspliced",
  reduction = "umap",
  result_id = "hermann_scvelo",
  backend_control = list(
    velocity_mode = "deterministic",
    enforce_normalization = TRUE,
    log1p_transform = TRUE,
    n_top_genes = 1500L,
    n_neighbors = 30L,
    pixi = pixi_control
  ),
  seed = 717L
)

velocity <- sn_get_result(velocity_object, "velocity", "hermann_scvelo")
head(velocity$tables$cells)
head(velocity$tables$transition_edges)
velocity$diagnostics
velocity_manifest <- velocity$models$artifacts
velocity_manifest$output_h5ad <- file.exists(velocity_manifest$output_h5ad)
velocity_manifest

sn_plot_velocity(velocity, color_by = "confidence")
```

Managed velocity runs retain their output directory by default because the
stored H5AD is the input to `sn_run_fate()`. The redundant CSV/Matrix Market
input export is removed after a successful run, and the manifest records the
retained directory and reason. If fate inference is not needed, set
`backend_control$keep_run_dir = FALSE`; Shennong then imports the velocity
tables, removes its owned run directory, and intentionally stores no reusable
H5AD path. An explicit `backend_control$run_dir` must be empty.

Low confidence, weak splicing signal, or unstable directionality remains
visible in the stored diagnostics. The H5AD artifact recorded in
`models$artifacts` is the transition evidence consumed by CellRank; a smooth
arrow field is not sufficient evidence by itself.

Imported velocity output must contain unique known cell IDs, either two finite
vector components or two missing components per row, and at least one finite
vector overall. Transition endpoints must name known cells and weights must be
finite. A custom runner or precomputed result is recorded as such in `backend`;
it is not mislabeled as a pixi execution.

## Estimate terminal states and fate

CellRank runs only after the real scVelo result has been stored. The default
stability rule is kept explicit rather than silently replaced by a fixed number
of terminal states.

```{r fate, eval=run_velocity_backend}
velocity_object <- sn_run_fate(
  velocity_object,
  source_result_id = "hermann_scvelo",
  reduction = "umap",
  result_id = "hermann_fate",
  backend_control = list(
    terminal_method = "stability",
    pixi = pixi_control
  ),
  seed = 717L
)

fate <- sn_get_result(velocity_object, "fate", "hermann_fate")
fate$tables$terminal_states
head(fate$tables$probabilities)
head(fate$tables$lineage_drivers)
fate$diagnostics

sn_plot_fate(fate)
```

If no macrostate meets the stability threshold, the backend raises an error.
A documented project may instead choose `terminal_method = "top_n"` or supply
explicit terminal states, but that decision should not be made automatically
inside a teaching article.

Fate import also verifies that every terminal state occurs in the returned
probability table and that probability cells belong to the object. Its
package-owned temporary run directory is removed after successful import by
default. Supplying `backend_control$run_dir` retains that explicit directory;
set `keep_run_dir = FALSE` to request cleanup while treating the supplied path
as a parent rather than deleting the parent itself.

## RegVelo requires a real, versioned prior

RegVelo is not run with a toy regulatory network. It becomes eligible only when
the extended backend is ready and `SHENNONG_REGVELO_PRIOR` names an existing
regulator-target CSV supplied by the project.

```{r regvelo-status}
run_regvelo <- run_velocity_backend && nzchar(regvelo_prior) &&
  file.exists(regvelo_prior)

tibble::tibble(
  prior_path = regvelo_prior,
  prior_available = nzchar(regvelo_prior) && file.exists(regvelo_prior),
  will_run_regvelo = run_regvelo
)
```

```{r regvelo, eval=run_regvelo}
regvelo_object <- sn_run_velocity(
  velocity_object,
  method = "regvelo",
  spliced_assay = "RNA",
  spliced_layer = "spliced",
  unspliced_assay = "RNA",
  unspliced_layer = "unspliced",
  reduction = "umap",
  result_id = "hermann_regvelo",
  backend_control = list(
    prior_grn = regvelo_prior,
    max_epochs = 1500L,
    early_stopping = TRUE,
    seed = 717L,
    pixi = pixi_control
  )
)

regvelo <- sn_get_result(regvelo_object, "velocity", "hermann_regvelo")
head(regvelo$tables$cells)
head(regvelo$tables$transition_edges)
regvelo$models$artifacts
regvelo$diagnostics

sn_plot_velocity(regvelo, color_by = "pseudotime")
```

The prior GRN is a modeling assumption and must be versioned with the analysis.
The stored result records the actual prior-edge and model artifacts returned by
the backend; this article never substitutes fabricated adapter output.
