---
title: "RNA Velocity and Fate Mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RNA Velocity and Fate Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Shennong runs scVelo, RegVelo, and CellRank in a managed, isolated CPU pixi environment.
The two stages remain separate stored analyses: velocity estimates a directed
transition process from spliced/unspliced counts, while fate mapping analyzes
that transition process with CellRank's GPCCA estimator.

## Prepare the backend

```{r setup-pixi}
sn_prepare_pixi_environment(
  "trajectory",
  install_environment = TRUE
)
```

The Seurat object must contain raw `spliced` and `unspliced` layers with shared
features and cells, plus a dimensional reduction used only to project the
high-dimensional velocity vectors.

## Estimate velocity

```{r velocity}
object <- sn_run_velocity(
  object,
  spliced_layer = "spliced",
  unspliced_layer = "unspliced",
  reduction = "umap",
  store_name = "velocity",
  backend_control = list(
    velocity_mode = "stochastic",
    n_neighbors = 30,
    seed = 717
  )
)

velocity <- sn_get_result(object, "velocity", "velocity")
head(velocity$tables$cells)
head(velocity$tables$transition_edges)
velocity$diagnostics
sn_plot_velocity(object, "velocity", color_by = "confidence")
```

Low confidence, weak splicing signal, or unstable directionality must remain
visible in interpretation. The H5AD artifact recorded in `models$artifacts`
contains the transition evidence consumed by CellRank.

## Add a regulatory prior with RegVelo

RegVelo uses the same spliced/unspliced and embedding inputs, plus an explicit
gene-regulatory prior. Supply either a regulator-target edge table, a named
target-by-regulator matrix, or a CSV path. Shennong retains the trained model,
H5AD, prior-edge count, velocity vectors, transition graph, and RegVelo latent
time under the standard velocity result contract.

```{r regvelo}
prior_grn <- data.frame(
  regulator = c("SOX10", "SOX10", "TFAP2A"),
  target = c("ERBB3", "PLP1", "SOX10"),
  weight = c(1, 0.8, 1)
)

object <- sn_run_velocity(
  object,
  method = "regvelo",
  spliced_layer = "spliced",
  unspliced_layer = "unspliced",
  reduction = "umap",
  store_name = "regvelo",
  backend_control = list(
    prior_grn = prior_grn,
    max_epochs = 1500,
    early_stopping = TRUE,
    seed = 717
  )
)

regvelo <- sn_get_result(object, "velocity", "regvelo")
regvelo$tables$cells
regvelo$models$artifacts
```

The prior GRN is a modeling assumption and must be versioned with the analysis.
Do not interpret a smooth vector field as independent validation of that prior.

## Estimate terminal states and fate

CellRank can consume the retained H5AD produced by either scVelo or RegVelo.

```{r fate}
object <- sn_run_fate(
  object,
  velocity_name = "velocity",
  store_name = "fate",
  backend_control = list(
    n_states = 6,
    terminal_method = "stability"
  )
)

fate <- sn_get_result(object, "fate", "fate")
head(fate$tables$terminal_states)
head(fate$tables$probabilities)
head(fate$tables$lineage_drivers)
sn_plot_fate(object, "fate")
```

If no macrostate meets the default stability threshold, CellRank raises an
actionable error. A documented workflow may use `terminal_method = "top_n"`,
`terminal_n_states`, or explicit `terminal_states`; Shennong records those
choices in the result parameters rather than silently relaxing the rule.
