---
title: "RNA Velocity and Fate Mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RNA Velocity and Fate Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
real_data_root <- Sys.getenv("SHENNONG_REAL_DATA_DIR", unset = "")
velocity_fixture <- if (nzchar(real_data_root)) {
  file.path(real_data_root, "trajectory", "hermann_spermatogenesis.qs2")
} else {
  ""
}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true") &&
  nzchar(velocity_fixture) && file.exists(velocity_fixture)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_extended <- run_vignette && identical(real_profile, "all")
pixi_info <- list(installed = FALSE, path = NA_character_)
trajectory_environment <- ""
run_velocity_backend <- FALSE
regvelo_prior <- Sys.getenv("SHENNONG_REGVELO_PRIOR", unset = "")
run_regvelo <- FALSE

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Shennong keeps velocity and fate mapping as separate stored analyses. Velocity
estimates a directed transition process from spliced and unspliced counts;
CellRank then analyzes the retained transition process with GPCCA.

This article uses the real Hermann mouse spermatogenesis fixture under
`SHENNONG_REAL_DATA_DIR`. The core profile loads and checks the real layers,
runs Shennong clustering, and plots the biological annotations. Managed Python
backends run only in the `all` profile and only when an existing pixi binary and
an already materialized trajectory environment are both available.

## Verify the real layers and embedding

The fixture was prepared from source spliced and unspliced assays. No layer is
synthesized inside the article. Cells without a source annotation are removed
before clustering so every displayed group is traceable to the source data.

```{r load-velocity-fixture}
library(Shennong)
library(dplyr)

velocity_object <- qs2::qs_read(velocity_fixture)
annotated <- !is.na(velocity_object$real_cell_type) &
  nzchar(as.character(velocity_object$real_cell_type))
velocity_object <- subset(
  velocity_object,
  cells = colnames(velocity_object)[annotated]
)

required_layers <- c("counts", "spliced", "unspliced")
available_layers <- SeuratObject::Layers(velocity_object[["RNA"]])
if (!all(required_layers %in% available_layers)) {
  stop(
    "The Hermann fixture is missing required RNA layers: ",
    paste(setdiff(required_layers, available_layers), collapse = ", "),
    call. = FALSE
  )
}

dplyr::bind_rows(lapply(required_layers, function(layer) {
  matrix <- SeuratObject::LayerData(velocity_object, assay = "RNA", layer = layer)
  tibble::tibble(layer = layer, features = nrow(matrix), cells = ncol(matrix))
}))

sort(table(velocity_object$real_cell_type), decreasing = TRUE)

velocity_object <- sn_run_cluster(
  object = velocity_object,
  normalization_method = "seurat",
  nfeatures = min(1500L, nrow(velocity_object)),
  npcs = 30,
  dims = 1:20,
  resolution = 0.6,
  cluster_algorithm = "louvain",
  cluster_name = "velocity_cluster",
  species = "mouse",
  verbose = FALSE
)

sn_plot_dim(
  velocity_object,
  reduction = "umap",
  group_by = "real_cell_type",
  title = "Hermann spermatogenesis before velocity inference"
)
```

## Check the managed backend without installing it

`sn_run_velocity()` invokes a managed Python environment. To keep site builds
reproducible, this article never calls an installer. The extended chunks become
eligible only when `SHENNONG_REAL_PROFILE=all`, pixi is discoverable, and the
trajectory environment already exists on disk.

```{r velocity-backend-status}
pixi_info <- sn_check_pixi(quiet = TRUE)
trajectory_paths <- sn_pixi_paths("trajectory")
trajectory_environment <- file.path(
  trajectory_paths$workspace_env_dir,
  "default"
)
run_velocity_backend <- run_extended && isTRUE(pixi_info$installed) &&
  dir.exists(trajectory_environment)

tibble::tibble(
  profile = real_profile,
  pixi_installed = isTRUE(pixi_info$installed),
  environment = "trajectory/default",
  environment_ready = dir.exists(trajectory_environment),
  will_run_velocity = run_velocity_backend
)

pixi_control <- list(
  pixi = if (isTRUE(pixi_info$installed)) pixi_info$path else NULL,
  pixi_environment = "default",
  install_pixi = FALSE,
  quiet = TRUE
)
```

## Estimate velocity in the extended profile

The source spliced and unspliced matrices contain finite, non-negative
fractional molecule estimates. `enforce_normalization = TRUE` prevents scVelo
from mistaking those non-integer values for already log-transformed data; the
normalized expression is then log1p-transformed before Seurat-flavor HVG
selection. Both choices are retained in the unified result parameters and
backend manifest. This fixture uses scVelo's explicit deterministic model;
stochastic inference remains available through `backend_control` but is not
silently substituted if its generalized least-squares backend fails.

```{r velocity, eval=run_velocity_backend}
velocity_object <- sn_run_velocity(
  velocity_object,
  method = "scvelo",
  spliced_assay = "RNA",
  spliced_layer = "spliced",
  unspliced_assay = "RNA",
  unspliced_layer = "unspliced",
  reduction = "umap",
  store_name = "hermann_scvelo",
  backend_control = list(
    velocity_mode = "deterministic",
    enforce_normalization = TRUE,
    log1p_transform = TRUE,
    n_top_genes = 1500L,
    n_neighbors = 30L,
    seed = 717L,
    pixi = pixi_control
  )
)

velocity <- sn_get_result(velocity_object, "velocity", "hermann_scvelo")
head(velocity$tables$cells)
head(velocity$tables$transition_edges)
velocity$diagnostics
velocity_manifest <- velocity$models$artifacts
velocity_manifest$output_h5ad <- file.exists(velocity_manifest$output_h5ad)
velocity_manifest

sn_plot_velocity(velocity, color_by = "confidence")
```

Low confidence, weak splicing signal, or unstable directionality remains
visible in the stored diagnostics. The H5AD artifact recorded in
`models$artifacts` is the transition evidence consumed by CellRank; a smooth
arrow field is not sufficient evidence by itself.

## Estimate terminal states and fate

CellRank runs only after the real scVelo result has been stored. The default
stability rule is kept explicit rather than silently replaced by a fixed number
of terminal states.

```{r fate, eval=run_velocity_backend}
velocity_object <- sn_run_fate(
  velocity_object,
  velocity_name = "hermann_scvelo",
  reduction = "umap",
  store_name = "hermann_fate",
  backend_control = list(
    terminal_method = "stability",
    seed = 717L,
    pixi = pixi_control
  )
)

fate <- sn_get_result(velocity_object, "fate", "hermann_fate")
fate$tables$terminal_states
head(fate$tables$probabilities)
head(fate$tables$lineage_drivers)
fate$diagnostics

sn_plot_fate(fate)
```

If no macrostate meets the stability threshold, the backend raises an error.
A documented project may instead choose `terminal_method = "top_n"` or supply
explicit terminal states, but that decision should not be made automatically
inside a teaching article.

## RegVelo requires a real, versioned prior

RegVelo is not run with a toy regulatory network. It becomes eligible only when
the extended backend is ready and `SHENNONG_REGVELO_PRIOR` names an existing
regulator-target CSV supplied by the project.

```{r regvelo-status}
run_regvelo <- run_velocity_backend && nzchar(regvelo_prior) &&
  file.exists(regvelo_prior)

tibble::tibble(
  prior_path = regvelo_prior,
  prior_available = nzchar(regvelo_prior) && file.exists(regvelo_prior),
  will_run_regvelo = run_regvelo
)
```

```{r regvelo, eval=run_regvelo}
regvelo_object <- sn_run_velocity(
  velocity_object,
  method = "regvelo",
  spliced_assay = "RNA",
  spliced_layer = "spliced",
  unspliced_assay = "RNA",
  unspliced_layer = "unspliced",
  reduction = "umap",
  store_name = "hermann_regvelo",
  backend_control = list(
    prior_grn = regvelo_prior,
    max_epochs = 1500L,
    early_stopping = TRUE,
    seed = 717L,
    pixi = pixi_control
  )
)

regvelo <- sn_get_result(regvelo_object, "velocity", "hermann_regvelo")
head(regvelo$tables$cells)
head(regvelo$tables$transition_edges)
regvelo$models$artifacts
regvelo$diagnostics

sn_plot_velocity(regvelo, color_by = "pseudotime")
```

The prior GRN is a modeling assumption and must be versioned with the analysis.
The stored result records the actual prior-edge and model artifacts returned by
the backend; this article never substitutes fabricated adapter output.
