---
title: "Visualization with Shennong"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization with Shennong}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
real_data_file <- file.path(
  Sys.getenv("SHENNONG_REAL_DATA_DIR"),
  "single-cell",
  "kotliarov_pbmc.qs2"
)
real_profile <- Sys.getenv("SHENNONG_REAL_PROFILE", unset = "core")
run_requested <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")
run_vignette <- run_requested &&
  requireNamespace("qs2", quietly = TRUE) &&
  file.exists(real_data_file)
run_extended <- run_vignette && identical(real_profile, "all")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Visualization is a first-class part of Shennong. The plotting helpers wrap the
common Seurat and ggplot2 patterns used in single-cell reports: automatic point
sizing, stable palettes, compact panel sizing, rasterized scatter plots for
large objects, and direct support for stored Shennong results.

This article uses the public Kotliarov PBMC CITE-seq subset as the running
example. Its RNA and ADT values, 20 biological samples, response groups, and
acquisition batches are all retained from the source study. The chunks execute
in real-data builds and remain safe during ordinary package checks when the
local fixture is unavailable.

## Prepare one clustered object

The plot wrappers expect the same Seurat object produced by the analysis
workflow. For a fast demonstration, this example clusters the real PBMC subset with fewer PCs
and variable features than a full analysis.

```{r setup-kotliarov-pbmc}
library(Shennong)
library(Seurat)
library(dplyr)
library(ggplot2)

pbmc <- qs2::qs_read(real_data_file)

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  npcs = 20,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

table(pbmc$seurat_clusters, pbmc$real_response)
```

## Dimensional maps: labels, palettes, and rasterization

`sn_plot_dim()` is the default map for categorical metadata. It hides axes by
default, chooses a practical point size from the number of cells, and uses a
named palette registry instead of requiring every script to hand-code colors.

```{r plot-dim}
sn_plot_dim(
  object = pbmc,
  reduction = "umap",
  group_by = "seurat_clusters",
  label = TRUE,
  label_halo = FALSE,
  palette = "Paired",
  title = "Kotliarov PBMC clusters"
)
```

When you need colors programmatically, use the same registry directly.

```{r palette-api}
sn_list_palettes()

cluster_cols <- sn_get_palette(
  palette = "Paired",
  n = length(levels(pbmc$seurat_clusters)),
  palette_type = "discrete"
)

cluster_cols
```

## Feature maps: expression and density modes

`sn_plot_feature()` covers standard expression maps and density-style maps. The
same function handles legends, shared scales, rasterization, and panel sizing.
Rasterization is enabled by default; when `ggrastr` is available the point layer
is rasterized after the regular ggplot point size is applied, so small values
such as `pt_size = 0.01` behave like ordinary `geom_point()` sizes. When
`ggrastr` is unavailable, the plot remains usable through the vector point
layer.

```{r plot-feature-expression}
sn_plot_feature(
  object = pbmc,
  features = c("IL7R", "MS4A1", "LYZ"),
  reduction = "umap",
  mode = "expression",
  palette = "YlOrRd",
  title = "Canonical PBMC markers"
)
```

Density mode is useful when expression is sparse and you want to emphasize
where a marker-defined state sits in the embedding.

```{r plot-feature-density, eval = run_vignette && requireNamespace("Nebulosa", quietly = TRUE)}
sn_plot_feature(
  object = pbmc,
  features = "IL7R",
  reduction = "umap",
  mode = "density",
  density_style = "galaxy",
  title = "IL7R density"
)
```

The matched ADT assay can be shown on the WNN embedding after the real protein
counts have been CLR-normalized by the multimodal workflow.

```{r plot-adt-features, eval = run_vignette}
pbmc_cite <- sn_run_multimodal(
  object = qs2::qs_read(real_data_file),
  method = "wnn",
  assay = "RNA",
  adt_assay = "ADT",
  nfeatures = 1500,
  dims = 1:15,
  adt_dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)

sn_plot_feature(
  object = pbmc_cite,
  features = c("CD3-PROT", "CD19-PROT", "CD14-PROT"),
  assay = "ADT",
  layer = "data",
  reduction = "wnn.umap",
  palette = "YlOrRd",
  title = "Measured ADT proteins"
)
```

## Focused heatmaps for selected genes

When the question is about a known marker panel, a heatmap is usually clearer
than another embedding. `sn_plot_heatmap()` validates the requested genes and
can draw either cell-level or group-averaged heatmaps. The default
`mode = "cells"` follows Seurat's heatmap backend but hides cell names and ticks,
uses an 8 pt group label, rasterizes by default, and colors the group bar with a
Paired palette.

```{r plot-heatmap}
sn_plot_heatmap(
  object = pbmc,
  features = c("IL7R", "CCR7", "MS4A1", "CD79A", "LYZ", "NKG7"),
  group_by = "seurat_clusters",
  palette = "RdBu",
  label_angle = 45
)
```

Use `mode = "average"` when the comparison is about the group-level pattern
rather than every individual cell. This averages expression within each group
and then z-scores each gene by default, giving a compact marker-panel summary.

```{r plot-heatmap-average}
sn_plot_heatmap(
  object = pbmc,
  features = c("IL7R", "CCR7", "MS4A1", "CD79A", "LYZ", "NKG7"),
  group_by = "seurat_clusters",
  mode = "average",
  palette = "RdBu"
)
```

Use `split_by` when the same marker panel should be inspected separately across
samples, conditions, or batches.

```{r plot-heatmap-split, eval = run_vignette}
sn_plot_heatmap(
  object = pbmc,
  features = c("IL7R", "CCR7", "MS4A1", "CD79A", "LYZ", "NKG7"),
  group_by = "seurat_clusters",
  split_by = "real_response"
)
```

## Distribution plots for metadata and scores

Shennong wraps common summary plots so a report can move from cell-level maps to
sample-level summaries without switching visual grammar.

```{r plot-violin-box}
sn_plot_violin(
  object = pbmc,
  features = c("nFeature_RNA", "pctMT"),
  group_by = "real_response",
  palette = "Set2"
)

qc_summary <- pbmc[[]] |>
  dplyr::group_by(real_sample, real_response) |>
  dplyr::summarise(
    median_features = median(nFeature_RNA),
    median_mito = median(pctMT),
    .groups = "drop"
  )

head(qc_summary)

sn_plot_boxplot(
  qc_summary,
  x = real_response,
  y = median_features,
  y_label = "Median detected genes",
  panel_widths = 80,
  panel_heights = 70
)
```

## Marker dot plots from stored DE results

The DE workflow can store markers under `object@misc$de_results`. Once stored,
`sn_plot_dot(features = "top_markers")` can reuse that result directly instead
of asking users to manually copy marker lists between scripts.

```{r plot-dot}
pbmc <- sn_find_de(
  object = pbmc,
  analysis = "markers",
  group_by = "seurat_clusters",
  layer = "data",
  store_name = "cluster_markers",
  return_object = TRUE,
  verbose = FALSE
)

sn_plot_dot(
  x = pbmc,
  features = "top_markers",
  de_name = "cluster_markers",
  group_by = "seurat_clusters",
  n = 4,
  palette = "Purples"
)
```

For a curated marker panel, pass a named list. Each list element becomes a
free-width facet, so marker families remain visually separated even when the
groups contain different numbers of genes.

```{r plot-dot-groups}
level1_markers <- list(
  "Lymphoid" = c("CD3D", "CD3E", "MS4A1"),
  "Myeloid" = c("LYZ", "FCGR3A"),
  "Platelets" = c("PPBP")
)

sn_plot_dot(
  x = pbmc,
  features = level1_markers,
  group_by = "seurat_clusters"
)
```

## Composition plots for sample-level summaries

`sn_calculate_composition()` creates the table; `sn_plot_composition()` displays
it. Keeping those steps separate makes the denominator and grouping explicit.

```{r plot-composition}
composition <- sn_calculate_composition(
  x = pbmc,
  group_by = "real_sample",
  variable = "seurat_clusters",
  measure = "both",
  min_cells = 10
)

sn_plot_composition(
  composition,
  x = real_sample,
  fill = seurat_clusters,
  y = proportion,
  position = "stack",
  palette = "Paired"
)
```

## Standard ggplot2 still fits

When a plot is not covered by an `sn_plot_*()` wrapper, use plain ggplot2 and
resolve colors through Shennong. If `catplot` is installed, its compact theme is
a good final layer for manuscript-style figures.

```{r custom-ggplot}
response_levels <- sort(unique(qc_summary$real_response))
response_cols <- sn_get_palette(
  "Set2",
  n = length(response_levels),
  palette_type = "discrete"
)
names(response_cols) <- response_levels

p <- ggplot(
  qc_summary,
  aes(x = real_response, y = median_mito, colour = real_response)
) +
  geom_boxplot(outlier.shape = NA, colour = "grey35") +
  geom_jitter(width = 0.08, height = 0, size = 2) +
  scale_colour_manual(values = response_cols) +
  labs(x = NULL, y = "Per-sample median mitochondrial percent", colour = NULL)

if (requireNamespace("catplot", quietly = TRUE)) {
  p <- p + catplot::theme_cat(panel_widths = 80, panel_heights = 70)
}

p
```
