---
title: "Visualization with Shennong"
author:
- name: Songqi Duan
  email: duan@songqi.org
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization with Shennong}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_vignette <- identical(Sys.getenv("SHENNONG_RUN_VIGNETTES"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  eval = run_vignette
)
```

Visualization is a first-class part of Shennong. The plotting helpers wrap the
common Seurat and ggplot2 patterns used in single-cell reports: automatic point
sizing, stable palettes, compact panel sizing, rasterized scatter plots for
large objects, and direct support for stored Shennong results.

This article uses PBMC3k as the running example. The chunks are displayed by
default and are evaluated only with `SHENNONG_RUN_VIGNETTES=true`.

## Prepare one clustered object

The plot wrappers expect the same Seurat object produced by the analysis
workflow. For a fast demonstration, this example clusters PBMC3k with fewer PCs
and variable features than a full analysis.

```{r setup-pbmc3k}
library(Shennong)
library(Seurat)
library(dplyr)
library(ggplot2)

pbmc <- sn_load_data("pbmc3k")

pbmc <- sn_run_cluster(
  object = pbmc,
  normalization_method = "seurat",
  nfeatures = 1500,
  npcs = 20,
  dims = 1:15,
  resolution = 0.6,
  species = "human",
  verbose = FALSE
)
```

## Dimensional maps: labels, palettes, and rasterization

`sn_plot_dim()` is the default map for categorical metadata. It hides axes by
default, chooses a practical point size from the number of cells, and uses a
named palette registry instead of requiring every script to hand-code colors.

```{r plot-dim}
sn_plot_dim(
  object = pbmc,
  reduction = "umap",
  group_by = "seurat_clusters",
  label = TRUE,
  palette = "Paired",
  title = "PBMC3k clusters"
)
```

When you need colors programmatically, use the same registry directly.

```{r palette-api}
sn_list_palettes()

cluster_cols <- sn_get_palette(
  palette = "Paired",
  n = length(levels(pbmc$seurat_clusters)),
  palette_type = "discrete"
)

cluster_cols
```

## Feature maps: expression and density modes

`sn_plot_feature()` covers standard expression maps and density-style maps. The
same function handles legends, shared scales, rasterization, and panel sizing.

```{r plot-feature-expression}
sn_plot_feature(
  object = pbmc,
  features = c("IL7R", "MS4A1", "LYZ"),
  reduction = "umap",
  mode = "expression",
  palette = "YlOrRd",
  title = "Canonical PBMC markers"
)
```

Density mode is useful when expression is sparse and you want to emphasize
where a marker-defined state sits in the embedding.

```{r plot-feature-density, eval = FALSE}
sn_plot_feature(
  object = pbmc,
  features = "IL7R",
  reduction = "umap",
  mode = "density",
  density_style = "galaxy",
  title = "IL7R density"
)
```

## Distribution plots for metadata and scores

Shennong wraps common summary plots so a report can move from cell-level maps to
sample-level summaries without switching visual grammar.

```{r plot-violin-box}
pbmc$sample <- rep(paste0("donor", 1:6), length.out = ncol(pbmc))
pbmc$condition <- ifelse(pbmc$sample %in% paste0("donor", 1:3), "control", "stim")

sn_plot_violin(
  object = pbmc,
  features = c("nFeature_RNA", "percent.mt"),
  group_by = "condition",
  palette = "Set2"
)

qc_summary <- pbmc[[]] |>
  dplyr::group_by(sample, condition) |>
  dplyr::summarise(
    median_features = median(nFeature_RNA),
    median_mito = median(percent.mt),
    .groups = "drop"
  )

sn_plot_boxplot(
  qc_summary,
  x = condition,
  y = median_features,
  y_label = "Median detected genes",
  panel_widths = 80,
  panel_heights = 70
)
```

## Marker dot plots from stored DE results

The DE workflow can store markers under `object@misc$de_results`. Once stored,
`sn_plot_dot(features = "top_markers")` can reuse that result directly instead
of asking users to manually copy marker lists between scripts.

```{r plot-dot}
pbmc <- sn_find_de(
  object = pbmc,
  analysis = "markers",
  group_by = "seurat_clusters",
  layer = "data",
  store_name = "cluster_markers",
  return_object = TRUE,
  verbose = FALSE
)

sn_plot_dot(
  x = pbmc,
  features = "top_markers",
  de_name = "cluster_markers",
  group_by = "seurat_clusters",
  n = 4,
  palette = "Purples"
)
```

## Composition plots for sample-level summaries

`sn_calculate_composition()` creates the table; `sn_plot_composition()` displays
it. Keeping those steps separate makes the denominator and grouping explicit.

```{r plot-composition}
composition <- sn_calculate_composition(
  x = pbmc,
  group_by = "sample",
  variable = "seurat_clusters",
  measure = "both",
  min_cells = 10
)

sn_plot_composition(
  composition,
  x = sample,
  fill = seurat_clusters,
  y = proportion,
  position = "stack",
  palette = "Paired"
)
```

## Standard ggplot2 still fits

When a plot is not covered by an `sn_plot_*()` wrapper, use plain ggplot2 and
resolve colors through Shennong. If `catplot` is installed, its compact theme is
a good final layer for manuscript-style figures.

```{r custom-ggplot}
condition_cols <- sn_get_palette("Set2", n = 2, palette_type = "discrete")
names(condition_cols) <- c("control", "stim")

p <- ggplot(qc_summary, aes(x = condition, y = median_mito, fill = condition)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = condition_cols) +
  labs(x = NULL, y = "Median mitochondrial percent")

if (requireNamespace("catplot", quietly = TRUE)) {
  p <- p + catplot::theme_cat(panel_widths = 80, panel_heights = 70)
}

p
```
